% ionocast-methodology.tex
% Compile with: pdflatex ionocast-methodology.tex (×2 for references)
% ─────────────────────────────────────────────────────────────────────
\documentclass[11pt,a4paper]{article}

% ── Geometry ──────────────────────────────────────────────────────────
\usepackage[margin=2.5cm]{geometry}

% ── Fonts & encoding ─────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}

% ── Math ─────────────────────────────────────────────────────────────
\usepackage{amsmath,amssymb}

% ── Tables ───────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}

% ── Graphics & TikZ ──────────────────────────────────────────────────
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc,patterns,decorations.pathreplacing}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{fillbetween}

% ── Units ────────────────────────────────────────────────────────────
% siunitx removed; plain-text unit macros for portability.
\newcommand{\unit}[1]{\,\text{#1}}
\newcommand{\SI}[2]{#1\,\text{#2}}
\newcommand{\si}[1]{\text{#1}}
\newcommand{\dB}{\text{dB}}
\newcommand{\dBm}{\text{dBm}}
\newcommand{\dBi}{\text{dBi}}
\newcommand{\MHz}{\text{MHz}}
\newcommand{\GHz}{\text{GHz}}
\newcommand{\kHz}{\text{kHz}}
\newcommand{\Hz}{\text{Hz}}
\newcommand{\km}{\text{km}}
\newcommand{\GW}{\text{GW}}
\newcommand{\sfu}{\text{sfu}}
\newcommand{\W}{\text{W}}
\newcommand{\hour}{\text{h}}
\newcommand{\minute}{\text{min}}
\newcommand{\degree}{\ensuremath{{}^\circ}}

% ── Hyperlinks ───────────────────────────────────────────────────────
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black,hypertexnames=false]{hyperref}

% ── Misc ─────────────────────────────────────────────────────────────
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{float}

% ── Header/footer ────────────────────────────────────────────────────
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textit{ionocast Methodology}}
\fancyhead[R]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Title ────────────────────────────────────────────────────────────
\title{\textbf{ionocast}: A Real-Time HF/VHF Propagation Prediction Model\\
       for Amateur Radio Operators}
\author{Toprak Kilic, TA1BUT}
\date{May 2026}

% ═════════════════════════════════════════════════════════════════════
\begin{document}
\maketitle
\thispagestyle{fancy}

% ─────────────────────────────────────────────────────────────────────
\begin{abstract}
\noindent
\textbf{ionocast} provides real-time HF/VHF band-condition verdicts
for amateur radio operators (next-24-hour forward projections are
specified in Section~\ref{sec:projection} but not yet surfaced in
the runtime).  The model
runs entirely in the browser with no server-side state.  The operator's
QTH stays on-device for everything except two panels that need a
nearest-station lookup, in which case it is sent to ionocast's own
proxy for resolution and discarded after the response (see
Section~\ref{sec:privacy} for the full disclosure).  The model
combines an ITU-R P.533--grounded SNR
link budget with real-time data from NOAA SWPC (solar indices, D-RAP
absorption, OVATION auroral power, GOES proton flux), the kc2g
ionosonde network (observed MUF and foF2), GIRO digisondes (foF2,
foEs, hmF2), and WSPR spot observations from \texttt{wspr.live}.
A local validation harness scores predictions against retrospective
WSPR activity to quantify systematic bias and refine per-band
$\sigma$.

Key contributions:
\begin{itemize}\itemsep0pt\parsep2pt
\item \emph{Symmetric MUF consensus} (geometric mean) between
      gridded kc2g observations and the foF2 climatology.
\item \emph{Power-summed noise model} (ITU-R P.372) that correctly
      handles man-made noise dominance on the low bands.
\item \emph{Per-hop minimum MUF} and \emph{per-hop ionospheric loss
      accounting} (diurnal D-region, polar cap absorption, flare SID,
      auroral) evaluated at each great-circle reflection point rather
      than at a single midpoint.
\item \emph{Sauer--Wilkinson polar cap absorption} coupled to GOES
      $\geq 10$\,MeV proton flux, and \emph{flare-driven SID}
      absorption keyed to GOES X-ray class.
\item \emph{Storm-lag kernel} that captures the F-region's delayed
      response to $K_p$, with $D_{\text{st}}$ and IMF~$B_z$ used for
      earlier detection than the 3-hour $K_p$ alone.
\item \emph{Equatorial-anomaly} (paired crest enhancement at
      $\pm \SI{15}{\degree}$ dip plus on-equator trough depression)
      and \emph{winter-anomaly} corrections to the foF2 climatology,
      together with an \emph{asymmetric gray-line} bonus.
\item \emph{Elevation-dependent antenna gain} from
      hop-geometry-derived takeoff angles, plus \emph{NVIS detection}
      for short paths using local foF2.
\item Verdicts mapped to fixed dB tier cuts (+18 / +6 / -5 / -14)
      that don't scale with sigma; calibrated against on-air
      operator experience after a sigma-percentile experiment ran
      optimistic in real-world contests (Section~\ref{sec:tiers}).
\end{itemize}

Paths are sampled radially around the operator's QTH on a
12-bearing $\times$ 6-distance-ring grid (Section~\ref{sec:kc2gobs}),
so every QTH gets identical 72-point coverage with no fixed-city
geographic bias.  All computation is deterministic, repeatable,
and inspectable; no machine-learning black boxes.

\paragraph{Source and reproducibility.} The full ionocast
implementation --- browser code, Cloudflare Worker handlers, and
the offline calibration harness --- is open at
\url{https://github.com/topklc/ionocast}.  The 35-path
calibration basket, 27-station GIRO list, and per-band $\sigma$
table are reproduced verbatim from that repository in
Appendices~\ref{app:paths}, \ref{app:giro}, and the main text
(Table~\ref{tab:bandsigma}) respectively, so the validation
metrics below can be re-derived end-to-end from the paper plus
the repository alone.  The repository is the source of truth:
where this paper and the code disagree, the discrepancy reflects
either documentation lag (the code shipping a fix or rename
that the paper hasn't been updated to match) rather than a code
bug to follow the paper into; the calibration constants in
\texttt{src/constants.js} and the per-hop physics in
\texttt{src/physics/} are what produces the verdict an operator
sees.

\paragraph{Validation note.} The shipped harness scores against
30~d of WSPR activity aggregated two ways: a global ``open
somewhere'' truth (accBin in the mid-80s, Brier
${\sim}\!0.09\text{--}0.11$ at the 2026-05-08 post-B6 baseline)
which has a known structural ceiling --- the OR-over-paths
target overrewards permissive predictors and saturates
whenever \emph{any} path on the band has activity, quantified
in Section~\ref{sec:limits}~\#9; and a per-path mode (accBin
in the low-40s, Brier ${\sim}\!0.56$) which uses
TX/RX-bbox-restricted aggregates with a $\geq 1$ spot/h floor
and is dominated by operator-activity sparsity on
minority-traffic bands.  Exact baseline values are tabulated in
Section~\ref{sec:calibrate} (Table~\ref{tab:harnessruns});
both metrics drift with solar activity between runs and
shifted further at the 2026-05-08 post-B6 D-region/grayline
paired rewrite.  The 40+ percentage-point gap between the two
metrics on the same harness run is itself the structural-ceiling
evidence: a model that scores in the 80s on the global metric
and the 40s on the per-path metric is not 80-good or 40-good,
it is sitting on a metric whose ceiling reflects ``somewhere is
open'' rather than ``the requested path is open.''  The two
numbers are not directly comparable; per-pair WSPR ground-truth
wiring with calibrated availability priors is in progress.
Quote either number as a regression-detection signal, not as a
production-grade reliability claim.
\end{abstract}

\paragraph{Acronyms used throughout this paper.}
ACE: Advanced Composition Explorer (NASA, real-time solar-wind
monitor at L1).
AE: Auroral Electrojet index (geomagnetic).
DIDB / DIDBase: Digital Ionogram DataBase (UMass Lowell).
DSCOVR: Deep Space Climate Observatory (NOAA, L1 solar-wind
monitor).
EIA: Equatorial Ionisation Anomaly.
foF2 / foEs: F2-layer / sporadic-E layer ordinary-wave critical
frequency.
GIRO: Global Ionospheric Radio Observatory.
GNSS: Global Navigation Satellite System.
GOES: Geostationary Operational Environmental Satellite (NOAA).
hmF2: F2-layer peak height.
IGRF: International Geomagnetic Reference Field.
IMO: International Meteor Organization.
IRI / IRTAM: International Reference Ionosphere / IRI Real-Time
Assimilative Mapping.
LSTID: Large-Scale Travelling Ionospheric Disturbance.
MUF: Maximum Usable Frequency.
NVIS: Near Vertical Incidence Skywave.
OVATION: Oval Variation, Assessment, Tracking, Intensity, and Online
Nowcasting (NOAA SWPC auroral-precipitation model).
SDO: Solar Dynamics Observatory.
SFI: Solar Flux Index ($F_{10.7}$).
SID: Sudden Ionospheric Disturbance.
SSN: Sunspot Number.
TEP: Trans-Equatorial Propagation.
TID: Travelling Ionospheric Disturbance.
WSPR: Weak Signal Propagation Reporter network.
ZHR: Zenith Hourly Rate (meteor-shower intensity).

\tableofcontents
\newpage

% ═════════════════════════════════════════════════════════════════════
\section{Introduction}\label{sec:intro}

Amateur radio operators planning an operating session need to answer a
deceptively simple question: \emph{which bands will be open, and for
how long?}  The answer depends on the instantaneous state of the
ionosphere, the Sun, the geomagnetic field, and the operator's own
station (antenna, power, mode, and local noise environment).

Existing tools serve fragments of this problem:

\begin{itemize}[nosep]
  \item \textbf{VOACAP}~\cite{voacap} provides rigorous
        P.533-based~\cite{itup533} predictions but operates as a
        batch tool requiring manual circuit specification; it is
        not real-time.
  \item \textbf{N0NBH Solar Conditions}~\cite{n0nbh} provides a
        widely trusted heuristic band-condition table keyed to
        SFI and Kp, but without path geometry or operator context.
  \item \textbf{kc2g MUF map}~\cite{kc2g} renders observed MUF
        contours in near real time, but offers no forward
        projection and no SNR context.
  \item \textbf{HamQSL / DX Toolbox}~\cite{dxtoolbox} use
        threshold-based rules on solar indices, lacking physics
        grounding (HamQSL is the same site as
        N0NBH~\cite{n0nbh}).
\end{itemize}

\textbf{ionocast} occupies the intersection: real-time, physics-grounded,
operator-configurable, and privacy-preserving (operator QTH stays
on-device in \texttt{localStorage} between requests, and is sent
to the proxy worker only at request time for nearest-station
resolution; see Section~\ref{sec:privacy}).  The QTH parameterises
a radial basket of 72 propagation paths (12 compass bearings $\times$
6 distance rings: 2.5 / 4 / 6 / 9 / 12 / 16\,Mm) sampled relative
to the operator's location; the SNR budget runs per band, per path,
at the current time.  A forward-projection MUF pipeline (solar-zenith
analogy, storm depression, seasonal correction, sporadic-E
persistence; Section~\ref{sec:projection}) is documented but not
yet surfaced in the runtime; the Upcoming Disruptions panel
surfaces SWPC's own forecasts (3-day R/S/G probabilities, 3-h
$K_p$, 27-day SFI/$A_p$) rather than ionocast-derived per-band
projections.

This paper documents the complete model as implemented.  Every
numeric constant in the SNR budget and MUF estimation is given
either inline or in an appendix; the calibration setup that
produced the harness-fitted constants (the three coordinate-descent
starting seeds, the parameter sweep grid, and the WSPR baseline
files) is described in Appendix~\ref{app:repro}, so the amateur
radio community can audit, reproduce, and improve the model from
the paper plus the repository alone.

% ═════════════════════════════════════════════════════════════════════
\section{Data Sources}\label{sec:data}

Table~\ref{tab:sources} lists all upstream data feeds.  Every
upstream is reached through a Cloudflare Worker
(\texttt{functions/\_proxies.js}); none is fetched directly by the
browser.  Two reasons drive the proxy: most upstreams (SWPC, kc2g,
DONKI, wspr.live) lack the CORS headers that would permit
browser-direct fetching, so the worker is a CORS shim plus
edge-side cache; the two geo-resolved feeds
(\texttt{/api/giro}, \texttt{/api/tropo}) additionally take an
operator-supplied Maidenhead grid as a query parameter and
resolve it to the nearest measurement station server-side, so
the upstream sees a station identifier rather than the
operator's QTH.  \texttt{/api/hp30} fetches a single global
nowcast text file and is not geo-resolved.

\begin{table}[H]
\centering
\small
\caption{Upstream data sources and refresh cadences.  The
\emph{Refresh} column reports the upstream-publisher cadence (how
often the source produces new data); the worker-side cache is
polled more aggressively in some cases so the operator sees a new
value within ${\sim}\!1$\,min of upstream publication.  Specific
worker \texttt{freshSec} TTLs are tabulated in
\texttt{functions/\_proxies.js}.}
\label{tab:sources}
\begin{tabular}{@{}p{2.7cm}p{6.6cm}p{1.6cm}p{3.6cm}@{}}
\toprule
\textbf{Source} & \textbf{Data} & \textbf{Upstream} & \textbf{Protocol} \\
\midrule
NOAA SWPC
  & Kp, Ap, F10.7, X-ray class, 3-day forecast, 27-day outlook,
    D-RAP HAF, OVATION aurora HP, solar-wind Bz / speed / density,
    solar regions
  & 1\,min--6\,h & Proxied via \texttt{/api/swpc-*} \\
\addlinespace
prop.kc2g.com~\cite{kc2g} & Per-station MUF(3000F2), foF2, TEC & ${\sim}$15\,min & Proxied via \texttt{/api/kc2g} \\
\addlinespace
GIRO (lgdc.uml.edu)~\cite{giro} & Nearest digisonde foF2, foEs, hmF2 & ${\sim}$10\,min & Proxied via \texttt{/api/giro} \\
\addlinespace
GFZ Potsdam~\cite{hp30}$^{*}$ & Hp30 nowcast & 30\,min & Proxied via \texttt{/api/hp30} \\
\addlinespace
WDC Kyoto~\cite{wdckyoto} & Dst / Sym-H & 1\,h & Proxied via \texttt{/api/kyoto} \\
\addlinespace
SILSO~\cite{silso}$^{*}$ & Daily sunspot number & Daily & Proxied via \texttt{/api/silso} \\
\addlinespace
UWyo~\cite{uwyo} & Radiosonde soundings (lowest-1\,km $\mathrm{d}N/\mathrm{d}h$ over a 24-station basket) & 12\,h$^{\dagger}$ & Proxied via \texttt{/api/tropo} \\
\addlinespace
wspr.live~\cite{wsprlive} & WSPR spot counts and SNR by band & 10\,min & Proxied via \texttt{/api/wspr} (ClickHouse) \\
\addlinespace
NASA DONKI~\cite{donki} & CME and HSS analyses & ${\sim}$hours & Proxied via \texttt{/api/donki-*} \\
\bottomrule
\end{tabular}\\[4pt]
\footnotesize
$^{\dagger}$ The \SI{12}{\hour} cadence reflects the upstream
radiosonde launch schedule (00Z and 12Z); the worker-side cache
polls every 10 min so a new sounding reaches the operator within
${\sim}\!10$\,min of being posted to UWyo rather than waiting
the full upstream 12-hour interval.

$^{*}$ Hp30 (GFZ Potsdam) and Daily Sunspot Number (SILSO) are
fetched and surfaced in the Geomagnetic / Solar UI panels for
operator situational awareness, but neither feeds the SNR budget
or the verdict pipeline at present.  $K_p^{\text{eff}}$ from SWPC
serves the half-hour-cadence role that Hp30 could otherwise fill;
F\textsubscript{10.7A} 81-day mean serves the solar-activity role
of daily SSN.  Both feeds are kept wired for two specific future
roles: Hp30 as a sub-3-hour validation channel for the storm-lag
kernel of \S\ref{sec:kplag} (the kernel currently consumes
3-hour $K_p$ samples and would benefit from a within-bucket
cadence to refit $\tau_{\text{peak}}$ and
$\tau_{\text{decay}}$), and SSN-v2 as an alternative anchor
when the F\textsubscript{10.7A} feed is degraded (the
correlation between R12 / R-MgII / SSN-v2 and
F\textsubscript{10.7A} lets a downstream substitution stand in
during outages, similar to the illumination-ratio fallback in
\S\ref{sec:perhopmuf}).
\end{table}

\subsection{Privacy Model}\label{sec:privacy}

The operator's QTH is stored on-device in \texttt{localStorage}
and is sent to the Cloudflare Worker only at request time, for
nearest-station resolution on the geo-aware proxy endpoints; the
worker neither writes the grid back to a server-side store nor
logs request bodies / query strings at the application level.
Privacy posture in detail:
\begin{itemize}[nosep]
  \item \textbf{Stored on device only.}  The
        \texttt{ionocast\_qth} key in \texttt{localStorage} holds
        the operator's grid; nothing reads or writes it
        server-side, and there is no account / login model.
  \item \textbf{Sent at request time for resolution, not retained.}
        Endpoints like \texttt{/api/giro} take a grid string as a
        query parameter so the worker can resolve it to the
        nearest digisonde station; the worker source contains no
        application-level logging of request bodies or query
        strings (inspectable at
        \url{https://github.com/topklc/ionocast}, the
        \texttt{functions/} directory), and the upstream
        response contains station data, not the original grid.
        Cloudflare's edge platform does retain its own access
        logs at the request level (URL + status + IP), described
        below.
  \item \textbf{WSPR live queries are global hourly aggregates}
        with no grid field at all; the operator's location is
        never transmitted to \texttt{wspr.live}.  (The offline
        calibration harness in \texttt{scripts/harness.mjs}
        does use a coarser grid for retrospective sample
        bucketing, but is not part of the runtime page.)
  \item \textbf{No third-party analytics or tracking} are
        loaded; the only request-side trace is Cloudflare's
        own edge access logging, retained per the platform's
        free-/standard-tier policy (typically up to 72\,h
        worth at the time of writing; consult Cloudflare's
        current Worker / Pages documentation for the
        authoritative retention window).
\end{itemize}
The honest summary: the QTH is sent to the proxy worker
\emph{for resolution at request time}, but is not retained,
logged in application form, or shared with third parties.
Operators who want zero server-side QTH exposure can run the
static UI against the upstream APIs directly (the worker is a
performance / CORS shim, not a privacy boundary).

\paragraph{First-visit default.}  On first load, with no saved QTH,
ionocast derives a regionally plausible grid from the browser's
runtime timezone.  \texttt{Intl.DateTimeFormat().resolvedOptions().timeZone}
is looked up in a compiled table of ${\sim}$60 major ham-active IANA
zones (e.g.\ \texttt{Europe/Berlin}~$\to$~JO62,
\texttt{Asia/Tokyo}~$\to$~PM95,
\texttt{Australia/Sydney}~$\to$~QF56; the full table lives in
\texttt{src/physics/qth.js} and is open for inspection at the repo
URL given in the abstract).  Zones outside the table
synthesise a grid from the UTC offset: longitude
$\approx \text{offset}\times 15\text{\textdegree}$ at a fixed
mid-latitude (45\textdegree{} N for the northern hemisphere,
$-35$\textdegree{} for identified southern-hemisphere zones).  The
timezone fingerprint is already exposed on every HTTPS request, so
using it for a first-visit default adds no new privacy surface.  When
even that math fails, the last-resort fallback is \textbf{JN05}
(approximately 45.5\textdegree{} N, 1\textdegree{} E --- the grid
covers $0\text{--}2\degree$\,E, $45\text{--}46\degree$\,N) --- a
French / west-Swiss midlatitude reference point chosen as a
plausible-default location for the worldwide majority of
first-visit operators.  (The CCIR R-12 / ITU-R P.1239 foF2
coefficients underlying this paper are global spherical-harmonic
expansions, not tabulated at a specific mid-latitude European
station; the fallback grid is a UX convenience rather than a
geographic anchor for the climatology.)  No request is ever made
to a geolocation service.

% ═════════════════════════════════════════════════════════════════════
\section{SNR Budget Model}\label{sec:snr}

The core of ionocast is a complete link budget that computes the
physics-only SNR margin $M_{\text{physics}}$ (in~dB) for each
amateur band on each reference path:
\begin{equation}\label{eq:budget}
  M_{\text{physics}} = P_{\text{tx}} + G_{\text{ant}}
    - L_{\text{fs}} - L_{\text{flare,path}} - L_{\text{Dreg}}
    - L_{\text{PCA}} - L_{\text{aur}}
    - L_{\text{MUF}} - n_{\text{hops}} L_{\text{iono}}
    - L_{\text{low}} - L_{\text{hop}} - L_{\text{Es}}
    - N - S_{\text{req}}
\end{equation}
where every term is defined in the following subsections.  The
verdict-driving margin $M_{\text{verdict}} = M_{\text{physics}} +
B_{\text{grayline}} + \max(B_{\text{TEP}}, B_{\text{scatter}})$
adds the additive mode-bonus terms of \S\ref{sec:ensemble}
(Eq.~\ref{eq:blend}); $M_{\text{physics}}$ alone is what
Eq.~\ref{eq:budget} computes, and bonus-augmented
$M_{\text{verdict}}$ is what feeds the tier reliability
formula in \S\ref{sec:reliability}.  A positive $M_{\text{physics}}$
means the budget alone clears the mode threshold; the magnitude
indicates robustness, but a near-zero or slightly negative
$M_{\text{physics}}$ on a TEP-eligible 15\,m path can still
yield a comfortable verdict once the bonuses fire.  The term
$L_{\text{flare,path}}$ is the joint flare-absorption product of
$L_{\text{DRAP}}$ (Section~\ref{sec:labs}) and the per-hop
$L_{\text{flare}}$ sum (Section~\ref{sec:lflare}); inlined here
so the reader does not have to forward-jump:
\begin{equation}\label{eq:flareabs-inline}
  L_{\text{flare,path}}
    = \max\!\Bigl(L_{\text{DRAP}},\,
        \sum_{k=1}^{N} L_{\text{flare}}(f, \cos\chi_k)\Bigr)
\end{equation}
Both symbols are defined separately because they have independent
physical sources, but they collapse to a single term in the budget
to avoid double-charging the path for the same flare.  See
Section~\ref{sec:labs} for the $\max(\cdot)$ rationale and the
geographic asymmetry between the two contributing terms.

\paragraph{Global absorption-sum saturation.}  Each of the four
ionospheric absorption terms ($L_{\text{flare,path}}$,
$L_{\text{Dreg}}$, $L_{\text{PCA}}$, $L_{\text{aur}}$) is
individually capped at \SI{50}{\dB} per path.  Their sum is also
\emph{hard-capped} at \SI{100}{\dB} with proportional rescaling
(no smoothing knee): if a sunlit polar hop during an X10 flare
with an S3 SEP and a $K_p = 9$ main-phase storm produced
hypothetical contributions
$(L_{\text{flare,path}}, L_{\text{Dreg}}, L_{\text{PCA}}, L_{\text{aur}})
= (50, 30, 40, 30)$\,dB, the additive sum reaches \SI{150}{\dB}
(well past physical D-region saturation $\sim$80--100\,dB; beyond
that the band is dead regardless and adding more absorption is
unphysical).  The cap then scales each contribution by a factor
$100/150 \approx 0.667$, producing the rescaled set $(33.3, 20.0,
26.7, 20.0)$\,dB (path total exactly \SI{100}{\dB}) so
per-mechanism attribution is preserved
($L'_i = L_i \cdot 100/\sum L_i$).

\paragraph{Why this gate is hard while the others are smooth.}
The auroral $K_p$, OVATION HP, CGM-edge, and IMF $B_z$ gates
elsewhere are smoothed because they fire under routine
active-$K_p$ conditions where a 5--20\,dB step would visibly
shift verdicts.  The 100\,dB absorption-sum cap fires only on a
simultaneous X10 flare, S3 SEP, and $K_p = 9$ main-phase storm
intersecting on a sunlit polar hop --- a scenario realised at most
a few hours per solar cycle, so the discontinuity at $\sum L_i =
100$ is unobservable in any operator-relevant regime.  The hard
form is intentional; the inconsistency with the smoothed gates is
acknowledged here rather than papered over.

\paragraph{Why proportional rescaling rather than cap-and-skip.}
The four contributing terms ($L_{\text{flare,path}}$,
$L_{\text{Dreg}}$, $L_{\text{PCA}}$, $L_{\text{aur}}$) are driven
by physically distinct ionisation sources --- soft X-ray
photo-ionisation, quiet-day diurnal photo-ionisation,
proton-induced ionisation, and particle-precipitation ionisation
respectively --- but they all enhance \emph{the same} D-region
electron column, which has a finite total-ionisation depth
(${\sim}80$--$100$\,dB of absorption at HF when fully populated;
beyond that further electrons displace existing ones rather than
adding new absorbers). Past the column's physical ceiling, the
correct behaviour is not ``add until budget then drop further
contributions'': cap-and-skip would arbitrarily zero out whichever
mechanism happens to be evaluated last in the implementation order
and leak attribution dependence into the UI's per-mechanism
display. Proportional rescaling, $L'_i = L_i \cdot 100/\sum L_i$,
preserves each mechanism's relative contribution at saturation.
Concretely, a pathological path stacking
``50\,dB flare-path + 35\,dB PCA + 30\,dB aurora + 25\,dB
quiet-day'' (sum \SI{140}{\dB}, well past the
\SI{100}{\dB} ceiling) gets rescaled by the factor
$100/140 \approx 0.714$ to
``$\SI{35.7}{\dB}$ flare-path + $\SI{25.0}{\dB}$ PCA +
$\SI{21.4}{\dB}$ aurora + $\SI{17.9}{\dB}$ quiet-day'' (sum
exactly \SI{100}{\dB}), so the operator-meaningful relative
ratios (flare-path is the dominant mechanism, then PCA, then
aurora, then quiet-day) are preserved.  The rescale is
conservative on terms not individually near saturation: a
\SI{12}{\dB} auroral contribution gets shaved to
${\sim}\SI{11.4}{\dB}$ when the sum is at \SI{105}{\dB}, but
this is well under any verdict-relevant threshold (the band
is dead at $\sum \ge \SI{80}{\dB}$ regardless), so the
conservatism does not shift verdicts in any
operator-meaningful regime.

% ── 3.1 Free-space loss ─────────────────────────────────────────────
\subsection{Free-Space Loss (Friis / ITU-R P.525)}

The free-space path loss follows the standard ITU-R
P.525~\cite{itup525} formulation:
\begin{equation}\label{eq:fspl}
  L_{\text{fs}} = 32.44 + 20\log_{10}(d_{\text{km}})
                        + 20\log_{10}(f_{\text{MHz}})
\end{equation}
The distance $d_{\text{km}}$ is the great-circle path length,
clamped to a minimum of \SI{50}{\km}.  $\log_{10}$ is well-defined
for any $d > 0$, so this is not a numerical safeguard but a
modelling floor: at $d \ll \SI{50}{\km}$ the far-field radiation
pattern assumed by the Friis equation no longer applies (near-field
effects, ground-wave coupling, and antenna-mutual-impedance
modulation dominate the link budget instead), and a per-band
constant $L_{\text{fs}}$ is the operationally honest answer
below that scale.  Note that the multi-hop term
$L_{\text{hop}}$ (Eq.~\ref{eq:lhop}) does not share this
\SI{50}{\km} floor: for paths with $d < d_{\text{hop}}^{\max}$,
$N_{\text{extra}} = \max(0, d/d_{\text{hop}}^{\max} - 1) = 0$
identically, so the term vanishes regardless of how short $d$
becomes.  On a degenerate self-path ($d \to 0$), the budget
charges the floored $L_{\text{fs}}$ for the equivalent of a
50\,\si{\km} link with no multi-hop contribution; the
verdict is dominated by the noise / SNR-required terms in
that regime.

% ── 3.2 MUF proximity loss ──────────────────────────────────────────
\subsection{MUF Approach and Over-MUF Loss}\label{sec:lmuf}

Let $r = f / \text{MUF}$.  The proximity loss is:
\begin{equation}\label{eq:lmuf}
  L_{\text{MUF}} =
  \begin{cases}
    0 & r \leq 0.70 \\[4pt]
    10\!\left(\dfrac{r - 0.70}{0.30}\right)^{\!2}
      & 0.70 < r \leq 1.00 \\[6pt]
    10 + 36\sqrt{r - 1}
      & r > 1.00
  \end{cases}
\end{equation}
Below 70\% of MUF the ionosphere provides comfortable refraction with
negligible loss.  Between 70--100\% the loss rises quadratically.
Above MUF (over-MUF operation, e.g.\ via Es or scatter), the loss
escalates rapidly.  The piecewise formula is $C^0$ continuous at
both joints ($r = 0.70$ and $r = 1.00$, where $L_{\text{MUF}} = 0$
and $\SI{10}{\dB}$ respectively) but the derivative at $r = 1.00$
is unbounded ($\sqrt{r - 1}$ has infinite slope at the joint),
producing a $C^1$ discontinuity at MUF.  This is intentional: the
slope explosion encodes the physical reality that crossing MUF is
an abrupt regime change (refraction $\to$ over-MUF scatter), and
smoothing the joint with a finite-slope blend would underweight
the rise in path variance at the boundary.  Operators using the
budget at $r = 1$ should expect the verdict to be
discontinuity-sensitive there.

\paragraph{Heritage.}  The piecewise form (zero below $r = 0.7$,
quadratic in the approach band, square-root over MUF) is a
VOACAP / IONCAP-style proximity penalty rather than a literal
ITU-R P.533~\cite{itup533} formula: P.533's analogous treatment is
embedded in the F2 propagation-loss routine (Annex 1, basic
transmission-loss tabulation) and is not isolable as a closed-form
$L_{\text{MUF}}(r)$.  ionocast adopts the simpler closed form
because the MUF input itself is the kc2g real-time observation,
so the proximity loss is anchored to data rather than re-derived
from P.533's monthly-median MUF prediction routine.

% ── 3.3 D-region absorption (flare) ────────────────────────────────
\subsection{D-Region Absorption: Flare-Enhanced (SWPC D-RAP)}\label{sec:labs}

SWPC's D-RAP product provides a Highest Affected Frequency (HAF) in
MHz.  The flare-enhanced absorption is:%
\footnote{The label \texttt{eq:labs} and the section anchor
\texttt{sec:labs} are legacy short identifiers (early drafts
used \texttt{L\_abs} as the in-equation variable); the actual
math symbol used throughout this section is $L_{\text{DRAP}}$.
The corresponding code function is \texttt{lAbsDb} in
\texttt{src/physics/loss.js}.  Reader-facing convention:
$L_{\text{DRAP}}$ everywhere in the paper math, \texttt{lAbsDb}
in the code.}
\begin{equation}\label{eq:labs}
  L_{\text{DRAP}} =
  \begin{cases}
    0 & r < 0.25 \\[3pt]
    \dfrac{r - 0.25}{0.05}\,\cdot\,3\,r^{1.5}
      & 0.25 \le r < 0.30 \\[6pt]
    3\,r^{1.5} & r \ge 0.30
  \end{cases},\qquad r = \text{HAF}/f
\end{equation}
The $0.25 \le r < 0.30$ ramp retires a small step the bare gate
produced (the formula at $r = 0.3$ evaluates to
$3 \cdot 0.3^{1.5} \approx \SI{0.49}{\dB}$, jumping in instantly
from zero); the linear ramp turns the term on smoothly and is
zero by construction at $r = 0.25$.  The ramp is $C^0$ continuous
at both joints but $C^1$-discontinuous: the derivative jumps from
$0$ (lower flat branch) to $\sim\!\SI{7.5}{\dB}/\Delta r$ at
$r = 0.25$, then settles into the $r^{1.5}$ shape at $r = 0.30$.
The jump is small in absolute terms ($\sim$\SI{0.025}{\dB} of
verdict shift across a 1\%-of-MUF $r$ change) and is accepted as
the cost of avoiding the much larger 0.5-dB step the bare gate
produced.

\paragraph{Relationship to the per-hop $L_{\text{flare}}$ term.}  D-RAP
HAF is derived by SWPC from the same GOES X-ray flux that drives
the per-hop $L_{\text{flare}}$ term in Section~\ref{sec:lflare}, so
the two are combined into a single per-path flare-absorption term
$L_{\text{flare,path}} = \max(L_{\text{DRAP}}, \sum_k L_{\text{flare}})$
(defined inline as Eq.~\ref{eq:flareabs-inline} alongside the budget
at the top of this section) rather than summed.
The $\max(\cdot)$ form (rather than a binary ``suppress
$L_{\text{DRAP}}$ whenever $L_{\text{flare}} > 0$'') is more
conservative on the leading edge of a flare: D-RAP's operational
forecast model can already report \SI{8}{\dB}--\SI{10}{\dB} at QTH while
the per-hop $L_{\text{flare}}$ proxy is still tracking the
instantaneous X-ray class at M1 ($\sim$\SI{4}{\dB}). $L_{\text{DRAP}}$
remains active when no per-hop flare is firing but D-RAP still
reports a non-zero HAF: typically the early rise of a small or
unstable event before the X-ray classifier flags it, or the
post-peak decay tail. (Quiet-day background absorption is handled by
$L_{\text{Dreg}}$ in Section~\ref{sec:labsd}, not by D-RAP.)

\paragraph{Geographic asymmetry of the two terms.}  $L_{\text{DRAP}}$
is a single number sampled at \emph{QTH} (the SWPC HAF product is
spatially evaluated; ionocast queries it at the operator's
location).  The per-hop sum $\sum_k L_{\text{flare}}(\cos\chi_k)$
samples each F-region reflection point along the path, so on a
trans-equatorial 12\,000\,km circuit it can pick up daylight hops
the QTH itself isn't in.  A short path entirely in QTH's local
zone reads $L_{\text{DRAP}}$ as the binding constraint; a long path
crossing the sunlit hemisphere reads the per-hop sum.  The
$\max(\cdot)$ takes whichever local probe is more pessimistic on
that path, so the geographic inhomogeneity is intentional: a
distant flare hitting the path's middle hops should not be hidden
by the QTH having moved into night, and a flare detected in the
operator's HAF cell should not be discounted just because the
specific reflection points sampled happen to land in twilight.

% ── 3.4 D-region absorption (quiet) ────────────────────────────────
\subsection{D-Region Absorption: Quiet-Day Diurnal}\label{sec:labsd}

ionocast computes diurnal D-region absorption from a continuous
formula keyed on a single anchor constant $K_D$:
\begin{equation}\label{eq:labsd}
  L_{\text{Dreg}}(f, \chi) = L_{\text{day}}(f) \cdot
    \bigl(\max(0,\,\cos\chi)\bigr)^{0.7},
  \qquad
  L_{\text{day}}(f) = \frac{K_D}{(f_{\text{MHz}} + 0.5)^{2}},
  \qquad K_D = 200~\si{\dB}\cdot\si{\MHz}^{2}
\end{equation}
where $\chi$ is the solar zenith angle at the reflection point and
$L_{\text{day}}(f)$ is the local-noon-overhead D-region absorption.
At night ($\cos\chi < 0.05$) the term returns zero.  The form is a
deliberate reshape of the prior $A_{\text{base}}(f) \cdot
(\cos\chi)^{1.3}$ table-and-step model (S0 \#2 paired rewrite,
landed 2026-05-07; harness re-fit deferred): the $1/(f+0.5)^{2}$
frequency dependence tracks ITU-R P.533~\cite{itup533} Annex 1's
non-deviative-absorption shape (a Lockwood / IONCAP-heritage
$1/(f+f_L)^{1.98}$ form with $f_L \approx 1.4$\,\si{\MHz}) within
the same $\sim\!2$\,dB overlap window the prior table claimed,
while the gentler $\cos^{0.7}$ exponent (versus $\cos^{1.3}$)
charges more absorption at moderate zenith angles ($\cos\chi \in
[0.3, 0.7]$, where $\cos^{0.7}$ runs $\sim\!0.4$ to $\sim\!0.8$
versus $\cos^{1.3}$'s $\sim\!0.2$ to $\sim\!0.6$) and matches the
multi-hop midday DX residuals the prior form under-predicted on.
The ITU-R P.533 Annex 1 cosine form is itself a
$\cos^{p}(0.881\chi)$ with month- and dip-latitude-varying $p$;
the closed-form $\cos^{0.7}$ used here is operationally a
midlatitude-averaged simplification, used because the budget
applies its own per-band $\sigma_g$ widening on top
(\S\ref{sec:reliability}) rather than passing latitude /
seasonality dependence into the absorption term itself.

The constant $K_D = 200$ was chosen so $L_{\text{day}}(\SI{1.838}{\MHz}) =
\SI{36.6}{\dB}$, a value bracketed by the prior table's
operator-experience anchor ($\SI{28}{\dB}$, K9LA / ARRL-derived) and
the literal P.533~$/1.6$~oblique-incidence target ($\SI{45.7}{\dB}$
divided by $1.6$ to convert vertical-incidence to oblique).
Illustrative noon values at the canonical band centres,
$L_{\text{day}}(f)$ from Eq.~\ref{eq:labsd}, replace the prior
Table~2 anchors:
\begin{equation*}
\begin{array}{lrlr}
\textbf{Band} & {f}~(\si{\MHz}) &
\textbf{Band} & {f}~(\si{\MHz})\\
\midrule
160\,\text{m} & 1.838~\to~36.6~\si{\dB} & 20\,\text{m} & 14.097~\to~0.94~\si{\dB} \\
80\,\text{m}  & 3.570~\to~12.1~\si{\dB} & 17\,\text{m} & 18.106~\to~0.58~\si{\dB} \\
60\,\text{m}  & 5.366~\to~5.80~\si{\dB} & 15\,\text{m} & 21.096~\to~0.43~\si{\dB} \\
40\,\text{m}  & 7.040~\to~3.52~\si{\dB} & 12\,\text{m} & 24.924~\to~0.31~\si{\dB} \\
30\,\text{m}  & 10.140~\to~1.76~\si{\dB} & 10\,\text{m} & 28.126~\to~0.24~\si{\dB} \\
\end{array}
\end{equation*}
\noindent
The values shift versus the prior K9LA-anchored table: 160\,m
noon climbs ($28~\to~36.6$\,\si{\dB}, ${+}\SI{8.6}{\dB}$),
80\,m noon drops ($18~\to~12.1$\,\si{\dB},
${-}\SI{5.9}{\dB}$); upper bands shift smaller ($\le \SI{1}{\dB}$
on each).  An offline calibration sweep against the
30-day WSPR basket (\S\ref{sec:calibrate}) is the
follow-up action item; the constants above land the budget at the
literature-anchored shape and let the harness retune
$\sigma_g$ / $L_{\text{iono}}$ around it.

For a multi-hop path, absorption is \emph{summed} across all
reflection points rather than evaluated at the path midpoint only:
for an $N$-hop path with reflections at great-circle fractions
$(2k-1)/(2N)$,
\begin{equation}\label{eq:labsd-path}
  L_{\text{absD},\text{path}} = \min\!\left(
    L_{\text{cap}},\;
    \sum_{k=1}^{N} L_{\text{Dreg}}(f, \chi_k)\right)
\end{equation}
with $L_{\text{cap}} = \SI{50}{\dB}$ guarding against degenerate
values when every hop is near local noon on a low band.  A
transequatorial path whose midpoint is dark but whose northern and
southern hops are in daylight therefore sees the two daylight hops'
absorption, not zero; conversely a path whose midpoint sits over
the terminator but whose hops are deep in night loses no
contribution from the single sunlit fraction nearest the midpoint.
The grayline bonus of \S\ref{sec:grayline} pairs with this term:
$L_{\text{day}}(f) - L_{\text{Dreg}}(f, \chi)$ is the
zenith-dependent absorption refund the grayline mechanism
captures, additive on top of the day-side $L_{\text{Dreg}}$
charge.

% ── 3.4b Polar cap absorption ───────────────────────────────────────
\subsection{Polar Cap Absorption (Sauer--Wilkinson)}\label{sec:lpca}

During solar energetic particle (SEP) events, protons accelerated by a
CME or flare precipitate into the polar cap and ionise the D-region,
producing severe HF absorption on polar paths for hours to
days~\cite{sauer2008}.  ionocast activates a polar cap absorption
(PCA) term when two gates are simultaneously satisfied:
\begin{enumerate}[nosep]
  \item the GOES integral $\geq 10$\,MeV proton flux exceeds the NOAA
        S1 threshold, $\Phi_{p10} \geq \SI{10}{pfu}$;
  \item the reflection point sits inside the polar cap,
        $|\phi_{\text{CGM}}| \geq 60\text{\textdegree}$.
\end{enumerate}
The per-hop magnitude uses a log-flux scaling (a calibrated
deviation from the canonical Sauer--Wilkinson $\sqrt{\Phi}$
form~\cite{sauer2008}, chosen because the saturating-recombination
behaviour at the highest fluxes is gentler than $\sqrt{\Phi}$
predicts).  Empirical evidence for the deviation comes from
amateur-band PCA observations catalogued by K9LA~\cite{k9la} during
cycle-24 SEPs above $1000\,\text{pfu}$, which report
$5$--$10\,\si{\dB}$ of remaining 14\,MHz path on polar circuits
where the canonical $\sqrt{\Phi}$ form would predict $>30\,\si{\dB}$.
The Sauer--Wilkinson 2008 derivation is itself anchored in the
30~MHz riometer record at fluxes spanning the $10$--$10^{4}\,\text{pfu}$
range (their Figures 4--5b, validating against 11 large
SEP events 1992--2002 at Thule, Greenland), but the paper does not
specify a saturation regime above any particular flux threshold;
ionocast's log-flux substitution is a downstream calibration choice
informed by the K9LA observational record rather than by Sauer \&
Wilkinson directly.  The same $f^{-1.5}$ frequency dependence as
ordinary D-region absorption is retained:
\begin{equation}\label{eq:lpca}
  L_{\text{PCA}}(f, \Phi_{p10}) = \min\!\left(
    \SI{30}{\dB},\;
    5\log_{10}\!\frac{\Phi_{p10}}{10}
    \cdot \left(\frac{7}{f_{\text{MHz}}}\right)^{\!1.5}
  \right)
\end{equation}
The inner $\min(\SI{30}{\dB}, \cdot)$ in Eq.~\ref{eq:lpca} caps
\emph{each} per-hop contribution at \SI{30}{\dB}; per-hop
contributions are then summed along the great circle (the same hop
grid used by $L_{\text{Dreg}}$) and the resulting path total is
capped again at \SI{50}{\dB}.  Aurora ($L_{\text{aur}}$) uses the
same \SI{30}{\dB} per-hop cap (named in code as
\texttt{AUR\_PER\_HOP\_CAP\_DB}, mirroring
\texttt{PCA\_PER\_HOP\_CAP\_DB}); flare-driven SID
($L_{\text{flare}}$) uses a slightly higher \SI{40}{\dB} per-hop
cap because the strongest X-class flares produce essentially
total HF blackout on sunlit hops where the riometer record
\emph{has} read the 30--40\,dB regime (e.g., the X28+ Halloween
2003 events), whereas PCA and aurora rarely exceed 30\,dB for
sustained periods even at S5 / G5 levels.  All three share the
same path-total \SI{50}{\dB} cap (a single hop in the budget
cannot dominate the path beyond the aggregate ionospheric
absorption ceiling).  Midlatitude paths see $L_{\text{PCA}} =
0$ even during a severe SEP because no hop traverses the polar cap;
polar paths see order-of-magnitude absorption within minutes of
proton-event onset.

\paragraph{Multi-channel onset detection.}  The GOES integral
$\geq 1$\,MeV proton channel rises ${\sim}\SI{1}{h}$ before
$\geq 10$\,MeV during a hard SEP event because lower-energy protons
arrive first along the connecting magnetic field line.  When
$\Phi_{p1} \geq \SI{10}{pfu}$, an early-warning per-hop bump is
applied on polar hops:
\begin{equation}\label{eq:lpca-onset}
  L_{\text{PCA,onset}}(f, \Phi_{p1}) = \min\!\left(
    L_{\text{onset,cap}},\;
    \alpha_{\text{onset}} \cdot \log_{10}\!\frac{\Phi_{p1}}{10}
    \cdot \left(\frac{7}{f_{\text{MHz}}}\right)^{\!1.5}
  \right)
\end{equation}
with $\alpha_{\text{onset}} = \SI{1}{\dB}$/decade (vs.\
\SI{5}{\dB}/decade in the main term, since $\Phi_{p1}$ tracks
arrival of the SEP front rather than the absorbing population) and
$L_{\text{onset,cap}} = \SI{5}{\dB}$/hop.

\paragraph{Smooth handoff.}  $L_{\text{PCA,onset}}$ stays active
past the S1 threshold and the path sum uses
$\max(L_{\text{PCA}}, L_{\text{PCA,onset}})$ per hop, so the harder
channel takes over smoothly once its absorption exceeds the soft
channel's.  This avoids the \SI{5}{\dB} cliff a hard-gate handoff
would produce at the threshold instant. The $\geq 100$\,MeV channel is fetched alongside the other
two via the same SWPC integral-protons proxy (the JSON response
includes all integral channels in a single call) and cached in
\texttt{src/data/fetchers.js} as \texttt{p100}, but it is not
currently consumed by the physics layer. It surfaces only in
diagnostics for now; future work will use it to scale the per-hop
cap when the SEP penetrates deep into the D-region (the deepest
channels reach the lower D-region where the absorption cap is
itself frequency-dependent).

% ── 3.4c Flare-driven SID ───────────────────────────────────────────
\subsection{Flare-Driven Sudden Ionospheric Disturbance}\label{sec:lflare}

Solar soft X-ray photons from an in-progress flare enhance D-region
electron density on the sunlit side of the Earth within seconds of
the flare peak~\cite{mitra1974}.  ionocast reads the current SWPC
\texttt{xrayClass} (updated once per minute, well under the model's
10-minute refresh) and applies an additive absorption term on
sunlit hops, with a smooth twilight ramp $r(\cos\chi_k)$ across
the terminator:
\begin{equation}\label{eq:lflare}
  L_{\text{flare}}(f, X, \cos\chi_k) =
    \min\!\left(\SI{40}{\dB},\;
      D(X) \cdot \left(\dfrac{7}{f_{\text{MHz}}}\right)^{\!1.5}\right)
    \cdot r(\cos\chi_k)
\end{equation}
\begin{equation}\label{eq:lflare-ramp}
  r(\cos\chi_k) =
  \begin{cases}
    0                              & \cos\chi_k \leq 0 \\[3pt]
    \cos\chi_k \,/\, 0.05          & 0 < \cos\chi_k < 0.05 \\[3pt]
    1                              & \cos\chi_k \geq 0.05
  \end{cases}
\end{equation}
The ramp $r(\cos\chi_k)$ scales the daytime value linearly across a
$\sim$3\textdegree\ solar-elevation band ($\cos\chi_k \in [0,
0.05]$ corresponds to elevation $\in [0\degree, 2.87\degree]$).
The wall-clock duration of that 3\textdegree\ band depends on
latitude: the elevation rate at the horizon is
$\sim$15\textdegree/h at the equator (90\textdegree\ sun arc over
6\,h), $\sim$10\textdegree/h at midlatitudes,
$\sim$3\textdegree/h at $|\phi| = 60\degree$ in winter, and
${\to}\,0$ at the polar circles around solstice --- so the
twilight ramp spans about 12\,min at the equator, 17\,min at
midlat, an hour at high winter latitude, and (asymptotically)
the entire civil-twilight period inside the polar circles.  The
ramp matches the gentle climb of D-region photoionisation through
twilight at any latitude rather than snapping on at the geometric
terminator.

The log-flux driver is in units of dB at \SI{7}{\MHz}, calibrated so
M1 produces \SI{4}{\dB}, X1 produces \SI{12}{\dB}, and X10 saturates
at \SI{20}{\dB}:
\begin{equation}\label{eq:lflare-driver}
  D(X) = \max\!\Bigl(0,\;
    \min\!\bigl(\SI{20}{\dB},\;
      \SI{4}{\dB} + \SI{8}{\dB}\cdot\log_{10}\!\tfrac{X}{\text{M1}}\bigr)\Bigr)
\end{equation}
where $X/\text{M1}$ is the X-ray flux ratio to the M1 baseline
($10^{-5}\,\text{W/m}^{2}$).

\paragraph{Calibration source.}  The anchors (\SI{4}{\dB} at M1,
\SI{12}{\dB} at X1, \SI{20}{\dB} at X10) follow the classical
SID-absorption framework of Mitra~\cite{mitra1974}
(\emph{Ionospheric Effects of Solar Flares}, the canonical
print-only treatment of D-region absorption versus soft X-ray
peak flux), with the slope set to \SI{8}{\dB}/decade between M1
and X10 so the anchor points fit the per-decade ${\sim}8\,\si{\dB}$
scale that the riometer-record literature consistently
reproduces.  The specific M1=4 / X1=12 / X10=20 dB values are
not directly cross-checkable against an open digital source;
the WSPR-loss validation below is what anchors them in
ionocast. The values were cross-checked against WSPR loss
patterns observed during cycle-25 M-class and X-class events (the
wspr-live archive of 2024--2026 flare hours; see
\S\ref{sec:calibrate}): on 20\,m and 17\,m WSPR spot drops of
5--12\,dB versus the prior quiet-day median tracked the predicted
$D(X) \cdot (7/f)^{1.5}$ envelope to within ${\sim}\SI{2}{\dB}$
across M1 through X4. (No cycle-25 X10 events have occurred at
the time of writing, so the X10 anchor is set by the literature
extrapolation alone, hence the \SI{20}{\dB} hard saturation in
the equation rather than an open-ended slope.)

The outer $\max(0, \cdot)$ wraps the saturated $\min(20, \cdot)$
branch and suppresses contributions whose underlying log-driver
crosses zero, which happens at $X \approx 10^{-5.5}\,\text{W/m}^{2}$
(approximately C3.16); this is not an explicit ``if X $<$ C3
return 0'' gate but the practical effect is that the quietest
C-class events (C1--C3) contribute
$\sim$0\,dB at the 7\,MHz reference, while higher C-class flares
ramp up the contribution as the algebra allows:
$D(\text{C5}) \approx \SI{1.6}{\dB}$,
$D(\text{C9}) \approx \SI{3.6}{\dB}$,
$D(\text{M1}) = \SI{4}{\dB}$ by calibration construction (these
are at the 7\,MHz reference; the band-scaling factor
$(7/f)^{1.5}$ raises them on lower bands so a C5 event still
contributes a few dB on 80\,m / 160\,m).  The clamp removes only
the lowest of the C-class background.  As with PCA, sunlit-hop
contributions are summed and
the path total is capped (\SI{50}{\dB}).  The instantaneous-proxy
approximation omits the roughly 10--60\,minute exponential decay
of the SID, which is acceptable given the model's 10-minute
cadence; the next refresh picks up the decayed class.

\paragraph{Two flare thresholds, two distinct roles.}  The
physics zero-crossing at C3.16 (this section) suppresses the
quiet-day background contribution but is otherwise continuous;
the alert-banner threshold at M1 / M5 / X
(Table~\ref{tab:soft-alerts}) is escalation-grade signaling for
the operator UI.  The two thresholds are not redundant -- one is
the algebraic point where the physics term goes to zero, the
other is the operator-facing ``this is worth surfacing
top-of-page'' line.  Quiet-time C-class flares contribute
sub-dB absorption to the budget without firing a banner.

% ── 3.5 Auroral absorption ──────────────────────────────────────────
\subsection{Auroral Absorption (CGM-Gated)}\label{sec:laur}

Auroral absorption activates when the corrected geomagnetic
(CGM) latitude $|\phi_{\text{CGM}}|$ at a reflection point is
at or above the storm-dependent threshold.  In one line, the
per-hop absorption is
$L_{\text{aur}} = \min(30, D \cdot (30/f_{\text{MHz}}) \cdot
c(\phi_{\text{CGM}}))$ where $D$ is the storm driver
(Eq.~\ref{eq:laur}'s outer term, derived below from
$K_p^{\text{eff}}$ and OVATION HP), $30/f_{\text{MHz}}$ is the
$f^{-1}$ band scaling, and $c(\phi_{\text{CGM}})$ is the
CGM-edge ramp (Eq.~\ref{eq:cphi}).  Per-hop cap 30\,dB
(\texttt{AUR\_PER\_HOP\_CAP\_DB}); path cap 50\,dB
(\texttt{PATH\_AUR\_CAP\_DB}); the rest of this section
unpacks each piece.
\begin{equation}\label{eq:laur}
  L_{\text{aur}} = \min\!\left(30,\; D \cdot \frac{30}{f_{\text{MHz}}}
                                    \cdot c(\phi_{\text{CGM}})\right)
\end{equation}
where
\begin{equation}
  D = \max\!\Bigl(\,
    \max(0,\,5(K_p^{\text{eff}} - 4)),\;\max(0,\,\tfrac{\text{HP}_{\text{GW}} - 50}{5})
  \Bigr)
\end{equation}
ramps continuously: the $K_p^{\text{eff}}$ term begins
contributing at $K_p^{\text{eff}} > 4$ and reaches $D = 5$ at
$K_p^{\text{eff}} = 5$ (the G1 storm threshold), which maps to
\SI{5}{\dB} of $L_{\text{aur}}$ at the 30\,MHz reference
frequency $f$ in Eq.~\ref{eq:laur} but to substantially more at
HF: the $30/f_{\text{MHz}}$ band-scaling factor turns $D = 5$
into ${\sim}\SI{21}{\dB}$ on 40\,m ($5 \cdot 30/7$) and
${\sim}\SI{11}{\dB}$ on 20\,m ($5 \cdot 30/14.1$), before the
per-hop \SI{30}{\dB} cap and the $c(\phi)$ CGM ramp.  The $f^{-1}$
band scaling here is shallower than the $f^{-1.5}$ used for
quiet-day D-region (\S\ref{sec:labsd}), PCA (\S\ref{sec:lpca}),
and flare (\S\ref{sec:lflare}).  This is empirical rather than
purely theoretical: auroral precipitation has a broad
energy-and-altitude footprint (electrons across 1--100\,keV
deposit across both D and lower E regions, integrating multiple
collision-frequency regimes), and the high-latitude HF riometer
record across $7$--$\SI{30}{\MHz}$ fits a slope closer to
$f^{-1}$ than to $f^{-1.5}$.  PCA and flare absorption, by
contrast, are dominated by D-region deposition at one
relatively narrow altitude band (60--80\,km), where the
collisional $f^{-1.5}$ form is the closer match.
$K_p^{\text{eff}}$ is the storm-lagged effective index
(\S\ref{sec:kplag}), built from the 3-hour planetary $K_p$
history augmented by $D_{\text{st}}$ and IMF $B_z$ early-warning
bumps; references to bare $K_p$ in this section refer to the
underlying 3-hour SWPC sensor that feeds $K_p^{\text{eff}}$.
The OVATION hemispheric-power term contributes $\max(0,
(\text{HP} - 50)/5)$ above the \SI{50}{\GW} floor: $D = 0$ for
HP $\le \SI{50}{\GW}$ (the formula is wrapped in $\max(0,
\cdot)$, so quiet-aurora HP values below the floor return zero
rather than a negative driver), $D = 1$ at HP $= \SI{55}{\GW}$,
$D = 5$ at HP $= \SI{75}{\GW}$ (same $D$-scaling as the
$K_p^{\text{eff}}$ term). The two drivers are combined via $\max(\cdot)$ rather
than additively because they are not independent --- both are
forced by the same auroral-electrojet activity, and adding them
would double-count the underlying ionisation; but they are kept
as parallel sensors because they sample the same phenomenon on
different timescales.  The 3-hour $K_p$ that feeds
$K_p^{\text{eff}}$ is reported every 3\,h, so a sub-hour
auroral intensification driven by a fresh substorm onset is
invisible to that channel for up to 90\,minutes unless one of
the $D_{\text{st}}$ / $B_z$ early-warning bumps fires; OVATION
HP is updated every ${\sim}30$\,min and reflects the actual
hemispheric power dissipated in the auroral oval over the most
recent half hour.  The $\max(\cdot)$ takes whichever sensor is
reading hotter at any given moment, so a fast-onset substorm
shows up via HP and a multi-hour storm in steady state shows up
via $K_p^{\text{eff}}$ --- catching both regimes without
double-counting.

The continuous ramp lets the prediction reflect modest auroral
absorption already at active-storm ($K_p^{\text{eff}} \in [4, 5)$)
levels rather than waiting for a hard storm threshold; a hard
$K_p^{\text{eff}} \geq 5$ / $\text{HP} \geq \SI{50}{\GW}$ gate
would produce a \SI{21}{\dB} step in $L_{\text{aur}}$ on 40\,m
polar paths at the threshold instant, the same family of cliff
the $c(\phi)$ ramp below avoids on the CGM edge.
$c(\phi_{\text{CGM}})$ is a smooth ramp on the CGM-edge:
\begin{equation}\label{eq:cphi}
  c(\phi) = \begin{cases}
    0 & |\phi| \le \phi_{\text{thr}} - 5\text{\textdegree} \\
    \dfrac{|\phi| - (\phi_{\text{thr}} - 5\text{\textdegree})}{5\text{\textdegree}}
        & \phi_{\text{thr}} - 5\text{\textdegree} < |\phi| < \phi_{\text{thr}} \\
    1 & |\phi| \ge \phi_{\text{thr}}
  \end{cases}
\end{equation}
Linear ramp over the trailing $5\text{\textdegree}$, full effect
poleward of the threshold. Eliminates the hard cliff that the bare
gate produced.

\paragraph{Storm-zone expansion.} The CGM threshold ramps linearly
from $60\text{\textdegree}$ to $50\text{\textdegree}$ as
$K_p^{\text{eff}}$ climbs from 5 to 7, then holds at
$50\text{\textdegree}$ above $K_p^{\text{eff}} = 7$:
\begin{equation}\label{eq:auroralexpansion}
  \phi_{\text{thr}}(K_p^{\text{eff}}) = 60\text{\textdegree} -
    10\text{\textdegree} \cdot \min\!\Bigl(1,\,\max(0,\,(K_p^{\text{eff}} - 5)/2)\Bigr)
\end{equation}
This captures the equatorward expansion of the auroral oval during
storms.  Both sides of the auroral term (geometric extent and
absorption depth) respond on the same thermospheric timescale, so
a single $K_p^{\text{eff}}$ argument enters both this expansion
and the magnitude $D$ in Eq.~\ref{eq:laur}'s outer term.  The
linear ramp from $60\degree$ to $50\degree$ over the
$K_p^{\text{eff}} \in [5, 7]$ window lets the expansion start
gradually during moderate storms (G2 / G3) rather than snapping
at $K_p^{\text{eff}} = 7$, which would otherwise produce a
30+\,dB jump on a hop near $55\degree$ CGM at the threshold
instant.  The expansion is wired into the absorption term itself;
no external runtime gating is needed.

\paragraph{Per-hop evaluation.} On a multi-hop path,
$L_{\text{aur}}$ is computed per reflection point and summed
(capped at \SI{30}{\dB} per hop via
\texttt{AUR\_PER\_HOP\_CAP\_DB} and \SI{50}{\dB} total path
via \texttt{PATH\_AUR\_CAP\_DB}) by the same machinery as the
per-hop D-region sum (Eq.~\ref{eq:labsd-path}). A 4-hop
transatlantic path with the midpoint at \SI{55}{\textdegree}
CGM but two of its hops at \SI{65}{\textdegree} correctly
accrues auroral absorption on the polar hops; a midpoint-only
gate would miss it.  \emph{Legacy fallback:} when the budget
is invoked without source / destination geometry (callers that
supply only \texttt{midLat}, \texttt{cgmLatAbsValue}, no
src/dst pair --- e.g.\ smoke tests, the local-MUF fallback
branch in \texttt{src/derive/conditions.js} with no kc2g
paths), the auroral term degrades to a single-midpoint
evaluation against the supplied \texttt{cgmLatAbsValue}
without per-hop summation.  The per-hop branch is the
production path; the midpoint fallback is documented here so
a reader auditing the smoke-test budgets does not mistake
it for a regression.

\paragraph{Storm-main amplification (forward reference).}
Eq.~\ref{eq:laur} states the steady-state $L_{\text{aur}}$.  During
the \emph{main} phase of a geomagnetic storm
($\text{Dst}\le-50\,\text{nT}$ with $K_p^{\text{eff}}$ still
loading), \SI{+4}{\dB} is added to the per-hop $L_{\text{aur}}$
(then re-clamped against the \SI{30}{\dB} per-hop cap from
\texttt{AUR\_PER\_HOP\_CAP\_DB}) to capture the extra
particle-precipitation absorption that the steady-state
$K_p$-keyed form undercounts; recovery and quiet phases keep the
steady-state value.  The amplification is detailed together with
the storm-phase classifier in Section~\ref{sec:reliability}
(paragraph ``Storm-phase auroral amplification''), since both
the classifier and the $\sigma_{\text{recovery}}$ widening it
gates are defined there.  A reader computing the budget from
this section alone should add the \SI{+4}{\dB} bump when the
phase classifier reports \emph{main} and re-apply the per-hop
cap.

The CGM latitude is computed via a tilted-dipole approximation:
\begin{equation}\label{eq:cgm}
  \phi_{\text{CGM}} = \arcsin\!\Bigl(\,\text{clamp}\bigl(
    \sin\phi\sin\phi_P + \cos\phi\cos\phi_P\cos(\lambda - \lambda_P),\;
    -1,\, +1\bigr)\Bigr)
\end{equation}
The explicit clamp guards against floating-point overshoot of the
$\arcsin$ argument outside $[-1, +1]$ near the geomagnetic pole
itself, where $\sin\phi\sin\phi_P + \cos\phi\cos\phi_P\cos(\Delta\lambda)$
can drift by ${\sim}10^{-15}$ above $1$ from rounding error and
produce a NaN $\phi_{\text{CGM}}$ that would propagate through the
auroral CGM gate.
with the IGRF~2020~\cite{igrf13} north magnetic pole at
$\phi_P = +80.7\text{\textdegree}$,
$\lambda_P = -72.7\text{\textdegree}$ (signed-east-of-Greenwich
convention; held fixed for reproducibility;
IGRF-14 was published with the 2025.0 epoch but the centred-dipole
pole has drifted sub-degree since 2020, far below the
$\phi_{\text{CGM}}$ ramp's $5\text{\textdegree}$ smoothing width
in Eq.~\ref{eq:cphi} and well below any verdict-shifting threshold.
Centred-dipole drift is ${\sim}0.05\text{\textdegree}$\,/\,yr, so
the held-fixed values remain inside the smoothing width through
$\sim 2030$; review against IGRF-15 at the 2030.0 epoch).  The
operator's QTH longitude is whatever signed value the geolocation
returns (e.g. $-9.14\text{\textdegree}$ for Lisbon, $+139.69\text{\textdegree}$
for Tokyo).  Holding the IGRF pole fixed while accepting
F\textsubscript{10.7A} live looks asymmetric but reflects the
underlying timescales: F\textsubscript{10.7A} (the 81-day mean
of F\textsubscript{10.7}) varies on days to weeks as the centre
of the trailing window slides, so consuming the live value is
correct; the IGRF centred-dipole pole drifts
$\sim$0.05\textdegree/yr (decadal), so freezing it at the 2020
epoch produces sub-degree error well inside the
$\phi_{\text{CGM}}$ ramp's 5\textdegree{} smoothing width and
keeps the bibliography pinned to a citable model
(Alken~et~al.~\cite{igrf13}) rather than tracking a
continuously-updating model file.

% ── 3.6 Multi-hop reflection ────────────────────────────────────────
\subsection{Multi-Hop Ground Reflection Loss}\label{sec:lhop}

Each extra F-region hop adds a ground-reflection plus defocusing
penalty applied at the intermediate ground bounces. To avoid a
discrete cliff at the hop-ceiling boundary, the extra-hop count
is computed as a smooth function of distance:
\begin{equation}\label{eq:lhop}
  L_{\text{hop}} = N_{\text{extra}}(d) \cdot
    \bigl(L_{\text{gr}}(f, \theta) + D\bigr) \quad\text{dB},\qquad
  N_{\text{extra}}(d) = \max\!\left(0,\; \frac{d_{\text{km}}}{d_{\text{hop}}^{\max}(h_F)} - 1\right)
\end{equation}
where the per-hop ground-range ceiling scales as $\sqrt{h_F}$, the
leading-order spherical-Earth result for glancing-incidence
reflection at altitude $h_F$. At zero elevation the geocentric angle
to the reflection-point ground projection is
$\gamma = \arccos\!\bigl(R_\oplus / (R_\oplus + h_F)\bigr)$, giving
$d \approx 2\sqrt{2\,R_\oplus h_F}$ for $h_F \ll R_\oplus$.
ionocast preserves the empirical \SI{4000}{\km} anchor at the
quiet-cycle $h_F = \SI{300}{\km}$ and lets the rest follow the
square-root shape:
\begin{equation}\label{eq:hopceiling}
  d_{\text{hop}}^{\max}(h_F) = \SI{4000}{\km} \cdot
    \sqrt{\frac{h_F}{\SI{300}{\km}}}
\end{equation}
At $h_F = \SI{300}{\km}$ this reduces to $\SI{4000}{\km}$. On
enhanced-solar-flux days $h_{m\!F2}$ climbs to $\SI{340}{\km}$ or
above and the ceiling moves to ${\sim}\SI{4258}{\km}$.  ionocast
uses an observed $h_{m\!F2}$ from the GIRO digisonde panel when
available; otherwise defaults to \SI{300}{\km}.

The per-hop ionospheric loss bundle (Section~\ref{sec:labsd}) is
also continuous across the hop boundary, but in a different way
than the multiplier here.  At a non-integer $n_C = d /
d_{\text{hop}}^{\max}$, the bundle computes the four ionospheric
absorption sums (D-region, PCA, flare, auroral) for both
bracketing integer hop layouts $\lfloor n_C \rfloor$ and
$\lceil n_C \rceil$ --- each laying down a different set of
reflection-point coordinates --- and blends the totals by the
fractional part $\alpha = n_C - \lfloor n_C \rfloor$:
$L = (1 - \alpha) \cdot L_{\lfloor n_C \rfloor} +
     \alpha \cdot L_{\lceil n_C \rceil}$.
At integer $n_C$, $\alpha = 0$ and the ceil branch is skipped.
This keeps both budgets continuous across hop-count boundaries: a
4001\,km path on 80\,m at noon would otherwise jump from
$\sim$18\,dB of D-region absorption (single midpoint) to
$\sim$36\,dB (two reflection points) at the boundary, while the
ground-reflection multiplier $L_{\text{hop}}$ barely changes
($N_{\text{extra}}$ ramping from 0 to 0.0003).

$L_{\text{gr}}(f, \theta)$ is the per-bounce Fresnel reflection
loss and $D = \SI{0.25}{\dB}$ is the per-extra-hop defocusing constant
(earth-curvature spreading; ITU-R P.533 Annex~1 prescribes
${\sim}\SI{1}{\dB}$ as a population mean, but that figure mixes ground
types and elevations and overshoots typical amateur multi-hop paths
over ``average earth'' as established by 30\,d WSPR calibration).

\paragraph{Fresnel ground-reflection loss.}  At grazing angle
$\psi = \theta_{\text{elev}}$ (for a symmetric hop, takeoff equals
grazing at the bounce), with complex relative permittivity
\begin{equation}
  \varepsilon_c = \varepsilon_r - j\,60\,\sigma\,\lambda
  \quad\text{($\lambda$ in metres)}
\end{equation}%
\footnote{The minus sign on the imaginary term assumes the
$e^{-j\omega t}$ time-harmonic convention used by the
\texttt{lHopGroundReflectionDb} implementation in
\texttt{src/physics/loss.js}.  The opposite convention
($e^{+j\omega t}$) flips the sign of the imaginary term but
not of $|R_h|^2$, so the loss in dB is unchanged either way.
Code uses \texttt{eIm = -60 * GROUND\_AVG\_SIGMA * lambdaM},
matching the form printed here.}
and ``average earth'' constants ($\varepsilon_r = 13$,
$\sigma = \SI{0.005}{S/m}$, where $\sigma$ here denotes ground
conductivity, not the per-band prediction spread of
Table~\ref{tab:bandsigma}) following the ARRL Antenna
Book~\cite{arrl2023antenna} amateur-radio convention.  ITU-R
P.527-6~\cite{itup527} Table~1 lists slightly different
canonical values for medium-dry ground ($\varepsilon_r \approx 15$,
$\sigma \approx \SI{0.01}{S/m}$); at HF takeoff angles the
resulting $L_{\text{gr}}$ disagreement is sub-decibel, so the
amateur-convention pair is retained for continuity with the
antenna-modelling literature operators are most likely to
cross-check against.  The horizontal-polarisation Fresnel
coefficient is
\begin{equation}
  R_h = \frac{\sin\psi - \sqrt{\varepsilon_c - \cos^2\psi}}
             {\sin\psi + \sqrt{\varepsilon_c - \cos^2\psi}}.
\end{equation}
The implementation takes the H-pol power reflectance
\begin{equation}\label{eq:lgr}
  L_{\text{gr}}(f, \psi) = -10\log_{10}|R_h|^2,
  \qquad L_{\text{gr}} \in [0, 8]~\si{\dB}.
\end{equation}
Most amateur HF antennas are H-polarised (dipoles, inverted-V, OCF,
EFHW, Yagi/quad arrays), and 30\,d WSPR calibration confirms H-pol
as the right per-bounce model: H-pol Fresnel scored Brier $0.07$
(89\% accuracy) versus a polarisation-averaged
$\tfrac{1}{2}(|R_h|^2 + |R_v|^2)$ alternative at $0.13$ (83\%) and
V-pol at $0.27$ (66\%).  H-pol Fresnel predicts ${<}\SI{1}{\dB}$
per bounce at typical $5$--$15\text{\textdegree}$ grazing angles,
which is what the multi-hop budget needs to see for a 3--5 hop
path to predict ``open'' when the band is in fact open.

The $[0, 8]~\si{\dB}$ clamp brackets the H-pol Fresnel loss
between physically-meaningful bounds.  The lower bound $0$\,dB is
the perfect-grazing limit ($\psi \to 0$) where any real-earth
reflectance approaches unity; in practice the un-clamped formula
returns sub-decibel values at typical $5$--$30\degree$ HF
grazing angles, so the lower bound is rarely approached and is a
guard against numerical edge cases rather than a binding floor.
The upper bound $8$\,dB guards against the single-layer earth
model's behaviour at high (near-normal) incidence angles. Unlike
V-pol, H-pol has no Brewster zero; $|R_h|^2$ is monotonic with
grazing angle, peaking in loss at normal incidence
($\psi = 90\text{\textdegree}$) and decreasing as the geometry
goes more grazing. For ``average earth''
($\varepsilon_r = 13$, $\sigma = \SI{0.005}{S/m}$) on \SI{14}{\MHz}
the normal-incidence loss is ${\sim}\SI{4.7}{\dB}$, comfortably
under the clamp. The cap only binds for very dry / poor ground
($\varepsilon_r \approx 4$), where the single-layer model can
predict normal-incidence loss above \SI{8}{\dB}; a real earth at
those parameters is layered (ground-water profile, soil density
gradient) and the effective reflectance does not collapse that
hard, so the \SI{8}{\dB} cap is the empirical ceiling that keeps
the budget aligned with WSPR observations on those short-haul
paths. The clamp is rarely binding in practice: typical amateur HF
takeoff angles sit at $5$--$30\text{\textdegree}$ grazing
($\psi$ measured from the surface), where the unclamped Fresnel
form itself returns sub-3\,dB losses regardless of ground
parameters.

\paragraph{Typical magnitudes.}  At a representative
$\sim\SI{11144}{km}$ Lisbon--Tokyo path, hop geometry under the
saturating $d_{\text{hop}} = \min(d, d_{\text{hop}}^{\max})$ form
(Section~\ref{sec:antpattern}, Eq.~\ref{eq:takeoff}) yields
$\theta = \arctan(2\cdot 300 / 4000) = 8.5\text{\textdegree}$,
and per-bounce H-pol Fresnel loss at that angle on 14\,MHz with
$\varepsilon_r = 13$, $\sigma = \SI{0.005}{S/m}$ gives
$L_{\text{gr}} \approx \SI{0.66}{\dB}$. Eq.~\ref{eq:lhop}'s
smooth $N_{\text{extra}}(d) = d/d_{\text{hop}}^{\max} - 1$ at
$d_{\text{hop}}^{\max} = \SI{4000}{\km}$ gives
$N_{\text{extra}} = 1.79$, so
$L_{\text{hop}} = 1.79 \cdot (0.66 + 0.25) \approx \SI{1.63}{\dB}$
on the path.  A 5-hop trans-equatorial path approximately doubles
it.  Single-hop NVIS and short-haul paths see $L_{\text{hop}} = 0$
(no intermediate bounces).

\paragraph{Calibration coverage.}  The Fresnel piece is fully
physics-derived; only the defocusing scalar $D$ remains tunable,
and is exercised by the harness over a multi-path basket
spanning 1--5 hops continuously (single-hop NVIS through
5-hop trans-equatorial; see Appendix~\ref{app:paths}'s
geometry/use-case column for the per-path hop counts;
Section~\ref{sec:calibrate}).

\paragraph{Path-type bias (deferred).}  Sea-vs-land per-path
classification is deferred --- ``average earth''
($\varepsilon_r = 13$, $\sigma = \SI{0.005}{S/m}$) applies
everywhere until coastline-aware path geometry is added.  The
direction and magnitude of the bias is known and bounded:
\begin{itemize}[nosep]
  \item Seawater ($\sigma \approx \SI{5}{S/m}$, three orders of
        magnitude more conductive) produces ${\sim}0.3$\,dB
        H-pol Fresnel loss per bounce versus
        ${\sim}\SI{0.66}{\dB}$ on average earth, so
        \emph{transatlantic and transpacific paths are
        under-predicted by ${\sim}1\text{--}3$\,dB depending on
        hop count} (1\,dB on a 2-hop circuit, ${\sim}3$\,dB
        on a 5-hop trans-oceanic).
  \item Dry sand or arid soil ($\sigma \approx \SI{0.001}{S/m}$,
        five times less conductive) produces an extra
        ${\sim}0.5$\,dB H-pol loss per bounce, so
        \emph{paths over the Sahara, Australian desert, or central
        Asia are over-predicted by ${\sim}1\text{--}2$\,dB on a
        3--5 hop budget}.
\end{itemize}
Both biases are operator-relevant on close-margin verdicts where
the budget reads near a tier boundary but small in the open-band
regimes where the verdict is dominated by the F2 / MUF /
absorption terms. Until coastline-aware geometry lands, the UI
shows the average-earth verdict; expect transatlantic / transpacific
paths to perform somewhat better than the verdict suggests, and
desert-dominated paths somewhat worse.

% ── 3.7 Es screening ────────────────────────────────────────────────
\subsection{Sporadic-E Screening Loss}

When a strong Es layer is present ($f_{\text{oEs}} \geq \SI{5}{\MHz}$)
and the operating frequency is below twice the Es critical frequency
($f < 2\,f_{\text{oEs}}$), the layer partially screens the F2
reflection:
\begin{equation}
  L_{\text{Es}} = \SI{5}{\dB}
\end{equation}
The flat \SI{5}{\dB} value is a calibration constant rather than
a derived quantity --- the actual screening loss has a graded
$f$-vs-$f_{\text{oEs}}$ dependence (deeper screening as $f$
approaches $f_{\text{oEs}}$, vanishing at $f = 2\,f_{\text{oEs}}$
where the Es layer becomes transparent), but the operationally
relevant case is the binary ``F2 hop is partially blocked by
this Es layer'' decision, and a single mid-range value matches
the WSPR-validated bias on screened paths better than a
parametrised form would given the foEs measurement uncertainty.
$L_{\text{Es}}$ is a screening penalty applied \emph{only inside
the F2 budget} (Eq.~\ref{eq:budget}). When the parallel Es-mode budget
of Section~\ref{sec:es-prop-mode} wins the per-band margin
competition, that path uses its own loss bundle (Eq.~\ref{eq:esmargin})
and $L_{\text{Es}}$ is not part of its calculation.

% ── 3.8 Low-band extra loss ─────────────────────────────────────────
\subsection{Low-Band Extra Loss}\label{sec:llow}

A band-stepped term applies an empirical residual on the lower HF
bands (Table~\ref{tab:lowbandextra}). $L_{\text{low}}$ is added
\emph{on top of} the per-hop quiet-day D-region absorption
$L_{\text{Dreg}}$ (already \SI{36.6}{\dB} on 160\,m at local noon,
etc., per Eq.~\ref{eq:labsd}), not as a substitute for it. The
mechanisms it represents are propagation residuals that
$L_{\text{Dreg}}$'s mean-conditions formulation does not capture:
\begin{itemize}[nosep]
  \item Higher-angle refraction at low frequencies, which forces
        steeper takeoff angles and longer in-ionosphere ray paths
        than the geometric hop calculation assumes.
  \item Magneto-ionic mode coupling on F2 reflection (E-trap and
        ordinary/extraordinary splitting), which is stronger on
        low bands and is not modelled mechanistically in the
        current Fresnel/multi-hop bundle (Section~\ref{sec:lhop}).
  \item Ground-wave contamination on short-haul paths, where the
        sky-wave is in destructive interference with a residual
        ground-wave component that the F2-only budget does not
        account for.
\end{itemize}
The values in Table~\ref{tab:lowbandextra} are hand-set integers
chosen by inspection of the harness's per-band Brier residual on
the low bands against 30\,d of WSPR
observations (\S\ref{sec:calibrate}); $L_{\text{low}}$ is
\emph{not} part of the joint coordinate-descent set in
Appendix~\ref{app:repro} ($\{L_{\text{iono}}, D, w_{\text{sc}},
w_{\text{NVIS-tail}}, w_{\text{Es-prim}}, \sigma\text{-scale},
\text{fusion-flag}\}$), and the round-integer values (8, 5, 3,
2\,dB on 160\,m through 40\,m) are the visible signature of
that; 30\,m carries a half-step taper at 0.5\,dB to give the
log-frequency interpolation a non-zero anchor without
re-introducing a step at the 30/20\,m cutoff.  Bounds are guided by ITU-R
P.533~\cite{itup533} Annex~1's median-conditions correction term
$Y_p$ which the current budget otherwise omits
(Section~\ref{sec:liono} discusses why $Y_p$ is not included).
A future calibration pass should fold $L_{\text{low}}$ into the
descent set once the per-pair WSPR ground-truth wiring stabilises;
until then, the table is empirical-judgment rather than
machine-fitted.
The values are path-independent --- a per-distance refinement is
deferred until enough per-pair WSPR data accumulates to fit a
distance-dependent ramp without inflating noise.

\begin{table}[H]
\centering
\caption{Low-band extra loss term applied below 30\,m.  Empirical
residual on top of $L_{\text{Dreg}}$, calibrated against 30\,d
WSPR; see~\S\ref{sec:llow} for the mechanism breakdown.  Stored as
anchor points for a smooth log-frequency interpolation (refactored
2026-04-30 from a step-function form that had hard cutoffs at
$\SI{2}{\MHz}$, $\SI{4}{\MHz}$, $\SI{6}{\MHz}$, $\SI{8}{\MHz}$);
intermediate frequencies linearly interpolate between the
band-centre anchors in $\log f$ space.  The $\SI{14}{\MHz}$ anchor
holds the value at zero so the contribution tapers smoothly
through 30\,m and is exactly zero at 20\,m and above.}
\label{tab:lowbandextra}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Band} & {$f$ (\si{\MHz})} & {$L_{\text{low}}$ (\si{\dB})} \\
\midrule
160\,m & 1.838 & 8.0 \\
80\,m  & 3.570 & 5.0 \\
60\,m  & 5.366 & 3.0 \\
40\,m  & 7.040 & 2.0 \\
30\,m  & 10.140 & 0.5 \\
20\,m + & 14.000+ & 0.0 \\
\bottomrule
\end{tabular}\\[4pt]
\footnotesize
The 30\,m cutoff is empirical: at and above 30\,m the
$L_{\text{Dreg}}$ baseline (Eq.~\ref{eq:labsd}) plus the modeled
F2-region terms account for the band's quiet-day budget within
the harness's residual resolution, so $L_{\text{low}}$ adds
nothing.  Below 30\,m the residuals were systematic enough to
warrant the empirical correction.
\end{table}

% ── 3.9 Lumped ionospheric loss ──────────────────────────────────────
\subsection{Lumped Ionospheric Loss}\label{sec:liono}

\begin{equation}
  L_{\text{iono,path}} = n_{\text{hops}} \cdot L_{\text{iono}},
  \qquad L_{\text{iono}} = \SI{1}{\dB}
\end{equation}
This is the ITU-R P.533 median-conditions residual once the
individually-modelled mechanisms are subtracted: spatial
focusing/defocusing $L_z$ is now handled per intermediate hop by
$D$ (Section~\ref{sec:lhop}), polarisation coupling is now handled
inside the Fresnel $R_h$ branch of the multi-hop reflection model,
and the only term that remains a-priori invisible to the geometry is
the ${\sim}\SI{1}{\dB}$ Faraday-rotation / small-correction residual
\emph{per ionospheric reflection}.  Per ITU-R P.533 §A.2 the lumped
correction accumulates per reflection, so the path total scales with
hop count: a 1-hop NVIS path gets \SI{1}{\dB}, a 3-hop transcontinental
path gets \SI{3}{\dB}, etc.  The 1-dB magnitude is calibrated
against the production 35-path basket (Appendix~\ref{app:paths}),
which spans single-hop NVIS through 5-hop trans-equatorial
geometries; the per-hop multiplier closes the residual on
multi-hop DX paths where a single-charge convention would
under-count by 2--4 dB.
$L_{\text{iono}}$ does \emph{not} include D-region absorption
(handled per-hop by Eq.~\ref{eq:labsd-path}) or ground reflection
(Section~\ref{sec:lhop}), nor does it include the
above-median-reliability $Y_p$ margin
(${\approx}\SI{7}{\dB}$ at 90\% reliability) defined by ITU-R
P.842~\cite{itup842} (the recommendation that prescribes the
circuit-reliability framework around the per-band $\sigma$
fed by P.533~\cite{itup533}): $Y_p$ is the correct term when
the prediction answers
``will the circuit work 90\% of the time at this hour?'', but ionocast
answers ``is the band open right now?'', which is a median-conditions
question.  The $\SI{1}{\dB}$ value is what remains after each
mechanism formerly bundled into the lumped term ($Y_p$
above-median margin, multi-hop bounce loss, polarisation
coupling) was relocated to its own physics-derived per-hop
expression and the residual recalibrated against 30\,d of WSPR
truth.

% ── 3.10 Noise model ────────────────────────────────────────────────
\subsection{Noise Model (ITU-R P.372)}\label{sec:noise}

Following ITU-R P.372~\cite{itup372}, the received noise power is
the power sum of the natural floor (galactic plus atmospheric)
and a man-made contribution that fires only on suburban and
urban sites.  We write power-sum addition with the symbol
$\oplus$ defined by $A \oplus B \equiv 10\log_{10}\!\bigl(
10^{A/10} + 10^{B/10}\bigr)$ throughout this section, with the
indicator-gated form spelled out in full as the noise budget:
\begin{equation}\label{eq:noise}
  N = 10\log_{10}\!\left(10^{N_{\text{atmo}}/10}
                       + \mathbb{1}_{\text{env}\ne\text{rural}}\cdot
                         10^{N_{\text{mm}}/10}\right)
\end{equation}
where:
\begin{align}
  N_{\text{atmo}} &= N_{\text{base}}(f) + \Delta N_{\text{diurnal}}(f,\cos\chi) \\
  N_{\text{mm}}   &= N_{\text{base}}(f) + F_a(\text{env})
\end{align}
The indicator $\mathbb{1}_{\text{env}\ne\text{rural}}$ gates the
man-made channel off entirely on rural sites; on suburban / urban
sites the channel sits at an environment-specific level
$N_{\text{base}} + F_a$ that is roughly diurnally constant (man-made
QRM tracks the local interference environment, not the natural
floor's day/night variation).  The diurnal swing is:
\begin{equation}
  \Delta N_{\text{diurnal}} = -A(f) \cdot \text{clamp}(\cos\chi,\,-1,\,1)
\end{equation}
where $A(f)$ ramps linearly from \SI{10}{\dB} at $f \leq \SI{5}{\MHz}$
down to \SI{3}{\dB} at $f \geq \SI{15}{\MHz}$:
\begin{equation}
  A(f) = \begin{cases}
    10 & f \le \SI{5}{\MHz} \\
    10 - 7\,(f - 5)/10 & \SI{5}{\MHz} < f < \SI{15}{\MHz} \\
    3  & f \ge \SI{15}{\MHz}
  \end{cases}
\end{equation}
Galactic-vs-atmospheric dominance transitions gradually across the
\SI{5}{\MHz}--\SI{15}{\MHz} window (galactic noise from P.372
Figs.~2--3 vs atmospheric from P.372 Figs.~13--14 / 23--24,
which tabulate atmospheric noise factor $F_{am}$ for the four
seasonal-hemispheric cuts) rather than at any single frequency,
so a smooth ramp is consistent with the smooth-handoff philosophy
applied elsewhere in the budget.
The man-made noise factor $F_a$ depends on environment
(rural / suburban / urban classification follows ARRL
Handbook~\cite{arrl2023handbook} convention):
\emph{rural} carries no man-made contribution above the natural
floor (the indicator switches the second power-sum term off),
\emph{suburban} sets $F_a = \SI{+15}{\dB}$ above the rural baseline
$N_{\text{base}}$, and \emph{urban} sets $F_a = \SI{+25}{\dB}$.
P.372 reports its $F_a$ relative to the thermal floor $kT_0$ (290~K)
rather than relative to a quiet-rural baseline.  Per P.372's
linear man-made-noise fit $F_a = c - d\log_{10}(f_{\text{MHz}})$,
quiet-rural takes $c = 53.6$, $d = 28.6$, giving $F_a \approx
\SI{25}{\dB}$ above thermal at \SI{10}{\MHz}; the galactic
floor at 10\,MHz with $c = 52$, $d = 23$ contributes $F_a
\approx \SI{29}{\dB}$, so the noise channel an HF
quiet-rural operator actually sees is the power-sum of these
two components ($\sim\!\SI{30}{\dB}$ at 10\,MHz, dominated by
the galactic term in the upper HF).  ionocast's
$+\!\SI{15}{\dB}$ suburban and $+\!\SI{25}{\dB}$ urban add to
the lumped quiet-rural floor to give ${\sim}\SI{45}{\dB}$ and
${\sim}\SI{55}{\dB}$ above thermal, landing inside the P.372
$30$--$80$\,dB envelope spanning quiet-rural through
industrial.  The amplitude $A$ is
held latitude-independent for simplicity; tropical sites with
frequent thunderstorm activity see larger atmospheric-noise diurnal
swings (the P.372 atmospheric maps exceed midlat figures by
5--10\,dB in the equatorial band) and polar sites with stable
nighttime ionospheres see smaller swings, so the model is somewhat
low for tropical noise floors at night and somewhat high for polar
nights.  The error is bounded at 2--3\,dB on the noise side and is
dominated by other budget terms on most verdicts.

\paragraph{Why the rural-site indicator gates $N_{\text{mm}}$.}
On rural sites the indicator $\mathbb{1}_{\text{env}\ne\text{rural}}$
gates the man-made channel off entirely so $N = N_{\text{atmo}}$
exactly.  This lets $\Delta N_{\text{diurnal}}$ modulate the
floor in full ($-124\,\si{\dBm}$ at 15\,m noon, $-118\,\si{\dBm}$
at 15\,m midnight; $N_{\text{base}} = -121\,\si{\dBm}$ from
Table~\ref{tab:noisebase} modulated by the $A(f) = \SI{3}{\dB}$
diurnal swing on the upper-band branch of $A(f)$) rather than
being clamped by an
unconditionally-active $N_{\text{mm}}$ that would otherwise pin
the rural floor near $N_{\text{base}}$ regardless of time of
day.  Suburban and urban verdicts are unaffected: $N_{\text{mm}}$
dominates the power sum on those sites, so the indicator's
all-environment gating is functionally rural-only.

\paragraph{Mapping to ITU-R P.372 figures.}  $N_{\text{base}}(f)$ in
Table~\ref{tab:noisebase} is the lumped quiet-rural floor, the power
sum of the galactic noise (P.372 Figs.~2--3, slowly varying with
frequency at $F_a \approx 52 - 23\log_{10}(f_{\text{MHz}})$\,dB
above thermal in the 10--250\,MHz range) and the atmospheric
quiet-day baseline (P.372 Figs.~13--14 / 23--24, the seasonal
$F_{am}$ atmospheric-noise maps; the quiet-rural midlat midnight
summer reference is read from these maps). The diurnal swing $\Delta N_{\text{diurnal}}$ then
modulates this combined floor by the atmospheric component's
day-vs-night spread (the galactic component is nearly diurnally
constant; below 10\,MHz the atmospheric swing dominates so the
$\SI{10}{\dB}$ amplitude is appropriate, above 10\,MHz galactic
dominates and the amplitude shrinks to $\SI{3}{\dB}$). The man-made
channel $N_{\text{mm}}$ sits $F_a$ dB above the lumped rural
baseline $N_{\text{base}}$ on suburban and urban sites and is held
diurnally constant (man-made QRM is set by the local interference
environment, not by the natural floor's diurnal swing); on rural
sites the channel is gated off entirely
(Eq.~\ref{eq:noise}'s indicator), so a quiet rural QTH reads
exactly $N_{\text{atmo}}$ rather than the
$N_{\text{atmo}}\oplus N_{\text{base}}$ phantom-3\,dB the earlier
form produced. The power-sum then correctly recovers the dominant
component: man-made on suburban / urban, the natural floor on
rural (atmospheric-dominated on low bands, galactic-dominated on
upper bands).

\paragraph{Receiver-bandwidth scaling.}
The two noise components scale differently when the receiver narrows
from the reference bandwidth $B_0 = \SI{2500}{\Hz}$ to the mode's
effective bandwidth $B_{\text{mode}}$ (Section~\ref{sec:modebw}):
\begin{align}
  N_{\text{atmo}}(B_{\text{mode}}) &= N_{\text{atmo}}(B_0) + 10\log_{10}(B_{\text{mode}}/B_0) \\
  N_{\text{mm}}(B_{\text{mode}})   &= N_{\text{mm}}(B_0)   + \alpha \cdot 10\log_{10}(B_{\text{mode}}/B_0)
\end{align}
Atmospheric (galactic, thermal, distant lightning) is quasi-white and
collapses in full with the receiver filter.  Man-made noise is
impulse-dominated: a bright transient in a \SI{2500}{\Hz} filter is
mostly the same transient in a \SI{50}{\Hz} filter, since impulses
have broad spectra.  The scaling exponent $\alpha$ interpolates
between the ideal-white limit ($\alpha = 1$) and the pure-impulsive
limit ($\alpha = 0$); ionocast uses $\alpha = 0.5$ as a
mixed-source-typical compromise.  The Middleton Class A/B
impulse-noise model~\cite{middleton1977} gives a non-linear
bandwidth-scaling form that does not reduce to a single
power-law exponent in general, so $\alpha = 0.5$ is a parametric
simplification of a non-parametric scaling --- defensible at the
prediction level but not derivable as a midpoint of the Middleton
form itself.
The 0.5 value is a compromise tuned for typical mixed broadband
QRM (switching power supplies, motor controllers, plasma TVs,
LED drivers, ignition transients):
\emph{continuous} interferers (RTTY signals, broadcast bleed-over,
local switching mode power-supply hum) sit closer to $\alpha = 1$
and are slightly under-rewarded by the narrowband-filter gain;
\emph{purely impulsive} interferers (lightning crashes, isolated
arcing) sit closer to $\alpha = 0$ and are slightly over-rewarded.
A site dominated by one mode could in principle be tuned with a
different $\alpha$, but that level of per-site characterisation is
out of scope for the current model.

\paragraph{FT8-vs-SSB advantage --- numeric walkthrough.}
The end-to-end model produces about $\SI{31}{\dB}$ advantage on
rural and $\SI{22}{\dB}$ on suburban for FT8 over SSB, evaluated
on a midlat 20\,m path.  Decomposition: SSB and FT8 use mode-native
$S_{\text{req}}$ values from Table~\ref{tab:modesnr} ($+10$ vs
$-4\,\dB$), so the decoder-side advantage is exactly $\SI{14}{\dB}$.
Bandwidth scaling on a rural site (galactic / atmospheric noise
dominant, $\alpha = 1$) gives $10\log_{10}(2500/50) \approx
\SI{17}{\dB}$ of additional gain for FT8's narrower receive
bandwidth, totalling $\SI{31}{\dB}$.  On a suburban site
(man-made-dominated, $\alpha = 0.5$) the noise-side gain falls to
$\sim\SI{8.4}{\dB}$, totalling $\sim\SI{22}{\dB}$.  Anecdotal
operator-experience figures (no specific citation) of
$\sim\SI{25}{\dB}$ rural / $\sim\SI{15}{\dB}$ suburban
circulate in the WSJT-X / ham-radio community as floor-of-
experience benchmarks; these compress several effects --- partial
decoder calibration loss at very low SNR, RX-chain front-end
variability, the fact that operators typically compare the
modes at the noise-floor edge rather than deep in the decode
region, etc.~--- that the model does not attempt to reproduce.  The computed numbers are larger than
operator-reported because the model is the maximum-information
budget; treat the reported $\sim\!25 / \sim\!15$ figures as
floor-of-experience benchmarks, not as a target for the budget.

Table~\ref{tab:noisebase} gives the baseline noise power at
\SI{2.5}{\kHz} reference bandwidth.

\begin{table}[H]
\centering
\caption{Baseline noise $N_{\text{base}}$ from ITU-R P.372 at
         \SI{2.5}{\kHz} bandwidth.  Noon-time floor; midnight =
         $N_{\text{base}} + \Delta N_{\text{diurnal}}(\cos\chi=-1)$
         where the swing amplitude $A(f)$ is \SI{10}{\dB} at
         $f \le \SI{5}{\MHz}$ ramping linearly to \SI{3}{\dB} at
         $f \ge \SI{15}{\MHz}$.  The midnight column assumes the
         path midpoint is at the solar antipode ($\cos\chi = -1$),
         which obtains at the equator at midnight summer; at
         high latitudes in winter the sun never reaches the
         antipode in the diurnal cycle so the actual midnight
         $\cos\chi$ is shallower than $-1$ and the swing is
         attenuated proportionally.  Re-derived 2026-04-30 from
         P.372-15 Fig 13 (atmospheric) $\oplus$ Fig 23 (galactic)
         max-of at midlat midnight summer.}
\label{tab:noisebase}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Band} & {$N_{\text{base}}$ (\si{\dBm})} & {midnight (\si{\dBm})} \\
\midrule
160\,m & -100 & -90  \\
80\,m  & -113 & -103 \\
60\,m  & -118 & -108 \\
40\,m  & -121 & -112 \\
30\,m  & -120 & -114 \\
20\,m  & -120 & -116 \\
17\,m  & -120 & -117 \\
15\,m  & -121 & -118 \\
12\,m  & -122 & -119 \\
10\,m  & -123 & -120 \\
\bottomrule
\end{tabular}
\end{table}

The table is anchored to the P.372 midnight midlat-summer
reference at every band; calibration history and the WSPR-residual
validation that led to the current values are summarised in
Section~\ref{sec:limits}.

% ── 3.11 Required SNR ───────────────────────────────────────────────
\subsection{Required SNR and Receiver Bandwidth by Mode}\label{sec:modebw}

Each mode is defined by two numbers: the decoder-only SNR threshold
$S_{\text{req}}$ specified at the mode's own effective noise bandwidth
$B_{\text{mode}}$, and that bandwidth itself.  The receiver-bandwidth
advantage of narrow digital modes is not baked into $S_{\text{req}}$;
it is computed separately in the noise model (Section~\ref{sec:noise}),
which scales atmospheric noise with the receiver BW in full and
man-made noise with it only partially.  Quoting modes at a common
\SI{2.5}{\kHz} reference BW with the BW advantage folded into
$S_{\text{req}}$ would over-count the benefit of digital modes on
man-made-dominated (impulse-noise) sites.

\begin{table}[H]
\centering
\caption{Required SNR threshold and effective noise bandwidth by
         mode.  $S_{\text{req}}$ is the decoder threshold at the
         mode's own native bandwidth.  Digital-mode rows (FT4 /
         FT8 / WSPR) derive from WSJT-X~\cite{k1jt} published
         values at 2.5\,kHz reference via
         $S_{\text{native}} = S_{2500} + 10\log_{10}(2500 / B_{\text{mode}})$;
         the SSB and CW rows are operator-practice native values
         from the ARRL
         Handbook~\cite{arrl2023handbook} (comfortable-copy SSB
         at +10\,dB SNR, trained-ear CW at $\sim$$-7$\,dB at
         2.5\,kHz).  The bandwidth advantage of narrow modes is
         then computed separately in the noise model
         (Section~\ref{sec:noise}) using $B_{\text{mode}}$, not
         folded into $S_{\text{req}}$.}
\label{tab:modesnr}
\begin{tabular}{@{}lrrr@{}}
\toprule
\textbf{Mode} & {$S_{\text{req}}$ (\si{\dB})} & {$B_{\text{mode}}$ (\si{\Hz})} & {Source} \\
\midrule
SSB  & $+10$ &  2500 & ARRL Handbook comfortable-copy native ($S_{2500} = +10$, $\Delta = 0$) \\
CW   &  $0$  &   500 & ARRL Handbook trained-ear native ($S_{2500} \approx -7$; $-7 + 10\log_{10}(2500/500) = 0$) \\
FT4  & $-2$  &   100 & WSJT-X $-16.4$ at 2500 Hz ($-16.4 + 10\log_{10}(2500/100) = -2.4$, rounded) \\
FT8  & $-4$  &    50 & WSJT-X $-21$ at 2500 Hz ($-21 + 10\log_{10}(2500/50) = -4$) \\
WSPR & $-8$  &    25 & WSJT-X $-28$ at 2500 Hz ($-28 + 10\log_{10}(2500/25) = -8$) \\
\bottomrule
\end{tabular}\\[4pt]
\footnotesize
$B_{\text{mode}}$ here is the decoder \emph{noise-equivalent
bandwidth} integrated against the WSJT-X published threshold,
not the signal's occupied spectral width.  WSPR's 4-FSK
signal~\cite{wsprnet} occupies $\sim$\SI{6}{\Hz} ($1.46$\,Hz/symbol $\times$ 4 tones);
the listed \SI{25}{\Hz} captures the matched-filter / decoder
integration over symbol-time that WSJT-X's $-28$\,dB / 2500\,Hz
threshold already integrates against.  FT8's 50\,Hz signal is a
closer match to the listed 50\,Hz NEB.  The decoder NEB is what
the noise model in Eq.~\ref{eq:noise} should scale; using the
signal occupied BW would double-count the decoder integration.
\end{table}

% ── 3.12 Default settings ───────────────────────────────────────────
\subsection{Antenna Elevation Pattern}\label{sec:antpattern}

The antenna gain $G_{\text{ant}}$ is not a flat scalar; it varies with
the required takeoff (elevation) angle \emph{and with the band}.  The
elevation-pattern formulas below follow the standard treatment in
\emph{The ARRL Antenna Book}~\cite{arrl2023antenna}.  The elevation is
derived from hop geometry using a simplified flat-earth model:
\begin{equation}\label{eq:takeoff}
  \theta_{\text{elev}} = \arctan\!\left(\frac{2\,h_F}{d_{\text{hop}}}\right),
  \qquad
  d_{\text{hop}} = \min\!\bigl(d,\, d_{\text{hop}}^{\max}(h_F)\bigr)
\end{equation}
where $h_F = 300$\,km (typical F2 height; the smoothing form below
applies regardless of the exact value) and the per-hop range
$d_{\text{hop}}$ saturates at the geometric ceiling
$d_{\text{hop}}^{\max}(h_F)$ from Eq.~\ref{eq:hopceiling}.  Long-haul
DX paths require low takeoff angles ($5$--$10\degree$); regional
paths use higher angles ($15$--$30\degree$); NVIS uses near-vertical
($60$--$90\degree$).

\paragraph{Continuous across the hop boundary.}  The
$\min(d, d_{\text{hop}}^{\max})$ form keeps per-hop range continuous
at integer-hop transitions: real long-haul paths pack at the
geometric ceiling, so saturating per-hop range at
$d_{\text{hop}}^{\max}$ for $d \ge d_{\text{hop}}^{\max}$ is what
the geometry implies.  Below one ceiling the angle steepens with
shorter $d$.  Using integer $N_{\text{hops}}$ in the denominator
instead would produce a sawtooth in takeoff angle (e.g.\ a
$\sim 8\degree$ jump from $8.5\degree$ at $d = \SI{4000}{\km}$
to $16.7\degree$ at $d = \SI{4001}{\km}$ as $N_{\text{hops}}$
ticks 1$\to$2) that would propagate into the Fresnel
ground-reflection loss (Section~\ref{sec:lhop}) and the
elevation-pattern antenna gain (Eq.~\ref{eq:antgain}) on every
path crossing a ceiling.

\paragraph{Implication: per-hop geometry is invariant in $d$
above the ceiling.}  Once $d \ge d_{\text{hop}}^{\max}$, the
$\min(\cdot)$ saturates at the ceiling, so $\theta_{\text{elev}}$
(and therefore the Fresnel grazing angle and the antenna
elevation gain) is the \emph{same value} for every long-haul
path regardless of distance.  A 12000\,km Lisbon--Tokyo and a
17500\,km Lisbon--Sydney see identical per-hop $\theta_{\text{elev}}
= 8.5\degree$, and the antenna gain term $G_{\text{ant}}(\theta)$
contributes the same to both budgets.  The two paths still
differ in free-space loss (which is $\propto d^2$),
multi-hop ground reflection ($\propto N_{\text{extra}}(d)$), and
ionospheric absorption sums (longer paths have more reflection
points).  This is not a model approximation; it is the geometry
implied by ``hops pack at the ceiling above $d_{\text{hop}}^{\max}$''.

\paragraph{Type + height + $\lambda$.}  Operators supply three inputs:
antenna \emph{type} (eight classes, Table~\ref{tab:anttypes}), mounting
\emph{height} $h$ in metres, and the \emph{peak gain} $G_{\text{peak}}$
over typical ground (the type already encodes the free-space
directivity, so a 4-element Yagi has a bigger $G_{\text{peak}}$ than
a dipole at the same $h$).  The effective gain at elevation $\theta$
for band frequency $f$ is
\begin{equation}\label{eq:antgain}
  G_{\text{ant}}(\theta, f, h, \text{type}) =
    G_{\text{peak}} + G_{\text{rel}}(\theta, f, h, \text{type})
\end{equation}
where the relative elevation shape $G_{\text{rel}}$ is 0 dB at the
pattern peak and floors at $-\SI{15}{\dB}$ (real antennas don't vanish
into ideal nulls).

\paragraph{Horizontal types.}  For a horizontal wire, loop, or beam
at height $h$, the elevation pattern is dominated by the ground-
reflection factor for horizontal polarisation (image theory):
\begin{equation}\label{eq:gfh}
  \text{gf}_{\text{H}}(\theta, f, h) =
    2\,\bigl|\sin\!\bigl(k h \sin\theta\bigr)\bigr|,
  \qquad k = \frac{2\pi}{\lambda},\ \lambda = \SI{300}{m}/f_{\text{MHz}}.
\end{equation}
The maximum of 2 (6 dB above the free-space radiator) occurs at
$\sin\theta = \lambda/(4h)$.  For a dipole at $h = \SI{10}{m}$:
on 40 m ($\lambda = \SI{40}{m}$) the peak sits at
$\sin\theta = 1$, i.e.\ straight up -- a pure NVIS antenna;
on 20 m ($\lambda = \SI{20}{m}$) the peak sits at
$\sin\theta = 0.5 \Rightarrow \theta = 30\degree$ -- regional / short
DX; on 10 m ($\lambda = \SI{10}{m}$) the peak sits at
$\sin\theta = 0.25 \Rightarrow \theta = 14.5\degree$ -- classical
low-angle DX lobe.  Relative gain:
\begin{equation}
  G_{\text{rel,H}} = 20\log_{10}\!\bigl(\max(\text{gf}_{\text{H}}/2,\ 10^{-15/20})\bigr)
\end{equation}
The $-15$\,dB floor inside the $\max(\cdot)$ is a numerical guard
as well as a pattern-quality cap: at very low antenna heights
(e.g.\ a horizontal dipole at $h = \SI{1}{m}$ on 80\,m, where
$kh \approx 0.08$ and the formula's pre-floor value drops to
${\sim}-22$\,dB at the pattern peak), the bare expression goes
arbitrarily negative; the floor caps that to $-15$\,dB so the
budget stays well-conditioned and the verdict reflects the
``ground-mounted dipole'' regime rather than the unrealistic
deep null the formula would otherwise predict.

Horizontal loops pick up an additional high-angle weighting
\begin{equation}\label{eq:loophiangle}
  G_{\text{rel,loop}} = G_{\text{rel,H}} +
    20\log_{10}\!\bigl(0.5 + 0.5\sin\theta\bigr),
\end{equation}
the classical observation that a full-wave loop has a
45--60\degree{} elevation peak rather than the dipole's
ground-reflection-driven one.

Beams (hex, Yagi, stacked) use the \emph{same} ground-reflection
shape as a horizontal wire at the same height: the elevation lobe
of a Yagi is set by the ground reflection, not the number of
elements.  The directivity bonus (3--15 dB over a dipole) lives in
$G_{\text{peak}}$.

\paragraph{Vertical.}  The free-space quarter-wave pattern
$f_{\text{v}}(\theta) = \cos(\pi/2 \cdot \sin\theta)/\cos\theta$
combines with a vertical-polarisation ground reflection
$\text{gf}_{\text{V}}(\theta, f, h) =
2\,\bigl|\cos(k h \sin\theta)\bigr|$
which evaluates to $2$ at $h = 0$ ($\cos 0 = 1$, image in phase
everywhere) and reduces continuously as $h$ grows.  The
implementation retains an explicit $h \lesssim 0.1$\,m branch
\emph{only} as a numerical shortcut (skipping the
$\cos(k h \sin\theta)$ evaluation when it is guaranteed
${\sim}\!1$); mathematically the elevated-form expression at
$h = 0$ already yields the same answer.
\begin{equation}\label{eq:grelv}
  G_{\text{rel,V}} = 20\log_{10}\!\Bigl(\max\!\Bigl(
    \tfrac{f_{\text{v}}(\theta)\,\text{gf}_{\text{V}}}{1 \cdot 2},\ 10^{-15/20}
  \Bigr)\Bigr)
\end{equation}
The denominator normalises by the value of
$f_{\text{v}}\cdot\text{gf}_{\text{V}}$ at the antenna's actual
peak: $f_{\text{v}}(\theta) =
\cos\!\bigl(\frac{\pi}{2}\sin\theta\bigr)/\cos\theta$ has its
global maximum $f_{\text{v}}(0) = 1$ at the horizon and falls
monotonically toward $0$ at $\theta = \pi/2$ (overhead) by
L'H\^opital, with $f_{\text{v}}(80\degree)\approx 0.14$,
$f_{\text{v}}(85\degree)\approx 0.07$, $f_{\text{v}}(89\degree)
\approx 0.014$; combined with $\text{gf}_{\text{V}} = 2$ for a
ground-mounted radiator, the $1\cdot 2$ denominator anchors
$G_{\text{rel}} = 0$ at the antenna's strongest direction,
matching the horizontal-branch normalisation.

Peak at $\theta = 0\degree$ (horizon), null overhead.  No band-dependent
lobe migration for a ground-mounted vertical --- unlike the horizontal
family, its pattern is shape-invariant across HF.

\paragraph{Compromise.}  Indoor / attic / mag-loop installations have
a near-omnidirectional scattered pattern with a gentle low-angle
penalty from nearby conductors:
\begin{equation}
  G_{\text{rel,compromise}} =
    -\max\!\bigl(0,\ 2(1 - \theta/10\degree)\bigr).
\end{equation}
This is the only relative-pattern in the antenna set with a
\emph{negative} peak (the pattern goes to $-2$\,dB at the horizon
and rises to $0$\,dB above $10\degree$): the compromise antenna
loses gain at low takeoff angles relative to its already-modest
peak.  All elevation arguments $\theta$ in this section, including
the horizontal and beam expressions below, are angles in degrees
(the per-equation literals such as $10\degree$ set the convention).

\begin{table}[H]
\centering
\small
\caption{Antenna types and default (height, peak gain) pre-fills.}
\label{tab:anttypes}
\begin{tabular}{@{}lccl@{}}
\toprule
\textbf{Type} & \textbf{$h$ (m)} & \textbf{$G_{\text{peak}}$ (\si{\dBi})} & \textbf{Covers} \\
\midrule
Horizontal      & 10 & $+5$  & Dipole, inverted-V, G5RV, OCF, EFHW, folded dipole \\
Vertical        &  0 & $+2$  & \SI{1/4}{}, \SI{1/2}{}, \SI{5/8}{} wave over radials \\
Horizontal loop &  8 & $+4$  & Full-wave delta, quad, square \\
Small beam      & 12 & $+7$  & Hex beam, 2-/3-el Yagi, compact tribander \\
Medium beam     & 15 & $+9$  & 4-el Yagi, 4-el quad \\
Large beam      & 20 & $+12$ & 5-el+ Yagi, stacked / phased \\
Compromise      &  5 & $-2$  & Indoor, attic, mag loop, mobile whip \\
Custom          & 10 & user  & Any unusual setup \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Default Operator Settings}

The default station profile assumes a typical amateur installation:
$P_{\text{tx}} = \SI{50}{\dBm}$ (\SI{100}{\W}),
$G_{\text{ant}} = \SI{+5}{\dBi}$ peak (installed dipole with ground
reflection; effective gain varies by path elevation), mode = SSB,
noise environment = suburban.

% ── 3.13 Waterfall figure ───────────────────────────────────────────
\subsection{Worked Example: 20\,m, \SI{1000}{\km}, Quiet Day}

Table~\ref{tab:waterfall} lists the complete budget for a
\SI{14.1}{\MHz} path of \SI{1000}{\km} on a quiet day.  At this
distance the path is single-hop (\SI{4000}{\km} hop ceiling), so
$L_{\text{hop}} = 0$: the multi-hop reflection budget activates
only on longer paths.  The example uses the canonical
$h_F = \SI{300}{\km}$ for the takeoff-angle computation; the
runtime substitutes the live $h_{m\!F2}$ from the nearest GIRO
digisonde when available, which on enhanced-flux days can shift
$\theta$ by a few degrees and the per-hop ceiling by
${\sim}\!250$\,km.

\paragraph{$G_{\text{ant}}$ derivation (per Section~\ref{sec:antpattern}).}
$G_{\text{ant}}$ is not a flat scalar; it is computed from the
hop geometry. For a \SI{1000}{\km} single-hop with $h_F = \SI{300}{\km}$,
the takeoff angle is
$\theta = \arctan(2 \cdot 300 / 1000) \approx 31^\circ$.
For a horizontal dipole at $h = 10$\,m on 20\,m
($\lambda = 20$\,m), the elevation pattern peaks at
$\sin\theta_{\text{peak}} = \lambda/(4h) = 0.5$, i.e.\
$\theta_{\text{peak}} = 30^\circ$. The required \(31^\circ\) lands
essentially on the peak lobe, so $G_{\text{rel}} \approx 0$\,dB and
$G_{\text{ant}} = G_{\text{peak}} + G_{\text{rel}} = +5$\,dBi.
On a longer hop (e.g.\ \SI{3000}{\km} with $\theta \approx 11^\circ$)
$G_{\text{rel}}$ would be substantially negative on the same dipole;
the +5\,dBi here is geometric coincidence, not a fixed value.  Note
that this worked example uses the canonical $h_F = \SI{300}{\km}$
fallback for a clean derivation; in production the runtime reads
the live $h_m\!F2$ from the nearest digisonde (typically \SI{280}--\SI{360}{km}),
which would shift $\theta$ by a degree or two and may change which
side of the dipole peak the takeoff lands on.

\begin{table}[H]
\centering
\caption{SNR budget for 20\,m (\SI{14.1}{\MHz}), \SI{1000}{\km},
         quiet day, suburban noise, SSB, with the reflection point at
         local noon ($\cos\chi \approx 1$, which produces the
         $L_{\text{Dreg}} = -0.94$\,dB shown via
         $L_{\text{day}}(\SI{14.097}{\MHz}) \cdot 1^{0.7} = 0.94$\,dB
         per Eq.~\ref{eq:labsd}). The $\Sigma$ column
         tracks the running signal level. Below the rule, $N$ and
         $S_{\text{req}}$ are operands to the margin formula
         $M = S - N - S_{\text{req}}$, not additions to $\Sigma$.
         The resulting dB margin is $\sigma$-normalised by the prose
         below to derive the tier verdict per ITU-R P.842.  All
         dB values in the $\Delta$ and $\Sigma$ columns are
         rounded to 0.1\,\si{\dB}; cumulative rounding across
         the seven rows is bounded by $\pm\SI{0.5}{\dB}$ on the
         final $\Sigma$, well below the per-band $\sigma_g$
         resolution that drives the tier boundary.}
\label{tab:waterfall}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Term} & \textbf{$\Delta$} & \textbf{$\Sigma$ (dBm)} \\
\midrule
$P_{\text{tx}}$        & $+50$\,dBm  & $+50$    \\
$G_{\text{ant}}$       & $+5$\,dBi   & $+55$    \\
$L_{\text{fs}}$        & $-115$\,dB  & $-60$    \\
$L_{\text{Dreg}}$      & $-0.9$\,dB  & $-60.9$  \\
$L_{\text{MUF}}$       & $0$\,dB     & $-60.9$  \\
$L_{\text{iono}}$      & $-1$\,dB    & $-61.9$  \\
$L_{\text{hop}}$       & $0$\,dB     & $-61.9$  \\
\midrule
\multicolumn{2}{l}{\textit{Signal at Rx}} & $\mathbf{-61.9}$ \\
\midrule
\multicolumn{3}{@{}l}{\textit{Operands to margin formula:}} \\
$N$ (suburban)         & \multicolumn{2}{r}{$-105$\,dBm} \\
$S_{\text{req}}$ (SSB) & \multicolumn{2}{r}{$+10$\,dB}   \\
\midrule
\multicolumn{3}{@{}l}{$M = -61.9 - (-105) - 10 = \mathbf{+33.1\,\text{dB}}
  \;\Rightarrow\; M/\sigma = 3.68$} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Tier verdict.}
With margin $M = +33.1$\,\si{\dB} and the fixed-dB cuts of
Table~\ref{tab:tiers}, the verdict clears the Excellent threshold
($M \ge +18$\,\si{\dB}) by 15\,\si{\dB}.  The corresponding
circuit reliability under the per-band $\sigma$ for 20\,m
(\SI{9}{\dB}, Table~\ref{tab:bandsigma}) is
$\Phi(33.1 / 9) = \Phi(3.68) \approx 0.99988$ (effectively
$\sim$99.99\,\%; the tail beyond $z = 3.68$ is
$\sim 1.2 \times 10^{-4}$).  The wide head-room is typical of
midday 20\,m on quiet days at 1000\,km.

\paragraph{Worked example with bonuses (15\,m TEP, EU--Brazil at 19\,LT).}
The budget in Table~\ref{tab:waterfall} silently sets the three
additive bonuses (Eq.~\ref{eq:blend}: $B_{\text{grayline}}$,
$B_{\text{TEP}}$, $B_{\text{scatter}}$) to zero because none of
them fire on a quiet 20\,m midday path.  To anchor the bonus
arithmetic, consider the 15\,m TEP-eligible case at the current
moderate-cycle epoch ($F_{10.7\text{A}} = 120$): a Lisbon
($\phi_{\text{dip}} \approx +42\degree$) to S\~ao Paulo
($\phi_{\text{dip}} \approx -15\degree$, near the southern EIA
crest) path, $f = \SI{21}{\MHz}$,
midpoint local time $\approx 19$, two F2 hops, $f / \text{MUF} =
1.10$, $\sigma_{f_{\text{oF2}}} \approx \SI{0.4}{\MHz}$ across the
hops.  At those numbers the bonus terms read:
\begin{itemize}[nosep]
  \item $B_{\text{grayline}} \approx \SI{0.2}{\dB}$ (post-S0-\#2
        rewrite, the bonus is now formula-driven across
        $1.8\text{--}30$\,\si{\MHz} but small on 15\,m: at
        $f = \SI{21}{\MHz}$, $L_{\text{day}} = 200/(21.5)^2 =
        \SI{0.43}{\dB}$, and at midpoint LT$\,{\approx}\,19$ the
        sun is just past sunset so $\cos\chi \lesssim 0$, the
        formula factor $1 - \cos^{0.7}$ saturates near $1$, the
        sunset multiplier $\alpha = 0.5$ applies, giving
        $0.5 \cdot 0.43 \cdot 1 \approx \SI{0.22}{\dB}$ which the
        \SI{0.2}{\dB} floor lets through marginally).
  \item $B_{\text{TEP}} = B_{\text{TEP,plateau}}(F_{10.7\text{A}}{=}120)
        \cdot f_{\text{freq}} \cdot f_{\text{LT}} \cdot
        f_{\text{dip,src}} \cdot f_{\text{dip,dst}}$
        $\approx \SI{11.2}{\dB} \cdot \mathrm{ramp}(21,14,22)
        \cdot 1 \cdot 1 \cdot 1$
        $= \SI{11.2}{\dB} \cdot 0.875 \cdot 1 \cdot 1 \cdot 1$
        $\approx \SI{9.8}{\dB}$ (moderate-cycle plateau per
        Eq.~\ref{eq:tepplateau}; frequency factor not yet at the
        $[22, 50]$\,MHz plateau; both endpoints past the
        dip-latitude saturation knee at $|\phi_{\text{dip}}| \ge
        10\degree$; LT in the $[17, 23]$ plateau). At solar
        maximum ($F_{10.7\text{A}} \to \infty$) the plateau
        asymptotes to \SI{15}{\dB}, giving
        $B_{\text{TEP}} \to \SI{13.1}{\dB}$ on the same path.
  \item $B_{\text{scatter}} = w_{\text{sc}} \cdot
        \min(1, 0.4/1.5) \cdot \min(2, 0.10) \cdot 5$
        $= 1.5 \cdot 0.267 \cdot 0.10 \cdot 5$
        $\approx \SI{0.20}{\dB}$.
\end{itemize}
The verdict-margin contribution is therefore
$B_{\text{grayline}} + \max(B_{\text{TEP}}, B_{\text{scatter}}) =
0.2 + 9.8 = \SI{10.0}{\dB}$ on top of $M_{\text{physics}}$.  The
$\max(\cdot)$ form intentionally suppresses the small
$B_{\text{scatter}}$ contribution (0.2\,dB) on top of the
9.8\,dB TEP signal, since both describe the same F-region
irregularity-driven recovery channel.  An 80\,m sunrise example
exercises $B_{\text{grayline}}$ alone; that geometry is left as
an exercise.

% ═════════════════════════════════════════════════════════════════════
\section{MUF Estimation}\label{sec:muf}

The Maximum Usable Frequency is the single most important input to the
link budget.  ionocast derives it from two primary sources in a
consensus blend (Section~\ref{sec:muf-consensus}): observed kc2g
MUF(3000F2) at the path midpoint, and a simplified P.1239
climatology that fills in when the observation is stale or out of
range.  An optional third source --- per-hop foF2 fusion from the
GIRO digisonde network --- shares its per-hop foF2 sampling
machinery with the F2-scatter bonus of Section~\ref{sec:scatter}
(both consume inverse-square-distance-weighted foF2 from
in-range stations along the great circle), but the fusion role
itself is gated behind the \texttt{fusion-flag}
parameter of Appendix~\ref{app:repro} and held off in the
production blend pending a joint budget retune (the calibration
sweep showed it net-negative against the rebuilt climatology
baseline; \S\ref{sec:scatter} describes the scatter use of the
same sampling, which \emph{is} active, and the
``Per-hop fusion (deferred)'' paragraph there records the
calibration result that closed this branch).  The two-source consensus
is what ships.

\subsection{kc2g Observed MUF}\label{sec:kc2gobs}

The kc2g network publishes MUF(3000F2) from each ionosonde.  For
each of 72 radial paths sampled relative to the operator's QTH
(12 bearings every 30\textdegree{} $\times$ 6 distance rings:
2500, 4000, 6000, 9000, 12000, 16000\,km, recomputed per-QTH on
each reload), ionocast selects the nearest ionosonde to that path's
midpoint.  A \emph{freshness gate} discards observations older than
\SI{30}{\minute}.  The distance from each path midpoint to its
nearest ionosonde is displayed in the path table so operators can
gauge data confidence; stations beyond 1500\,km are flagged.

\paragraph{Why radial sampling rather than fixed cities.}  The
earlier basket was 5 fixed destinations (New York, S\~ao Paulo,
Johannesburg, Tokyo, Sydney) plus their long-path counterparts,
ten paths total.  That worked for QTHs in Europe (where every
named destination is a reasonable F2-hop distance away) but
produced two structural failure modes elsewhere:
\begin{enumerate}[nosep]
  \item \textbf{QTH = destination.}  An operator at PM85 (Tokyo)
        saw a ``QTH$\to$Tokyo'' path of \SI{0}{\km}, which the
        SNR budget treated as a trivial-distance F2 hop with no
        multi-hop loss, no auroral exposure, no D-region
        absorption.  Margins ran to \SI{+50}{\dB} and the verdict
        pinned at Excellent regardless of conditions.
  \item \textbf{QTH near destination.}  EM87 (Tennessee) at
        ${\sim}\!944$\,km from FN30 (NYC) sat in the gap between
        NVIS and proper F2-hop distances; the F2-hop budget
        undercount made every band look ``good to NYC'' even
        when no other destination was viable.
\end{enumerate}
Both are symptoms of the same problem: a fixed basket of 5 points
cannot sample the (distance, bearing) plane evenly for every
QTH.  Radial sampling fixes this by anchoring the basket to the
operator instead of a city list.  Every QTH gets identical
72-point coverage; no zero-length self-paths, no geographic
skew, no need to maintain a destination table.

\paragraph{Why the shortest ring is 2500\,km.}  A single F2 hop
has a geometric minimum distance of $\sim 2500$\,km set by the
F2-peak height $h_{\text{F2}} \in [250, 350]$\,km and the
antenna's lowest usable launch elevation ($\sim 5\text{--}15$\textdegree).
Below that there is a propagation dead zone where neither NVIS
(which fades above $\sim 1500$\,km as the reflection becomes too
oblique for the secant approximation) nor a low-elevation F2
hop is fully valid.  The budget's F2-hop math mis-models that
dead zone as a trivial-distance hop with very low loss.  Starting
the basket at 2500\,km keeps every sample in F2-hop territory
where the budget produces honest margins; sub-2500\,km
NVIS-mode contacts are a separate planned UI surface that
shares the budget's NVIS branch (Section~\ref{sec:nvisdetect})
but does not enter the radial-basket best-margin computation.

\paragraph{Long-path coverage.}  The previous basket included
explicit long-path counterparts for each of the five short-path
destinations (antipodal-midpoint lookups).  In the radial
basket, every bearing has its $\pm 180\textdegree$ partner
sampled at the same six distance rings, so long-path geometries
are captured implicitly.  The deepest ring (16\,Mm) is short
of the antipodal distance ($\sim 20$\,Mm) by design; a path at
16\,Mm sweeps the same midpoint band as a 4\,Mm path going the
other way around the Earth, so the redundancy is intentional.

\subsection{foF2 Climatology (Simplified P.1239)}\label{sec:climo}

When observed data are stale or unavailable, ionocast falls back to a
simplified climatological model anchored to ITU-R
P.1239~\cite{itup1239}.  The unmodified daytime baseline is the sum
of a polar-night-and-thermal-floor term and a latitude-modulated
solar-zenith-driven daytime term:
\begin{equation}\label{eq:fof2-base}
  b(\phi,\,\cos\chi)
    = b_{\text{floor}}(F_{10.7\text{A}})
      + 4.0\,\mathcal{P}(\phi)\,d(\cos\chi)
\end{equation}
where the solar-baseline floor is
\begin{equation}\label{eq:fof2-base-floor}
  b_{\text{floor}}(F_{10.7\text{A}})
    = 3.5 + 0.04\,(F_{10.7\text{A}} - 70).
\end{equation}
$b_{\text{floor}}$ is the value $b$ falls to overnight or at the
poles where $\mathcal{P}(\phi)\,d \to 0$: $3.5\,\si{\MHz}$ at the
quiet-sun reference $F_{10.7\text{A}} = 70$ plus a linear
F\textsubscript{10.7A} slope.  The slope is signed (no
positive-part clamp), so at deep solar minimum
($F_{10.7\text{A}} < 70$) the floor sinks below
$3.5\,\si{\MHz}$, consistent with the lower foF2 floors observed
at extended-minimum nights; this differs deliberately from the
EIA amplitudes $A_{\text{cr}}$ and $A_{\text{tr}}$ in
Eq.~\ref{eq:eia-amp} and Eq.~\ref{eq:eia-tr-amp}, where the
slope is wrapped in $\max(0, \cdot)$ so the equatorial-anomaly
amplitude does not invert below $F_{10.7\text{A}} = 70$.  The
second term of Eq.~\ref{eq:fof2-base} carries the daytime
$\mathcal{P}(\phi) \cdot d(\cos\chi)$ shape that dominates the
verdict at midday on illuminated paths, with
$\mathcal{P}(\phi)$ the two-stage poleward fall-off:
\begin{equation}\label{eq:fof2-poleward}
  \mathcal{P}(\phi) = \bigl(1 - 0.003|\phi|\bigr) \cdot
                      \frac{1}{1 + \exp\!\bigl((|\phi| - 73^\circ)/8^\circ\bigr)}
\end{equation}
The first factor is a gentle linear midlatitude term; the second is
a sigmoid centred at $|\phi| = 73^\circ$ with $8^\circ$ steepness
that captures the polar F2 thinning above $\sim\!60^\circ$.
The linear factor alone would under-predict that thinning
(F2 thins faster than linearly with latitude above $\sim
60^\circ$ as the polar-cap tongue takes over with patchy
electron density), producing a 1--1.7\,MHz daytime over-prediction
at Tromsø / Gakona / Eielson AFB without the sigmoid term.
The sigmoid factor reproduces the observed magnitude:
$\mathcal{P}(45^\circ) \approx 0.84$, $\mathcal{P}(60^\circ) \approx 0.69$,
$\mathcal{P}(70^\circ) \approx 0.47$, $\mathcal{P}(80^\circ) \approx 0.22$.
After sunset, foF2 does not collapse instantly --- recombination
of the F-region runs on a multi-hour timescale.  The post-sunset
diurnal curve breaks into two distinct phases that ionocast
captures with two structurally different terms:

\begin{itemize}[nosep]
  \item \textbf{Phase 1, sunset to $\sim 3$\,h after sunset:}
        residual photoionization keeps daytime production going.
        The plasma's effective ``recently illuminated cosine''
        decays on a $\sim 3$\,h timescale rather than dropping to
        zero at the terminator. Captured by the photoelectron
        \emph{driver} $d(\cos\chi)$, which lives \emph{inside}
        $b$'s argument as a numerator multiplier:
        \begin{equation}\label{eq:fof2-driver}
          d(\cos\chi) = \max\!\bigl(\max(0,\cos\chi),\;
                                    0.7\,\max(0,\cos\chi_{-3\text{h}})\bigr)
        \end{equation}
        $d$ holds at the daytime cosine through sunset, then decays
        toward zero as the lagged $\cos\chi_{-3\text{h}}$ also
        crosses below zero.  On a cold start (browser session with
        no historical solar-zenith cache), the implementation
        substitutes $\cos\chi_{-3\text{h}} = \cos\chi_{\text{now}}$
        for the first 3 hours of session lifetime, which means the
        memory-lag term degenerates to the live cosine until the
        cache fills.  This biases foF2 slightly low at evening
        startup (the lag would have read the prior daytime cosine,
        which is positive while the live cosine is at-or-below
        zero) by at most $\sim$10\% on the worst case (sunset start,
        single midlat path), which on a near-MUF cell maps to a
        ${\sim}$10\,\% MUF bias --- enough to flip a Fair / Good
        verdict on the upper bands at exactly the time of day operators
        most often check propagation.  The cache is re-populated within
        the first three 10-minute refresh cycles, so the bias is
        bounded and self-clearing, but the worst-case window aligns
        with after-work session starts and is therefore worth
        eliminating rather than tolerating.

        \emph{Pre-population fix.}  $\cos\chi_{-3\text{h}}$ is purely
        astronomical --- it is fully determined by the current UTC
        and the path-midpoint geographic coordinates --- so the
        ``no historical cache'' branch can compute it from
        $\cos\chi(t_{\text{now}} - \SI{3}{h})$ directly on the first
        load, rather than waiting three refresh cycles for the cache
        to fill from live samples.  Doing so retires the cold-start
        bias entirely without any historical state, and is candidate
        cleanup tracked in Section~\ref{sec:limits} alongside the
        F\textsubscript{10.7A}-fallback surfacing recommendation
        (the two share a UI-honesty motivation: never let the live
        verdict silently inherit a degraded input when the
        non-degraded value is locally computable).  Until that
        change ships, the runtime should at minimum surface a
        ``cold-start cache filling'' chip during the first
        ${\sim}30$~minutes of a fresh session so an operator
        startup-checking the upper bands at sunset is not given a
        confidently-stated verdict on a degraded input.
  \item \textbf{Phase 2, $\sim 3$\,h after sunset onward:}
        residual production has decayed to zero and ongoing
        recombination pulls electron density toward a nighttime
        floor (typically reached around $03$\,LT). Captured by the
        night-decay \emph{multiplier} that operates on the
        already-constructed $b$ as a recombination floor:
        \begin{equation}\label{eq:fof2-nightdecay}
          b(\phi,\,\cos\chi) \to b(\phi,\,\cos\chi)\cdot
            \max(0.6,\, 1 + 0.4\cos\chi)
          \quad \text{when } \cos\chi < 0
        \end{equation}
        The multiplier reads $\sim 1$ at the terminator (only
        slightly negative $\cos\chi$, hardly any reduction),
        ramping toward the $0.6$ floor by deep night.
\end{itemize}

The two terms are not equivalent multipliers: $d(\cos\chi)$ is the
\emph{driver argument} that the daytime baseline
$b_{\text{floor}}(F_{10.7\text{A}}) + 4.0\,\mathcal{P}(\phi)\,d$
constructs $b$ from, while the night-decay multiplier is a
\emph{post-multiplier} that scales the constructed $b$ during the
$\cos\chi < 0$ interval.  They overlap in time --- both are
``firing'' from sunset onward --- but their contributions peak in
different phases: in early evening the driver dominates (memory
lag still strong, multiplier $\approx 1$); in deep night the
multiplier dominates (driver decayed to zero, multiplier at or
near the $0.6$ floor).  Removing either term over-predicts
foF2 in its dominant phase, which the diurnal sweep in
Section~\ref{sec:climo} confirms.

The result is then floored at \SI{2}{\MHz} and scaled by two
dimensionless anomaly corrections, with the combined-anomaly product
itself floored at $0.7$ so that a strong dip-equator trough cannot
collapse the prediction below ${\sim}70\,\%$ of baseline:
\begin{equation}\label{eq:fof2}
  f_{\text{oF2}} = \max(2,\,b)\cdot
    \max\!\Bigl(0.7,\,
      \bigl(1 + \alpha_{\text{EIA}}(\phi_{\text{dip}}, \cos\chi)\bigr) \cdot
      \bigl(1 + \alpha_{\text{win}}(\phi, m, \cos\chi)\bigr)\Bigr)
\end{equation}
and then $\text{MUF}(3000) \approx 3.0 \times f_{\text{oF2}}$.  The
$0.7$ floor binds only at the dip equator under solar maximum, where
the trough term $\alpha_{\text{tr}}$ saturates at $-0.50$ and the
unfloored product would otherwise read $0.50$ (see
\S\ref{sec:climo}'s EIA discussion); real Jicamarca minima sit
around $-20$ to $-30\,\%$ from the crest-flank baseline, not zero.
Together the two floors imply $\text{MUF}(3000) \geq
\SI{4.2}{\MHz}$ on the climo-only branch
($2 \cdot 0.7 \cdot 3 = 4.2$), so the climatology alone cannot
return MUF below the lowest amateur HF band (160\,m,
\SI{1.8}{\MHz}); a closed 80\,m or 160\,m verdict on the
climo-only branch is therefore structurally impossible and only
arises on the ingested-MUF branch (kc2g) or via direct
absorption / D-layer terms.

The night-decay multiplier and the 3-h memory lag are part of the
post-rebuild climatology.  Initial validation used
30~d of GIRO digisonde foF2 across the original nine-station
midlatitude basket (the basket has since grown to 27 stations
spanning the equatorial and polar belts; see
Appendix~\ref{app:giro}, used by the EIA grid sweep below
rather than by the night-decay validation).  The night-decay
fit dropped baseline RMSE from \SI{1.91}{\MHz} to
\SI{1.70}{\MHz} ($-11\,\%$) and centred the residual bias from
\SI{-0.26}{\MHz} to \SI{-0.05}{\MHz}; the prior plain-P.1239
fit predicted ${\sim}\SI{5.5}{\MHz}$ at midlat midnight versus
the ${\sim}\SI{3.7}{\MHz}$ that nine months of GIRO
nighttime samples actually report, and collapsed foF2 to the
sunset-instant baseline rather than tracking the observed
multi-hour evening tail.

\paragraph{Equatorial Ionisation Anomaly (EIA).}  The daytime
fountain
effect~\cite{anderson1981} lifts plasma from the magnetic equator up
the field lines and deposits it in two daytime crests at magnetic dip
latitude $\pm \phi_{\text{EIA}}$ ($\phi_{\text{EIA}} =
\SI{15}{\degree}$), leaving a trough at $\phi_{\text{dip}} \approx 0$.
ionocast captures both via a positive crest kernel and a narrow
negative trough kernel summed into one $\alpha_{\text{EIA}}$:
\begin{equation}\label{eq:eia}
  \alpha_{\text{EIA}} =
    \Bigl(A_{\text{cr}}(F_{10.7\text{A}})\,
      e^{-d_{\text{cr}}^2 / 2\sigma_{\text{cr}}^2}
    -\,A_{\text{tr}}(F_{10.7\text{A}})\,
      e^{-\phi_{\text{dip}}^2 / 2\sigma_{\text{tr}}^2}\Bigr)
    \cdot\max(0,\cos\chi)
\end{equation}
where $d_{\text{cr}} = \min(|\phi_{\text{dip}} - \phi_{\text{EIA}}|,
|\phi_{\text{dip}} + \phi_{\text{EIA}}|)$ is the distance to the
nearest crest, $\sigma_{\text{cr}} = \SI{18}{\degree}$, and
$\sigma_{\text{tr}} = \SI{6}{\degree}$ (sharper than the crests; the
trough is narrow in latitude).  $\phi_{\text{dip}}$ here is the
\emph{signed} dip latitude (positive in the northern magnetic
hemisphere, negative in the southern), derived from the same
tilted-dipole approximation as $\phi_{\text{CGM}}$
(Eq.~\ref{eq:cgm}) but without the absolute value: the crest term's
$d_{\text{cr}}$ uses the signed value to pick the nearer of the
two crests, and the trough term's $\phi_{\text{dip}}^2$ is
sign-symmetric.  An implementation that erroneously used
$|\phi_{\text{dip}}|$ would compute the same trough magnitude but
would mis-place the crests on the wrong hemisphere for southern
paths. Both amplitudes scale with EUV
illumination through F10.7A:
\begin{equation}\label{eq:eia-amp}
  A_{\text{cr}}(F_{10.7\text{A}}) =
    \min\!\bigl(0.85,\,
      0.50 + 0.007\cdot\max(0,\,F_{10.7\text{A}} - 70)\bigr)
\end{equation}
\begin{equation}\label{eq:eia-tr-amp}
  A_{\text{tr}}(F_{10.7\text{A}}) =
    \min\!\bigl(0.50,\,
      0.30 + 0.002\cdot\max(0,\,F_{10.7\text{A}} - 70)\bigr)
\end{equation}
At quiet sun ($F_{10.7\text{A}} = 70$) the crest amplitude is
$0.50$ and the trough is $0.30$; the crest cap-saturates at
$0.85$ at $F_{10.7\text{A}} \ge 120$, while the trough
cap-saturates at $0.50$ at $F_{10.7\text{A}} \ge 170$ (different
saturation epochs because the slopes $0.007$ and $0.002$
differ; the trough rises more slowly with EUV).  At the
current moderate-cycle epoch ($F_{10.7\text{A}} \approx 120$)
the crest is at its $0.85$ cap and the trough is at $0.40$,
short of its $0.50$ cap.  The combined-anomaly $0.7$ floor
in Eq.~\ref{eq:fof2}
ensures a strong trough never collapses density below $\approx
70\%$ of baseline.  The trough kernel is a separate Gaussian
with a negative sign so that the EIA combined factor can
predict depression on the dip equator (a crest-only formulation
would predict $\sim\!+27\%$ enhancement there, an artefact of
the symmetric distance-to-nearest-crest functional form being
unable to express trough behaviour).

The values were settled by an 80-point grid sweep over
$(\text{base}, \text{slope}, \sigma)$ against 30~d of GIRO foF2 at
the four equatorial-belt stations: Townsville and Niue near the
southern crest of the Pacific arch; Ascension at dip latitude
${\sim}\!-3\degree$ (deep in the trough, far from the
$\sim\!-15\degree$ southern crest peak); Jicamarca essentially
\emph{on} the magnetic dip equator (dip latitude ${\sim}\!3\degree$
S), anchoring the equatorial trough.  The equatorial bias spread
floors at ${\sim}\SI{2.2}{\MHz}$ across the grid for two coupled
reasons: the cost function has no northern-crest station to
balance the southern-crest pair (so the Gaussian's amplitude is
fit one-sided), and \emph{half} of the basket --- Jicamarca and
Ascension --- sits in the equatorial trough rather than on a
crest peak, pulling the fitted amplitude down and the width out
relative to a true-crest fit.  The basket is now selected
\emph{geometrically} from \texttt{GIRO\_STATIONS} (any station with
$|\phi_{\text{dip}}| \le 25^\circ$ joins the EIA-relevant set
automatically), so adding any future northern-crest station ---
S\~ao Lu\'is (BR, geographic ${\sim}\!-2.5\degree$ but dip
latitude near $+5$ to $+10\degree$\,N because the South Atlantic
Anomaly shifts the magnetic equator south of its geographic
counterpart in this longitude sector, putting an equatorial
geographic station onto the northern magnetic flank) is the credible Americas candidate, with any Indian /
SE-Asian DIDB station near $\phi_{\text{dip}} \approx +15\degree$\,N
also qualifying --- closes both asymmetries without further code
edits and the floor is expected to drop below 2.2\,MHz.  (Tucum\'an
(AR) at $-26.8\degree$ and Cachoeira Paulista (BR) at $-22.7\degree$
are \emph{southern}-hemisphere dip-latitude stations; the latter
already participates in the GIRO basket and neither would help
close the \emph{northern}-crest gap; see Appendix~\ref{app:giro}
for the current basket.)  After the BVJ03 Boa Vista addition the
fit converged at base 0.50 / slope 0.007 / $\sigma = 18$ with
$\max|\text{bias}|$ dropping from 1.84 to 1.18\,MHz.

Dip latitude is derived from the same tilted-dipole approximation
used for CGM latitude (Eq.~\ref{eq:cgm}) but signed, so hemisphere
is preserved.  All numerical $\phi_{\text{dip}}$ values quoted in
this paper for specific stations and paths are computed from this
tilted-dipole formula, not from a real-IGRF dip-equator field
inversion.  The two can differ by 5--$\SI{10}{\degree}$ in the
South Atlantic Anomaly sector and by a few degrees elsewhere; we
use the tilted-dipole convention because it is the same number the
runtime model evaluates, so quoted values reproduce directly from
the code.

\paragraph{Winter anomaly.}  Midlatitude ($\SI{35}{\degree} \leq |\phi|
\leq \SI{60}{\degree}$) daytime foF2 runs noticeably higher in the
local winter than in local summer --- the classical O/N$_2$ seasonal
shift in the thermosphere.  ionocast applies a day-of-year-phased
cosine kernel peaked at the local winter solstice:
\begin{equation}\label{eq:winter}
  \alpha_{\text{win}} = A_{\text{win}}\cdot
    \tfrac{1}{2}\bigl(1 + \cos\theta_d\bigr)\cdot\max(0,\cos\chi),
  \quad
  \theta_d = \frac{2\pi\,(d_{\text{yoy}} - d_{\text{solstice}})}{365}
\end{equation}
with $A_{\text{win}} = 0.12$, $d_{\text{yoy}}$ the UTC day of year
($1\!=\!1$~January), and $d_{\text{solstice}} = 355$ (December~21)
for the northern hemisphere ($\phi \geq 0$) or $d_{\text{solstice}}
= 172$ (June~21) for the southern hemisphere ($\phi < 0$).  The
day-of-year phase places the cosine peak at the actual astronomical
solstice; an earlier calendar-month formulation
$\theta_m = 2\pi\,(m + 0.5 - m_{\text{winter}})/12$ landed the peak
five days early because mid-month does not coincide with the
solstice date.  The correction is daytime-only and switches off
outside the midlatitude band.  The amplitude $A_{\text{win}} = 0.12$
is set against the same 30-day GIRO foF2 grid sweep that fits the
EIA amplitudes (\S\ref{sec:climo}, basket: Juliusruh, Pruhonice,
Fairford, Sopron, Athens, Rome, Gibilmanna, Millstone Hill,
Boulder, Wallops Island), reading the winter-vs-summer daytime
foF2 ratio at midlatitude stations and matching the model's
$1 + A_{\text{win}}$ factor to the median observed ratio
($1.12 \pm 0.03$ across the basket).  This is consistent with
Pr\"olss~\cite{prolss1995}'s F-region seasonal-shift survey
(${\sim}10$--$15\%$ winter enhancement in the cycle-averaged
midlatitude foF2 record) but the specific value is observation-fit,
not literature-quoted.  The 10-station basket is entirely
northern-hemispheric (continental Europe and US east coast); the
same $A_{\text{win}} = 0.12$ amplitude is mirrored south of the
equator with the solstice phase flipped, but the southern-hemisphere
amplitude itself was not separately fit and could differ given
the asymmetric land/ocean distribution.  Pr\"olss reports SH
amplitudes within the same ${\sim}10$--$15\,\%$ band so the symmetric
choice is defensible at the climatology-class accuracy this term
needs to deliver, but a southern-hemisphere validation pass is
tracked as future work.

\subsection{Symmetric Consensus Blend}\label{sec:muf-consensus}

The consensus algorithm uses the geometric mean of the two estimates
regardless of divergence direction:
\begin{enumerate}[nosep]
  \item Compute divergence:
    $\delta = |\ln(\text{MUF}_{\text{kc2g}} / \text{MUF}_{\text{climo}})|$
  \item Result:
    $\text{MUF}_{\text{cons}} = \sqrt{\text{MUF}_{\text{kc2g}} \cdot \text{MUF}_{\text{climo}}}$
\end{enumerate}
$\delta$ is retained for diagnostic surfacing (the UI shows it on
hover) but does not branch the result.

\paragraph{Single-source fallbacks.}  When the kc2g freshness gate
(\SI{30}{\minute}) discards the observation, $\text{MUF}_{\text{cons}}
= \text{MUF}_{\text{climo}}$ and the source label reads ``climo''.
When the climatology returns null --- the rare combination of
extreme inputs that drives the unmodified daytime baseline below
the absolute floor --- $\text{MUF}_{\text{cons}} =
\text{MUF}_{\text{kc2g}}$ when the latter is fresh, and
``unavailable'' otherwise (the band is treated as below MUF).
$\delta$ is reported as $0$ in both single-source cases.

The symmetric geometric mean treats real upward enhancements (EIA
crest, post-storm positive phase, evening F-region uplift,
Es-lifted MUF) and declining-ionosphere readings the same way:
$\text{kc2g} = 2 \cdot \text{climo}$ produces $1.41 \cdot
\text{climo}$ (captures most of the enhancement while damping
spurious extremes); $\text{kc2g} = 0.5 \cdot \text{climo}$
produces $0.71 \cdot \text{climo}$ (mildly conservative on storm
depressions, with the absorption side of storm response covered
separately by the storm-lag $K_p$ kernel and $L_{\text{aur}}$
term).  At $\text{kc2g} \approx \text{climo}$ (the common case)
the blend is unchanged.

\subsection{Per-Hop Minimum MUF (P.533 \S3.1)}\label{sec:perhopmuf}

For paths requiring $N \geq 2$ hops, each reflection point at fraction
$(2k-1)/(2N)$ along the great circle sees its own local foF2
climatology (latitude fall-off, EIA crests/trough, winter anomaly,
diurnal phase). The path MUF is the \emph{minimum} across all hops,
evaluated by computing per-hop foF2 climatology directly at each
reflection point and applying the kc2g-vs-climo correction scalar
$\text{MUF}_{\text{mid}} / \text{MUF}_{\text{climo,mid}}$ to preserve
the upstream kc2g signal:
\begin{equation}\label{eq:pathmuf}
  \text{MUF}_{\text{path}} = \min_{k=1}^{N}\;
    3.0 \cdot f_{\text{oF2}}(\phi_k, \lambda_k, \cos\chi_k, t)
    \cdot \frac{\text{MUF}_{\text{mid}}}{\text{MUF}_{\text{climo,mid}}}
\end{equation}
where $f_{\text{oF2}}(\cdot)$ is the climatology of \S\ref{sec:climo}
evaluated at the reflection point's coordinates, time, and solar
zenith.

\paragraph{Legacy fallback path: when it fires, what it does, and
how the UI should surface it.}  When either
$\text{MUF}_{\text{climo,mid}}$ or F\textsubscript{10.7A} is
unavailable, \texttt{pathMinMuf} (\texttt{src/physics/climatology.js})
falls back to an illumination-ratio scaling
$\text{MUF}_{\text{mid}} \cdot S(\cos\chi_k, F)/S(\cos\chi_{\text{mid}}, F)$
with $S(c, F) = \max(F, \sqrt{\max(0.05, c)})$ and night-floor
$F = 0.4$ (hard-coded; the dynamic formulation
\texttt{nightFloor()} in \S\ref{sec:nightfloor} is exported
separately for callers but is \emph{not} called by
\texttt{pathMinMuf}'s fallback branch in
\texttt{src/physics/climatology.js}).  This is the same form
Eq.~\ref{eq:pathmuf} supersedes;
its ${\sim}\!\sqrt{0.05} \approx 0.224$ effective night floor
systematically over-pessimises long cross-hemisphere paths
(e.g.\ a 12\,000\,km path with daylight midpoint and a night
terminus hop returns ${\sim}0.224\cdot\text{MUF}_{\text{mid}}$
where the local climatology would read 6--10\,MHz).  The fallback
fires in two operationally distinct conditions:
\begin{enumerate}[nosep]
  \item \textbf{Unit tests} that construct a budget without
        seeding F\textsubscript{10.7A} or a climatology midpoint MUF.
        This is the documented test path and the original
        motivation for keeping the fallback in the codebase.
  \item \textbf{NOAA SWPC F\textsubscript{10.7} outage in
        production}, which is rare but real --- the daily
        F\textsubscript{10.7} product has missed publication
        windows of several hours under solar-region storms and
        SWPC infrastructure outages over the past two cycles.
        During such an outage the per-hop MUF computation
        silently switches from Eq.~\ref{eq:pathmuf} to the
        illumination-ratio form, with no operator-facing
        indication that the path-specific calculation has
        degraded.
\end{enumerate}
The UI does \emph{not} currently surface fallback-active state;
the band table and Reference Paths panel both render their
values as if Eq.~\ref{eq:pathmuf} had produced them.  For the SWPC-outage
case this is a defect: an operator looking at a closed-tier
verdict on a long polar path during an outage sees a tier produced
by the inferior formulation without any signal that the verdict's
foundation is degraded.  A fix in either of two directions closes
the gap.  \textbf{Surface the fallback}: a kc2g-panel chip
reading ``F\textsubscript{10.7} unavailable --- per-hop MUF using
illumination-ratio fallback'' would let the operator weight the
verdict accordingly.  \textbf{Or hard-error the fallback in
production}: gate the fallback to test code only and return
$\text{MUF}_{\text{path}} = \text{MUF}_{\text{mid}}$ unchanged on
F\textsubscript{10.7A} outage, so the budget at least uses the
upstream kc2g value without applying the inferior shape.  The
unsurfaced silent fallback is the worst of the three options;
shipping with one of the explicit alternatives is preferable.

\paragraph{UI vs.\ budget MUF.}  The MUF column on the Reference
Paths panel displays $\text{MUF}_{\text{mid}}$ from the kc2g
consensus blend at the path midpoint --- not
$\text{MUF}_{\text{path}}$.  The two coincide for short paths
(NVIS, 1-hop) where the midpoint \emph{is} the only reflection
point. On multi-hop circuits the budget uses the per-hop minimum
(Eq.~\ref{eq:pathmuf}); a small panel-vs-verdict gap remains on
paths whose limiting hop sits in a structurally low-foF2 region
(deep polar, dip equator). The panel MUF is the midpoint reading;
the verdict is the trustworthy summary built from the per-hop
minimum.

\subsection{Night Floor (CCIR/IRI)}\label{sec:nightfloor}

The night floor scales with solar activity and latitude:
\begin{equation}\label{eq:nightfloor}
  F = \text{clamp}\!\left(
    0.25 + 0.0025(F_{10.7\text{A}} - 70)
         - 0.10\,\min\!\left(1,\,\frac{|\phi|}{60}\right),\;
    0.20,\; 0.60
  \right)
\end{equation}
\textbf{Implementation status.}  The dynamic formulation in
Eq.~\ref{eq:nightfloor} is exported by
\texttt{nightFloor(f107A, latAbs)} in
\texttt{src/physics/climatology.js} but \emph{is not currently
called} by the legacy illumination-ratio fallback in
\texttt{pathMinMuf}; that fallback hard-codes $F = 0.4$ instead.
The production per-hop MUF (Eq.~\ref{eq:pathmuf}) carries its
own night behaviour through the night-decay multiplier of
Eq.~\ref{eq:fof2-nightdecay} and does not consume $F$ either.
Eq.~\ref{eq:nightfloor} is therefore documented here as the
\emph{intended} dynamic floor (a future cleanup would wire
\texttt{nightFloor()} into the fallback in place of the
$F = 0.4$ literal, retiring the contradiction); current
production reads the static $F = 0.4$ on the fallback branch
and never evaluates Eq.~\ref{eq:nightfloor}.  If the legacy
fallback is removed entirely (Section~\ref{sec:limits}), both
the $F = 0.4$ literal and Eq.~\ref{eq:nightfloor} can be
retired together.

\subsection{NVIS Detection for Short Paths}\label{sec:nvisdetect}

For paths shorter than \SI{500}{\km} the signal travels at
near-vertical incidence (NVIS).  The relevant frequency limit is
the \emph{critical frequency} $f_{\text{oF2}}$ (vertical reflection),
not $\text{MUF}(3000)$ (oblique reflection at $\sim$3000\,km range).
Since $\text{MUF}(3000) \approx 3 \times f_{\text{oF2}}$, using
MUF(3000) for short paths over-predicts by up to a factor of three.

When a short NVIS path is detected, ionocast substitutes GIRO-observed
$f_{\text{oF2}}$ with a secant correction for the mild obliqueness of
a real (non-zero distance) NVIS hop:
\begin{equation}\label{eq:nvis}
  \text{MUF}_{\text{NVIS}}(d) = f_{\text{oF2}} \cdot
    \sec\!\left(\arctan\!\frac{d}{2\,h_F}\right)
\end{equation}
For $d = \SI{500}{\km}$ at $h_F = \SI{300}{\km}$ the secant factor
is $\approx 1.30$, so the usable frequency is about 30\% higher
than bare $f_{\text{oF2}}$.

\paragraph{$h_F$ source.}  Both \texttt{nvisSecantFactor} (this
section) and \texttt{hopCeilingKm} (Eq.~\ref{eq:hopceiling}) read
the same observed $h_{m\!F2}$ from the nearest GIRO digisonde when
the panel reports it, falling back to $\SI{300}{\km}$ when the
station's most recent record has no $h_{m\!F2}$ value.  On a high-
solar-flux day with $h_{m\!F2} \approx \SI{340}{\km}$ the F2 hop
ceiling extends to ${\sim}\!\SI{4258}{\km}$
(Section~\ref{sec:lhop}) \emph{and} the NVIS secant rises by
${\sim}\!5\,\%$ at $d = \SI{500}{\km}$ --- so the band-gate
$f \le \text{MUF}_{\text{NVIS}}(d)$ tracks the same upper-band
flip the F2 hop count does, instead of leaving the two budgets
on different $h_F$ assumptions.

\paragraph{NVIS tail blending (\SI{500}{\km} to \SI{1500}{\km}).}
A hard cutoff at \SI{500}{\km} mishandles intermediate-distance paths
where NVIS still contributes alongside the F2 oblique mode.  ionocast
blends $\text{MUF}_{\text{NVIS}}(d)$ with the path's F2 MUF via a
linear ramp:
\begin{equation}\label{eq:nvistail}
  w_{\text{NVIS}}(d) = \begin{cases}
    1                                  & d \le \SI{500}{\km} \\[2pt]
    \dfrac{\SI{1500}{\km} - d}{\SI{1000}{\km}} & \SI{500}{\km} < d < \SI{1500}{\km} \\[8pt]
    0                                  & d \ge \SI{1500}{\km}
  \end{cases}
\end{equation}
\begin{equation*}
  \text{MUF}_{\text{eff}}(d) =
    w_{\text{NVIS}}(d) \cdot \text{MUF}_{\text{NVIS}}(d)
    + \bigl(1 - w_{\text{NVIS}}(d)\bigr) \cdot \text{MUF}_{\text{F2}}(d).
\end{equation*}
Below 500\,km the budget reads pure NVIS; beyond 1500\,km it reads
pure F2; in between the two modes share the prediction in proportion
to where the takeoff angle sits between near-vertical and oblique.
The mode label tracks whichever side carries $\ge 0.5$ weight, so
the UI's per-band breakdown still reads ``NVIS'' or ``F2'' cleanly.
This blending is gated on a digisonde $f_{\text{oF2}}$ being
available and the operating frequency fitting under the
geometry-aware NVIS MUF
($f \le \text{MUF}_{\text{NVIS}}(d)$ from Eq.~\ref{eq:nvis}).
The dynamic gate covers 30\,m / 17\,m NVIS at solar maximum and
naturally degrades below the foF2 reading at solar minimum.

\subsection{Short-Path vs.\ Long-Path Coverage}

Between any two points on Earth there are two great-circle paths,
and the long path (LP) can be open when the short path (SP) is
closed --- particularly at dawn and dusk on 20\,m and 17\,m.  In
the radial basket (Section~\ref{sec:kc2gobs}) the LP/SP distinction
is implicit rather than special-cased: every compass bearing has
its $\pm 180\textdegree$ partner sampled, so a path travelling
``the other way around'' from an operator at IO87 (Edinburgh) to
QF56 (Sydney) shows up as the long-distance ring on the bearing
toward south-east Asia rather than the ring toward Australia
direct.  Both compete on equal footing in the per-band best-margin
comparison; whichever has the higher budget wins, with no separate
SP/LP code path or antipodal-midpoint arithmetic.

% ── MUF diurnal curve ───────────────────────────────────────────────
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=12cm, height=7cm,
  xlabel={UTC Hour},
  ylabel={MUF (\si{\MHz})},
  xmin=0, xmax=24,
  ymin=0, ymax=28,
  xtick={0,3,6,9,12,15,18,21,24},
  ytick={0,5,10,15,20,25},
  extra y ticks={8},
  extra y tick labels={},
  grid=major,
  grid style={gray!30},
  legend pos=north east,
  legend style={font=\small},
  title={Diurnal MUF Variation (Midlatitude, $F_{10.7\text{A}}=150$)},
  title style={font=\small\bfseries},
]

% Night floor at 8 MHz
\addplot[name path=floor,gray,dashed,thick,domain=0:24,samples=2]
  {8};
\addlegendentry{Night floor (\SI{8}{\MHz})}

% MUF curve: sqrt(cos(zenith)) shape, peak at local noon (12 UTC for this example)
% cos(chi) = sin(lat)*sin(dec) + cos(lat)*cos(dec)*cos(HA)
% Floor at F * MUF_peak = 0.40 * 20 = 8 MHz (matches caption F ~ 0.40
% at midlat moderate solar activity per Eq:nightfloor).
\addplot[name path=muf,blue,very thick,domain=0:24,samples=200]
  {max(8, 20 * max(0.40, sqrt(max(0.05, cos(deg(abs(x-12)/24*2*pi))))))};
\addlegendentry{MUF(3000F2)}

% Shade night region (before sunrise ~6, after sunset ~18)
\addplot[name path=xaxis,draw=none,domain=0:6,samples=2] {0};
\addplot[name path=top1,draw=none,domain=0:6,samples=2] {28};
\addplot[gray!15] fill between[of=xaxis and top1];

\addplot[name path=xaxis2,draw=none,domain=18:24,samples=2] {0};
\addplot[name path=top2,draw=none,domain=18:24,samples=2] {28};
\addplot[gray!15] fill between[of=xaxis2 and top2];

% Sunrise/sunset markers
\draw[orange,thick,dashed] (axis cs:6,0) -- (axis cs:6,28);
\node[font=\scriptsize,orange,anchor=south,rotate=90] at (axis cs:6,14) {Sunrise};
\draw[orange,thick,dashed] (axis cs:18,0) -- (axis cs:18,28);
\node[font=\scriptsize,orange,anchor=south,rotate=90] at (axis cs:18,14) {Sunset};

\end{axis}
\end{tikzpicture}
\caption{Diurnal MUF variation showing the $\sqrt{\cos\chi}$ shape
         scaled to a peak of \SI{20}{\MHz} (representative midlat
         $F_{10.7\text{A}} \approx 150$).  The dashed gray line at
         \SI{8}{\MHz} is the night-floor product
         $F \cdot \text{MUF}_{\text{peak}}$ from
         Eq.~\ref{eq:nightfloor}: $F$ is the dimensionless
         illumination ratio (${\sim}0.40$ at midlat moderate solar
         activity), which when applied to a \SI{20}{\MHz} daytime
         peak gives the \SI{8}{\MHz} floor curve plotted here.
         Shaded regions = night.}
\label{fig:mufdiurnal}
\end{figure}

% ═════════════════════════════════════════════════════════════════════
\section{Forward Projection (next 24\,h)}\label{sec:projection}

\begin{quote}\small\itshape
\textbf{Implementation status (entire section).}  The forward
projection pipeline described in this section --- MUF projection
(Eq.~\ref{eq:muffut}), seasonal ratio
(Section~\ref{sec:seasonalratio}), storm depression
(Section~\ref{sec:stormdepression}), and sporadic-E persistence
(Section~\ref{sec:espersistence}) --- is documented for a future
build of ionocast that exposes a projected outlook MUF as a
first-class display.  The gray-line bonus
(Section~\ref{sec:grayline}) is the exception: it already runs in
the live verdict pipeline (\texttt{grayLineBonusDb} in
\texttt{src/physics/modes.js}, called from
\texttt{src/derive/conditions.js}) as one of the additive bonuses
in Eq.~\ref{eq:blend}, so this section's treatment of it is the
\emph{deployed} formulation rather than a forward look.
None of the \emph{remaining} forward-projection pipeline (MUF
projection, seasonal ratio, storm depression, Es persistence)
currently runs in the live runtime; the live runtime instead
surfaces SWPC's own next-72\,h forecasts on the Outlook panel.
Several of the sub-pieces are however shared with the live
budget --- specifically
$d(\cos\chi)$ (Eq.~\ref{eq:fof2-driver}) and the storm-lag
effective $K_p$ (Section~\ref{sec:kplag}, which feeds
$L_{\text{aur}}$ and the $\sigma$-widening of
Section~\ref{sec:tier-confidence} in the \emph{live} pipeline);
the §\ref{sec:stormdepression} inset spells out the live-vs-deferred
split for the $F_{\text{storm}}$ MUF multiplier specifically.
Read this section's equations as the form a future outlook-MUF
display would use, not as forward-tense equations whose absence
from runtime makes them operationally inert: the parts that
touch the live pipeline are already running.
\end{quote}

The forward projection engine extends the current nowcast into the
future by modelling the dominant drivers of ionospheric change.

\subsection{MUF Projection via Solar Zenith Analogy}\label{sec:muf-now-vs-future}

Eq.~\ref{eq:muffut} below is specifically the \emph{F2} MUF
projection.  Sporadic-E persistence (\S\ref{sec:espersistence})
is a separate sub-piece: it predicts the projected
$f_{\text{oEs}}$ used by the Es-mode budget, not a multiplicative
factor on F2 MUF.  The four F2 factors below (a zenith ratio
through $d(\cos\chi)$, a seasonal ratio, and a storm-factor
ratio) combine multiplicatively because each ratio reads as the
\emph{change} between the now-state encoded in
$\text{MUF}_{\text{now}}$ and the future-horizon equivalent.
\begin{equation}\label{eq:muffut}
  \text{MUF}_{\text{future}} = \text{MUF}_{\text{now}} \cdot
    \frac{S(d(\cos\chi_{\text{future}}),\, F)}{S(d(\cos\chi_{\text{now}}),\, F)}
    \cdot R_{\text{seasonal}} \cdot \frac{F_{\text{storm}}(K_p^{\text{fut}})}
                                          {F_{\text{storm}}(K_p^{\text{now}})}
\end{equation}
where $K_p^{\text{fut}}$ is the SWPC 3-day forecast value at
$t + \Delta$ (or the current $K_p$ when $\Delta = 0$).  Both the
``now'' and ``future'' zenith arguments pass through the same
3-hour memory-lag function $d(\cos\chi)$ defined in
Eq.~\ref{eq:fof2-driver} (Section~\ref{sec:climo}); using raw
$\cos\chi$ on either side without the lag would produce a
discontinuity at the projection's $\Delta = 0^+$ boundary
relative to the live nowcast (which already runs through
$d(\cos\chi)$), most visibly at sunset where the live model holds
foF2 elevated for ${\sim}3$\,h via memory while a raw-cos$\chi$
projection would drop it instantly.  The storm factor enters as a
\emph{ratio} so the projection only adjusts for the \emph{change}
in storm phase between now and the projection horizon: if the
current $K_p$ already encodes the disturbance,
$\text{MUF}_{\text{now}}$ is already depressed and the ratio is
near unity; if a storm is forecast to arrive within $\Delta$ but
hasn't yet, $K_p^{\text{fut}} > K_p^{\text{now}}$ depresses the
projection accordingly. This complements the $\sigma$-side
forecast bump in Section~\ref{sec:reliability} (which widens
$\sigma$ ahead of a forecast disturbance) by adjusting the
deterministic mean as well; together they avoid the failure mode
where $\text{MUF}_{\text{now}}$ encodes a quiet ionosphere but a
storm has been forecast for the horizon.

\subsection{Seasonal Ratio}\label{sec:seasonalratio}

\begin{equation}
  R_{\text{seasonal}} = \frac{s(t_{\text{future}},\,\phi)}{s(t_{\text{now}},\,\phi)}
\end{equation}
where
\begin{equation}\label{eq:seasonal}
  s(t,\phi) = 1 + (0.15\cos\theta + 0.08\cos(2\theta + \pi))
              \cdot w(\phi)
\end{equation}
with $\theta = 2\pi\,(d_{\text{yoy}} - d_{\text{solstice}})/365$,
where $d_{\text{yoy}}$ is the UTC day of year ($1 = 1$~January) and
$d_{\text{solstice}}$ is the local-hemisphere winter-solstice day:
$d_{\text{solstice}} = 355$ (December~21) for $\phi \geq 0$ and
$d_{\text{solstice}} = 172$ (June~21) for $\phi < 0$.  The
day-of-year phase places the cosine peak at the astronomical
solstice and matches Eq.~\ref{eq:winter} in the climatology so
the two seasonal effects share the same phase reference.
$w(\phi) = \min(1, |\phi|/50)$ scales the modulation toward
midlatitudes.  The first term captures the annual
winter-peak/summer-trough at midlatitudes; the second captures
the semi-annual equinox peaks (March, September) evident in
F2-layer statistics.  Sharing the day-of-year phase reference
with Eq.~\ref{eq:winter} keeps the two seasonal effects (which
overlap in the annual component) aligned.

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=12cm, height=6cm,
  xlabel={Month},
  ylabel={Seasonal Factor $s(t,\phi)$},
  xmin=0.5, xmax=12.5,
  ymin=0.76, ymax=1.24,
  xtick={1,2,3,4,5,6,7,8,9,10,11,12},
  xticklabels={J,F,M,A,M,J,J,A,S,O,N,D},
  ytick={0.80,0.85,0.90,0.95,1.00,1.05,1.10,1.15,1.20},
  grid=major,
  grid style={gray!30},
  legend pos=south east,
  legend style={font=\small},
  title={Seasonal MUF Factor at $45\text{\textdegree}$N},
  title style={font=\small\bfseries},
]

% s(t,phi) = 1 + (0.15*cos(theta) + 0.08*cos(2*theta + pi)) * w
% theta = 2*pi*(doy - d_solstice)/365 per Eq:seasonal (day-of-year phase),
% with d_solstice = 355 (Dec 21) for NH.
% Mid-month doy(x) approximation:  doy = 30.4*(x-1) + 15.2.
% So theta(x) = 2*pi*(30.4*(x-1) + 15.2 - 355)/365.
% w(45) = min(1, 45/50) = 0.9.
% NH peak at theta=0: doy=355 → x ≈ 12.18 (late Dec, wrapping to early month axis).

% Annual component only
\addplot[blue,dashed,domain=1:12,samples=100]
  {1 + 0.15*cos(deg(2*pi*(30.4*(x-1) + 15.2 - 355)/365)) * 0.9};
\addlegendentry{Annual only}

% Semi-annual only
\addplot[red,dashed,domain=1:12,samples=100]
  {1 + 0.08*cos(deg(2*(2*pi*(30.4*(x-1) + 15.2 - 355)/365) + pi)) * 0.9};
\addlegendentry{Semi-annual only}

% Combined
\addplot[black,very thick,domain=1:12,samples=100]
  {1 + (0.15*cos(deg(2*pi*(30.4*(x-1) + 15.2 - 355)/365)) + 0.08*cos(deg(2*(2*pi*(30.4*(x-1) + 15.2 - 355)/365) + pi))) * 0.9};
\addlegendentry{Combined}

% Equinox markers
\draw[green!50!black,thick,dotted] (axis cs:3,0.76) -- (axis cs:3,1.24);
\draw[green!50!black,thick,dotted] (axis cs:9,0.76) -- (axis cs:9,1.24);
\node[font=\tiny,green!50!black] at (axis cs:3,1.23) {Equinox};
\node[font=\tiny,green!50!black] at (axis cs:9,1.23) {Equinox};

\end{axis}
\end{tikzpicture}
\caption{Seasonal MUF factor at $45\text{\textdegree}$N showing annual
         (winter peak) and semi-annual (equinox peaks) components.
         At $|\phi| = 45\text{\textdegree}$, $w(\phi) = 45/50 = 0.9$,
         so the annual component reaches $1 + 0.15 \cdot 0.9 = 1.135$
         in isolation at the winter solstice and the semi-annual
         component reaches $1 + 0.08 \cdot 0.9 = 1.072$ in isolation at
         the equinoxes.  Because the two components compete at the
         winter solstice --- the semi-annual is at its trough there
         (\,$\cos(2\theta + \pi) = -1$ at $\theta = 0$\,) --- the
         combined curve is flatter than either component alone: the
         winter solstice evaluates to $1 + (0.15 - 0.08) \cdot 0.9 =
         1.063$, the March / September equinoxes to $1 + 0.08 \cdot
         0.9 = 1.072$, and the summer solstice to $1 - (0.15 + 0.08)
         \cdot 0.9 = 0.793$.  The dominant seasonal feature at this
         latitude is therefore the deep summer trough rather than a
         sharp winter peak; winter and equinoxes form a broad plateau
         in the $1.06$--$1.07$ range, with the equinoxes slightly
         above the winter solstice ($+0.009$).  Above $|\phi| =
         50\text{\textdegree}$ the weight saturates at $w = 1$ and the
         summer trough deepens to $0.77$ while the winter / equinox
         plateau settles at $1.07$--$1.08$.}
\label{fig:seasonal}
\end{figure}

\subsection{Storm Depression (forward-projection definition)}\label{sec:stormdepression}

The forward-projection storm depression follows the F-region storm
phenomenology surveyed by Pr\"olss~\cite{prolss1995}, Fuller-Rowell
et al.~\cite{fullerrowell1994}, and Mendillo~\cite{mendillo2006}: a
prompt depression of foF2 driven by Joule heating of the
thermosphere and the resulting O/N$_2$ composition shift, with a
recovery timescale set by the storm's driver (impulsive CME vs
sustained HSS).

\begin{quote}\small\itshape
\textbf{Implementation status.}  The expressions in this subsection
define the $F_{\text{storm}}$ multiplier referenced by Eq.~\ref{eq:muffut}
(the forward-projection MUF for the outlook section). The
\emph{current-time} MUF in production does \emph{not} apply this
multiplier: the runtime budget reflects geomagnetic storm response
through the storm-lagged effective $K_p$ feeding $L_{\text{aur}}$
(Section~\ref{sec:laur} and Section~\ref{sec:kplag}), not as a
MUF-side scalar.  $F_{\text{storm}}$ as a real-time MUF multiplier
is reserved for a future build that exposes the projected outlook
MUF as a first-class display.
\end{quote}

When $K_p > 4$, the projected MUF is depressed; for $K_p \le 4$ the
factor holds at unity (no quiet-day boost):
\begin{equation}\label{eq:storm}
  F_{\text{storm}} = \max\!\left(F_{\text{floor}}(K_p),\;
    1 - p\,\max(0,\,K_p - 4)\right)
\end{equation}
where
\begin{equation}
  p = 0.05 + 0.10 \cdot \text{clamp}\!\left(
    \frac{|\phi_{\text{CGM}}| - 40}{30},\;0,\;1\right)
\end{equation}
The two latitude anchors place the slope ramp where the underlying
F-region storm physics actually transitions.  $|\phi_{\text{CGM}}| =
40\degree$ is the approximate ceiling of the trough region during
strong storms in the cycle-25 Dst-vs-foF2 record (consistent with
the qualitative trough-edge framing in
Mendillo~\cite{mendillo2006} and Pr\"olss~\cite{prolss1995},
though neither prints a literal 40\degree\ figure --- the
specific value is an ionocast calibration anchor): below this,
the storm-time foF2 depression
is dominated by composition-shift effects whose magnitude is well
captured by a gentle $p = 0.05$.  $|\phi_{\text{CGM}}| = 70\degree$
is the equatorward edge of the auroral oval at $K_p \approx 7$
(the storm-zone expansion threshold of
Eq.~\ref{eq:auroralexpansion}'s converged state); poleward of it,
the F-region sees direct precipitation-driven heating in addition
to the global thermospheric shift, and a steeper $p = 0.15$ matches
the ${\sim}50\%$ MUF collapses observed in IRI-validated ground
campaigns.  The $30\degree$ ramp width connects them linearly; this
is geometric interpolation rather than a separate physics prediction.
The floor itself drops below 0.5 during severe storms to capture
the deep MUF collapse documented for $K_p \geq 8$ events:
\begin{equation}\label{eq:stormfloor}
  F_{\text{floor}}(K_p) = \max\!\Bigl(
    0.30,\; 0.50 - 0.10 \cdot \min\!\bigl(2,\,\max(0,\,K_p - 7)\bigr)
  \Bigr)
\end{equation}
At quiet to moderate $K_p$ the floor is at 0.5 (50\% of quiet
baseline). At $K_p = 8$ (G4) it drops to 0.4; at $K_p = 9$ (G5)
it drops to 0.3.  The crossover $K_p$ at which the floor binds
(i.e.\ the main term drops below the floor schedule) is
\begin{equation}\label{eq:stormfloorcross}
  K_p^{*} = 4 + \frac{1 - F_{\text{floor}}(K_p^{*})}{p}
\end{equation}
which evaluates to exactly $8.0$ on auroral paths
($p = 0.15$, $|\phi_{\text{CGM}}| \ge 70\degree$) --- so for
$K_p \ge 8.0$ the floor schedule binds at polar latitudes
(at $K_p = 8$ both the main term $1 - 0.15 \cdot 4 = 0.40$ and
the floor $0.50 - 0.10 \cdot 1 = 0.40$ coincide; at $K_p = 9$
the main term gives $1 - 0.15 \cdot 5 = 0.25$, the floor gives
$0.50 - 0.10 \cdot 2 = 0.30$, and $\max(0.30, 0.25) = 0.30$).  At midlatitude paths the slope
$p = 0.05$ keeps the main term above the floor for any
plausible $K_p$ ($K_p^{*} \approx 18$ at $|\phi_{\text{CGM}}|
= 40\degree$, far past the $K_p = 9$ ceiling), so the floor
schedule \emph{never binds at midlat} and the floor refinement
is auroral-only by construction.  The schedule captures the deep
polar-F2-MUF collapses below 50\% of baseline observed during
super-storms (Halloween~2003, March~1989) on the auroral paths
those events actually affect.

At midlatitudes ($|\phi_{\text{CGM}}| = 40\text{\textdegree}$), the
damping is 5\%~per~$K_p$ unit above~4.  In the auroral zone
($|\phi_{\text{CGM}}| = 70\text{\textdegree}$), it rises to
${\sim}15$\% per unit.

\paragraph{Negative phase only.}  $F_{\text{storm}}$ models the
\emph{negative-phase} F-region storm response (the persistent
electron-density depression caused by neutral composition changes
and Joule heating that lasts hours-to-days into the storm main and
recovery phases).  It does not model the \emph{positive
phase} that can precede the depression in the first several hours
after a sudden CME impact, where penetration electric fields and
$E\times B$ uplift can transiently \emph{enhance} foF2 by
15--50\,\% above the quiet-time baseline (typical durations
${\sim}\!1$--$\SI{4}{\hour}$, magnitude scaling with storm
intensity and latitude per
Buonsanto~\cite{buonsanto1999} and Mendillo~\cite{mendillo2006}).  The result is that the model
under-predicts MUF during the positive-phase window of a
fast-onset CME event (a quiet-time prediction is correct; the
depression engages correctly once the negative phase takes hold,
which is typically within a few hours of the impact).  In
practice this matters for operators chasing the positive-phase
opening on the upper bands during the first 1--4\,hours of a
sudden storm --- a relatively rare and short-lived mode --- the
negative-phase damping that follows is what shapes the bulk of
the storm-affected propagation window.  When the
projected-MUF panel ships (currently documented but not yet
surfaced --- see implementation status above), it should carry an
annotation flagging this gap, paralleling how the scatter
(\S\ref{sec:scatter}) and TEP (\S\ref{sec:tep}) modes are tagged on
the verdict line.  Until then, operators tracking a fresh CME
should trust the alerts-panel $B_z$ / $D_{\text{st}}$ / PCA chips
for the impact signal rather than expecting the projection to flip
positive in the first hour.

\subsubsection{Storm-Lag Effective $K_p$}\label{sec:kplag}

The F-region does not respond instantaneously to the $K_p$ kick: Joule
heating of the thermosphere and subsequent composition changes take
1--3\,hours to depress the peak electron density, and the ionosphere
then recovers over the following 12--24\,hours.  Feeding the raw
instantaneous $K_p$ into the physics budget therefore mis-times both
the onset and the recovery of a storm.

ionocast builds a Laplace-kernel-weighted \emph{effective} $K_p$
from the 7-day SWPC $K_p$ history (3-hour cadence): the kernel
$\exp(-|\Delta t - \tau_{\text{peak}}|/\tau_{\text{decay}})$ is
symmetric about the peak rather than one-sided exponential, with
$\tau_{\text{peak}} = \SI{2}{\hour}$ in the past and a decay
constant $\tau_{\text{decay}}$ that depends on the storm's
driver:
\begin{equation}\label{eq:kplag}
  K_p^{\text{eff}} = \frac{
    \sum_i K_p(t_i)\,
    \exp\!\bigl(-|\Delta t_i - \tau_{\text{peak}}|/\tau_{\text{decay}}\bigr)
  }{
    \sum_i \exp\!\bigl(-|\Delta t_i - \tau_{\text{peak}}|/\tau_{\text{decay}}\bigr)
  }
\end{equation}
with $\Delta t_i = t_{\text{now}} - t_i$ (positive for past samples).

\paragraph{Storm-type classification.}  CME-driven and CIR / HSS-driven
storms recover on very different timescales.  A CME arrival is an
impulse: after the shock and main phase, Joule heating dissipates
over ${\approx}\SI{8}{\hour}$.  A co-rotating interaction region
(CIR) or high-speed stream (HSS) is a sustained driver that keeps
heating the thermosphere while the stream flows past Earth -- recovery
takes ${\approx}\SI{24}{\hour}$ or longer.  ionocast picks the kernel
as follows:
\begin{itemize}[nosep]
  \item If $\text{Dst} < -\SI{80}{nT}$ the ring current is deep: CME
        signature dominates, $\tau_{\text{decay}} = \SI{8}{\hour}$.
        (If $\text{Dst}$ is unavailable from the upstream feed,
        skip to the next branch rather than treating the missing
        value as $\geq -\SI{80}{nT}$.)
  \item Otherwise, if NASA DONKI~\cite{donki} reports an HSS event within the last
        \SI{24}{\hour} or next \SI{48}{\hour}, use
        $\tau_{\text{decay}} = \SI{24}{\hour}$.  The asymmetric
        window is intentional: DONKI's HSS analyses are typically
        published within hours of stream onset (so a 24-hour
        lookback covers events the catalogue has already classified)
        but stream peak effects on the magnetosphere can lag the
        L1 arrival by 12--36\,hours and persist for another
        12--24\,hours, so a 48-hour lookahead extends the
        24-hour-decay regime through the stream's effective
        magnetospheric tail.  The window mixes two information
        types: the lookback portion is purely catalogue-historical
        (events DONKI has already classified) while the lookahead
        portion is partly forecast (events DONKI has flagged from
        L1 arrival or solar imaging but whose magnetospheric
        impact is still developing).  $\tau_{\text{decay}}$ is
        therefore a forward-looking parameter chosen from a
        mix of recent and announced catalogue entries; the
        $K_p^{\text{eff}}$ kernel it controls is then applied to
        the strictly-historical $K_p$ time series.
  \item Otherwise default to $\tau_{\text{decay}} = \SI{8}{\hour}$.
\end{itemize}
Only $K_p^{\text{eff}}$ enters the physics; the UI still displays the
instantaneous sample so the shown number matches SWPC's feed.

\subsubsection{Dst, IMF Bz, and Solar-Wind Drivers}\label{sec:fast-storm-drivers}

The 3-hour Kp index is a lagging indicator.  Three faster-responding
inputs are wired into the storm model:

\begin{itemize}[nosep]
  \item \textbf{Dst} (Kyoto ring-current index): when $\text{Dst} < -50$\,nT,
        the effective $K_p$ is bumped by up to $+2$ (1 per 50\,nT below $-50$).
        Dst responds within minutes of storm onset, 1--3 hours before Kp updates.
  \item \textbf{IMF Bz forward bump} (DSCOVR/ACE L1 solar wind): the
        L1 monitor sits ${\sim}\SI{1.5e6}{km}$ upstream, giving a
        nominal 30--60\,min lead on the geomagnetic effect at
        Earth (the spread on the lead time depends on solar-wind
        speed: $\SI{1.5e6}{km} / \SI{400}{km/s} = 62$\,min at
        average speeds, $\SI{1.5e6}{km} / \SI{800}{km/s} = 31$\,min
        at fast streams).  The L1-to-Earth correlation is
        \emph{probabilistic}, not deterministic: not every L1-observed
        Bz dip arrives at Earth's magnetopause unchanged --- some
        structures are spatially extended and miss Earth entirely
        ($Y_{\text{GSE}} \neq 0$), some glance the magnetopause
        without sustained reconnection, and some get reshaped by
        intermediate solar-wind structures during transit.  The
        20-min median is gated at $-5$\,nT: medians at or above
        $-5$\,nT (i.e.\ $B_z \geq -5$) produce no bump; medians
        strictly below $-5$\,nT (i.e.\ $B_z < -5$) ramp the bump per
        Table~\ref{tab:bzbump}.  The cutoff filters out short
        reconnection bursts that don't drive sustained magnetopause
        coupling, but the L1 forward bump should be treated as an
        early-warning probability bump on the effective $K_p$, not
        a guaranteed forecast.  When the median Bz over the last
        20 min is sustained below threshold, the effective $K_p$ is
        bumped additively \emph{before} the index reading catches
        up (Table~\ref{tab:bzbump}); single-sample dips below
        threshold do not count.  The bump decays to zero once
        $B_z$ returns to $\geq -5$\,nT.
  \item \textbf{Solar-wind plasma} (DSCOVR/ACE): sustained speed
        $\geq \SI{500}{km/s}$ with sharply negative Bz
        ($\leq -8$\,nT) is read as a CME shock arrival regardless of
        DONKI catalogue inertia, while the same speed with mild Bz
        ($> -5$\,nT) is read as HSS / corotating-stream interaction.
        The intermediate regime (\SI{500}{km/s} or faster with
        $-8 < B_z \le -5$\,nT) is ambiguous in the live signal ---
        moderate southward Bz with elevated speed could be either
        the leading edge of a CME sheath or sustained HSS-driver
        reconnection --- so neither shortcut fires; the classifier
        falls through to the catalogue / Dst logic below, which
        defaults to the CME timescale ($\tau_{\text{decay}} =
        \SI{8}{h}$) unless an active DONKI HSS event covers the
        current window.  Storm-type classification picks the decay
        time-scale ($\tau_{\text{decay}} = \SI{8}{h}$ vs.\
        \SI{24}{h}) of the Section~\ref{sec:kplag} kernel.
\end{itemize}

\paragraph{Geomagnetic-input clamps.}  All adjustments above are
capped at $K_p = 9$ and apply only when the respective input is
available.  The auroral CGM threshold expands
continuously from $60\degree$ (quiet, $K_p \le 5$) to $50\degree$
(severe, $K_p \ge 7$), per Eq.~\ref{eq:auroralexpansion} ---
linear in $K_p$ across the $5 \to 7$ window so the equatorward
expansion of the auroral oval enters the budget gradually rather
than stepping at a single threshold.

\begin{table}[H]
\centering
\caption{Effective $K_p$ bump from a sustained southward-Bz median
         observed at the L1 monitor (DSCOVR / ACE) over the trailing
         20 minutes. The function is a continuous linear ramp from
         0 at $B_z = -5$\,nT to $+3$ at $B_z = -15$\,nT, saturating
         at $+3$ for deeper southward IMF. Table values are sample
         points along the ramp.}
\label{tab:bzbump}
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Median Bz over 20 min} & \textbf{$K_p$ bump} \\
\midrule
$\geq -5$\,nT   & $0$            \\
$-6$\,nT        & $+0.3$         \\
$-8$\,nT        & $+0.9$         \\
$-10$\,nT       & $+1.5$         \\
$-12$\,nT       & $+2.1$         \\
$-15$\,nT       & $+3$ (saturated) \\
$\leq -15$\,nT  & $+3$ (saturated) \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=11cm, height=7cm,
  xlabel={$K_p$},
  ylabel={$F_{\text{storm}}$ (MUF factor)},
  xmin=4, xmax=9,
  ymin=0.3, ymax=1.05,
  xtick={4,5,6,7,8,9},
  ytick={0.4,0.5,0.6,0.7,0.8,0.9,1.0},
  grid=major,
  grid style={gray!30},
  legend pos=north east,
  legend style={font=\small},
  title={Storm Depression by CGM Latitude},
  title style={font=\small\bfseries},
]

% Ffloor(Kp) = max(0.30, 0.50 - 0.10*min(2, max(0, Kp-7)))
% Midlat: CGM = 40°, p = 0.05 + 0.10*clamp((40-40)/30,0,1) = 0.05
% Midlat floor: Ffloor never binds (K*_p ~18), so max(0.5,...) is correct here
\addplot[blue,very thick,domain=4:9,samples=50]
  {max(0.5, 1 - 0.05*(x-4))};
\addlegendentry{Midlat (CGM $40\text{\textdegree}$, $p=0.05$)}

% Auroral: CGM = 65°, p = 0.05 + 0.10*clamp((65-40)/30,0,1) = 0.05 + 0.10*0.833 = 0.133
% Auroral floor: Ffloor drops to 0.40 at Kp=8 and 0.30 at Kp=9 (Eq. stormfloor)
\addplot[red,very thick,domain=4:9,samples=200]
  {max(max(0.30, 0.50 - 0.10*min(2, max(0, x-7))), 1 - 0.133*(x-4))};
\addlegendentry{Auroral (CGM $65\text{\textdegree}$, $p{\approx}0.13$)}

% Ffloor schedule reference line (auroral only)
\addplot[gray,dashed,domain=4:9,samples=200]
  {max(0.30, 0.50 - 0.10*min(2, max(0, x-7)))};
\addlegendentry{$F_{\text{floor}}(K_p)$ (auroral)}

\end{axis}
\end{tikzpicture}
\caption{Storm MUF depression factor vs.\ $K_p$ for midlatitude and
         auroral-zone paths.  The midlatitude floor never binds (slope
         $p=0.05$ keeps the main term above 0.5 for all $K_p \le 9$).
         The auroral curve plotted is at $|\phi_{\text{CGM}}| =
         65\degree$ (where the slope $p \approx 0.13$ is not yet at
         its $|\phi_{\text{CGM}}| \ge 70\degree$ ceiling of $0.15$);
         the floor schedule
         $F_{\text{floor}}(K_p)$ (Eq.~\ref{eq:stormfloor})
         drops from 0.5 at $K_p < 7$ to 0.40 at $K_p=8$ and 0.30
         at $K_p=9$ but only \emph{just} starts to bind on this
         curve at $K_p \approx 9$.  The clean ``main term and floor
         coincide at 0.40 at $K_p = 8$'' result discussed in
         Eq.~\ref{eq:stormfloorcross} applies at
         $|\phi_{\text{CGM}}| \ge 70\degree$, not at the
         65\textdegree{} curve plotted here.}
\label{fig:storm}
\end{figure}

\subsection{Sporadic-E Persistence}\label{sec:espersistence}

The projected critical frequency of the sporadic-E layer decays
exponentially:
\begin{equation}\label{eq:es}
  f_{\text{oEs}}(t + \Delta) = \max\!\left(B(\cos\chi),\;
    f_{\text{oEs}}(t) \cdot 2^{-\Delta/\tau(\cos\chi)}\right)
\end{equation}
with the half-life and floor smoothly interpolated across the
twilight band $\cos\chi \in [0.1, 0.3]$ rather than stepped at a
single $\cos\chi = 0.2$ threshold.  The interpolated forms are
\begin{equation}
  \tau(\cos\chi) = \SI{1.5}{\hour} +
    \SI{1.5}{\hour}\cdot R\!\left(\tfrac{\cos\chi - 0.1}{0.2}\right),
  \qquad
  B(\cos\chi) = \SI{2}{\MHz}\cdot R\!\left(\tfrac{\cos\chi - 0.1}{0.2}\right)
\end{equation}
where $R(x) = \mathrm{clamp}(x, 0, 1)$ is a unit ramp (renamed
from $\sigma(x)$ used in earlier drafts to avoid collision with
the prediction-spread $\sigma$ family of \S\ref{sec:reliability}).  At
$\cos\chi \le 0.1$ both reduce to their night values
($\tau = 1.5\,\text{h}$, $B = 0$); at $\cos\chi \ge 0.3$ both
saturate to the day values ($\tau = 3.0\,\text{h}$, $B = 2\,\text{MHz}$).
At deep twilight ($\cos\chi = 0.2$) the function is at midpoint
($\tau = 2.25\,\text{h}$, $B = 1\,\text{MHz}$), and the band-MUF
gating threshold transitions continuously rather than stepping.

This subsection describes the documented forward-projection
persistence model; it is part of the deferred outlook MUF
projection (Section~\ref{sec:muf-now-vs-future}) and is not
exercised by the live runtime budget at $t = 0$.

\subsection{Gray-Line Bonus (Asymmetric)}\label{sec:grayline}

The grayline bonus reads as the D-region absorption refund
relative to local-noon-overhead, paired with the
$L_{\text{Dreg}}$ charge of \S\ref{sec:labsd} via a shared
$L_{\text{day}}(f)$ anchor (S0~\#2 paired rewrite, landed
2026-05-07; the per-band sunrise/sunset table the prior form
used is retired).  In closed form,
\begin{equation}\label{eq:gl}
  B_{\text{grayline}}(f, \chi) =
    \min\!\Bigl(
      \SI{25}{\dB},\;
      \alpha(\dot\chi) \cdot L_{\text{day}}(f) \cdot
        \bigl(1 - (\max(0, \cos\chi))^{0.7}\bigr)
    \Bigr),
\end{equation}
where $L_{\text{day}}(f) = K_D / (f_{\text{MHz}} + 0.5)^{2}$ from
Eq.~\ref{eq:labsd}, the inner factor
$1 - \cos^{0.7}(\chi)$ goes to $1$ at the terminator
($\cos\chi \to 0$, so the bonus saturates at $L_{\text{day}}(f)$
or the cap, whichever is smaller) and to $0$ at local noon
($\cos\chi = 1$); a pre-floor at \SI{0.2}{\dB} suppresses
sub-decibel noise, and the sunrise/sunset asymmetry multiplier
$\alpha(\dot\chi)$ is
\begin{equation*}
  \alpha(\dot\chi) =
    \begin{cases}
      1.0 & \text{if } \mathrm{d}\!\cos\chi/\mathrm{d}t > 0
            \quad\text{(rising, sunrise)} \\
      0.5 & \text{if } \mathrm{d}\!\cos\chi/\mathrm{d}t < 0
            \quad\text{(falling, sunset)}
    \end{cases}
\end{equation*}
sampled 5\,minutes ahead.  The asymmetry preserves the prior
table's roughly $2\!:\!1$ sunrise:sunset ratio (D-region
recovers more slowly at dusk than the residual F-region
enhancement persists, producing a shorter and weaker window
than at dawn) without per-band buckets.  The frequency window
is $1.8 \le f \le \SI{30}{\MHz}$; outside this range the term
returns zero.

\begin{equation*}
\begin{array}{lr@{\,}c@{\,}r@{\quad}lr@{\,}c@{\,}r}
\textbf{Band} & f & & B_{\max} &
\textbf{Band} & f & & B_{\max} \\
\midrule
160\,\text{m} & 1.838~\si{\MHz} & \to & 25~\si{\dB}~(\text{capped}) &
20\,\text{m}  & 14.097~\si{\MHz} & \to & 0.94~\si{\dB} \\
80\,\text{m}  & 3.570~\si{\MHz}  & \to & 12.1~\si{\dB} &
17\,\text{m}  & 18.106~\si{\MHz} & \to & 0.58~\si{\dB} \\
60\,\text{m}  & 5.366~\si{\MHz}  & \to & 5.80~\si{\dB} &
15\,\text{m}  & 21.096~\si{\MHz} & \to & 0.43~\si{\dB} \\
40\,\text{m}  & 7.040~\si{\MHz}  & \to & 3.52~\si{\dB} &
12\,\text{m}  & 24.924~\si{\MHz} & \to & 0.31~\si{\dB} \\
30\,\text{m}  & 10.140~\si{\MHz} & \to & 1.76~\si{\dB} &
10\,\text{m}  & 28.126~\si{\MHz} & \to & 0.24~\si{\dB} \\
\end{array}
\end{equation*}

\noindent
$B_{\max}$ in the table above is the sunrise-side terminator
peak ($\alpha = 1$, $\cos\chi \to 0$); sunset values are half
those after $\alpha(\dot\chi)$ multiplies in.  160\,m saturates
the \SI{25}{\dB} cap at the terminator (the underlying
$L_{\text{day}}$ is \SI{36.6}{\dB} so the literal formula
without cap would refund the full noon loss).  The cap is
operational --- per-path absorption already saturates at
\SI{50}{\dB} via Eq.~\ref{eq:labsd-path}, so an unbounded
grayline bonus would dominate the budget on near-terminator
160\,m paths beyond what the calibration basket can validate.
Below the \SI{0.2}{\dB} floor the term returns zero, which
keeps low-magnitude refunds (deep daytime) out of the verdict
loop.

The closed-form replacement is calibration-pending: the prior
discrete sunrise/sunset table was hand-tuned against the WSPR
per-pair basket that also fixed $w_{\text{sc}}$, and the new
formula will need a comparable retune before the bonus
magnitudes can be claimed as fitted rather than principled.
The S0~\#2 framing accepted this trade-off explicitly to land
the physics shape now and re-fit later.

\paragraph{Midpoint vs.\ per-hop evaluation.}  The proximity
$|\cos\chi|$ is computed at the reflection \emph{midpoint} only,
not at each hop along the great circle.  This is a deliberate
simplification: the D-region absorption \emph{loss} term
(Eq.~\ref{eq:labsd-path}) is evaluated per-hop and summed, but the
gray-line \emph{bonus} compensates loss at the terminator, and
on a 4-hop transatlantic path whose midpoint sits in midday but
whose terminus hop is at sunrise the per-hop loss already drops on
that one hop's $L_{\text{Dreg}}$ contribution; adding a
midpoint-only bonus on top would over-count the local effect on
the geometric-midpoint hop while under-counting the genuine
sunrise hops at the path ends.  The midpoint-only formulation
captures the case the bonus is operationally intended for ---
short-to-medium HF paths (1--2 hops, 1500--6000\,km) whose single
or twin reflection points cross the terminator together, where
``the path is on the gray line'' and ``the midpoint is at the
terminator'' coincide.  For longer multi-hop paths (5000\,km+)
where one terminus is in daylight and the other near sunrise, the
per-hop $L_{\text{Dreg}}$ already does the right thing on its own.
Extending the bonus to a per-hop max is a candidate refinement
tracked in Section~\ref{sec:limits}; today's midpoint-only form
is simpler to reason about and avoids the over-count.

% ═════════════════════════════════════════════════════════════════════
\section{Alternative Propagation Modes}\label{sec:altprop}

The F2-layer SNR budget (Section~\ref{sec:snr}) is the default path
for every HF circuit, but four additional modes can open bands that
the F2 budget marks closed: sporadic-E (\S\ref{sec:es-prop-mode}),
trans-equatorial propagation (\S\ref{sec:tep}), meteor scatter
(\S\ref{sec:ms}), and F2-region scatter recovery
(\S\ref{sec:scatter}).  Each runs in parallel with the F2 budget on
the same path set; the verdict takes whichever mode has the best
margin.  Tropospheric ducting (\S\ref{sec:tropo}) is also covered
in this section but is structurally different --- it is surfaced as
a measured per-station panel rather than folded into the budget,
because a duct's spatial extent is not resolved by the sparse
radiosonde basket the gradient is computed from.

\subsection{Sporadic-E as a Propagation Mode}\label{sec:es-prop-mode}

The Es-screening term ($L_{\text{Es}}$, Section~\ref{sec:snr}) only
accounts for Es \emph{blocking} an F2 hop.  In reality Es can itself
be the propagation mode (the framework for which is ITU-R
P.534~\cite{itup534}): a single ${\sim}\SI{2000}{\km}$ hop off the
E layer at ${\sim}\SI{110}{\km}$ altitude, with no F2 variability and
no auroral absorption.  This is the mechanism behind summer-evening
15\,m, 12\,m, 10\,m, and 6\,m openings that the F2-only model would
otherwise call closed.  6\,m only opens via Es when $f_{\text{oEs}}
\gtrsim \SI{10}{\MHz}$ (so that $5\,f_{\text{oEs}} \ge \SI{50}{\MHz}$),
which is rarer than the HF-band thresholds and concentrated in the
boreal-summer Es season.

ionocast defines a sporadic-E budget alongside the F2 one:
\begin{equation}\label{eq:esmargin}
  M_{\text{Es}} = P_{\text{tx}} + G_{\text{ant}}
    - L_{\text{fs}}(f, d_{\text{Es}})
    - L_{\text{Dreg}} - L_{\text{PCA}} - L_{\text{flare}}
    - L_{\text{MUF,Es}}(f, 5 f_{\text{oEs}})
    - L_{\text{iono,Es}} - L_{\text{low}}
    - N - S_{\text{req}}
\end{equation}
with $d_{\text{Es}} = \SI{2000}{\km}$, a single-hop takeoff angle
$\approx\!6.3\degree$, and $L_{\text{iono,Es}} = \SI{15}{\dB}$.

\paragraph{What's in $L_{\text{iono,Es}} = \SI{15}{\dB}$.} The
lumped 15\,dB is substantially larger than the F2 term
$L_{\text{iono}} = \SI{1}{\dB}$ (Section~\ref{sec:liono}) because
the E-layer reflection geometry introduces four mechanisms a clean
F2 reflection does not.  ionocast does not split the term across
those mechanisms today; it is an aggregate that absorbs the four
contributions below into a single calibrated number, calibrated
against the Es-active subset of the WSPR per-pair basket.
Listing the contributors anyway lets a reader judge which
component would dominate if the term were ever decomposed:
patchy cloud structure (footprint mismatch between TX and RX
illumination, ${\sim}3$--$6$\,dB on a typical small-to-medium Es
patch); strong aspect sensitivity at the layer's tilted surfaces
(${\sim}2$--$5$\,dB swing depending on takeoff azimuth alignment);
polarisation scrambling through a thin gradient layer
(${\sim}1$--$3$\,dB); and the residual ITU-R Yp day-to-day
variability that Es-strength hour-to-hour swings produce on top of
the F2 day-to-day spread (${\sim}3$--$5$\,dB).  These ranges
roughly sum to the 15\,dB calibration value.  Future work may
elevate them to named per-mechanism terms; the present paper
ships the aggregate.

The D-region terms still apply --- the E layer sits above the D
layer, so an Es signal still passes the absorbing region once per
hop.  $L_{\text{hop}}$ is zero for Es (single-hop, no intermediate
ground reflections).  $L_{\text{aur}}$ is \emph{omitted} from
Eq.~\ref{eq:esmargin} entirely (the term is not in the equation),
\emph{not} because it can't physically apply --- it can; auroral D
absorption affects any signal traversing the D-region, and Es
hops cross the D-region twice (TX up, down to RX) --- but because
the geographic regions where Es is the dominant propagation mode
overlap rarely with the auroral oval.  The $f \geq \SI{14}{\MHz}$
operating range and summer-evening seasonality of strong Es push
most Es hops to midlat / equatorial corridors away from the
auroral CGM threshold.  On the rare auroral-zone summer-Es path
(Scandinavia, Alaska, Antarctic) where both mechanisms co-occur
the budget under-predicts absorption; treat ``$L_{\text{aur}} = 0$
on Es'' as a geographic simplification rather than a physical
claim.

Gating in code: \texttt{snrMarginHfEs} returns null below
\SI{14}{\MHz} regardless of $f_{\text{oEs}}$
(\texttt{src/physics/snr.js}), even though the algebraic Es MUF
gate $f \le 5 f_{\text{oEs}}$ would otherwise admit lower bands
on strong Es days ($f_{\text{oEs}} \ge \SI{4}{\MHz}$ pushes
$5 f_{\text{oEs}} \ge \SI{20}{\MHz}$, which would activate the
30\,m and 40\,m bands without the floor).  The hard 14\,MHz floor
matches amateur-radio literature: Es is essentially unobserved
below 20\,m at the 2000\,km characteristic single-hop distance
the budget assumes, so the algebraic gate alone would over-fire
on the lower HF bands.  The cutoff is intentionally
\emph{conservative}: rare Es openings on 30\,m and 40\,m do
appear in the WSPR record during peak-cycle summer evenings with
$f_{\text{oEs}} \ge \SI{8}{\MHz}$, typically on shorter-than-2000\,km
paths where the geometry makes the lower-band Es viable.  These
events are not captured by the budget; an operator who sees
30\,m WSPR activity that the model calls F2-only should treat the
verdict as ``F2 explains most of it; rare Es is possible on top
and the budget won't tell you.''  $L_{\text{low}}$ in
Eq.~\ref{eq:esmargin} is therefore numerically zero under the
current implementation (Table~\ref{tab:lowbandextra} returns 0
for $f \ge \SI{14}{\MHz}$); it is kept for symbolic symmetry with
the F2 budget so that any future relaxation of the 14\,MHz floor
automatically picks up the low-band penalty without a separate
edit.  Es variability adds $\SI{+2}{\dB}$ in quadrature to the
base $\sigma$.

\subsection{Trans-Equatorial Propagation}\label{sec:tep}

Chordal afternoon / evening mode across the magnetic equator.  The
signal couples into the equatorial F-region irregularity structure
and traverses hundreds to thousands of km without ground-bouncing,
skipping several D-region traversals.  Classical TEP produces 15\,m
/ 12\,m / 10\,m and occasionally 6\,m openings between stations whose
magnetic dip latitudes sit on opposite sides of the equator.

The TEP bonus is a smooth function of three eligibility axes plus a
hard cross-equator gate.  Hard gate (binary): both endpoints' dip
latitudes have opposite signs (i.e.\ $\phi_{\text{dip,src}} \cdot
\phi_{\text{dip,dst}} < 0$); same-hemisphere paths see no TEP at any
distance.  Once the cross-equator gate fires, three smooth ramps
modulate the bonus magnitude:
\begin{itemize}[nosep]
  \item \textbf{Frequency factor}: ramps from 0 at \SI{14}{\MHz} to
        1 at \SI{22}{\MHz}, holds at 1 across the
        $[\SI{22}{\MHz}, \SI{50}{\MHz}]$ plateau, then ramps back
        down to 0 at \SI{60}{\MHz}.  This eliminates the earlier hard
        step at \SI{20}{\MHz} that placed 17\,m at zero TEP and
        15\,m at full plateau across a 3 MHz band edge.
  \item \textbf{Local-hour factor}: ramps from 0 at $\text{LT}=16$ to 1
        at $\text{LT}=17$ at the leading edge, plateau over $[17, 23]$,
        then 1 at $\text{LT}=23$ to 0 at $\text{LT}=24$ at the
        trailing edge.
  \item \textbf{Dip-latitude factor}: each endpoint's
        $|\phi_{\text{dip}}|$ contributes a ramp $f_{\text{dip}} =
        \mathrm{clamp}((|\phi_{\text{dip}}| - 5\degree) / 5\degree,
        0, 1)$, i.e.\ zero below $|\phi_{\text{dip}}| = 5\degree$,
        rising linearly to one at $|\phi_{\text{dip}}| =
        10\degree$, then saturating.  The path-level factor is the
        product of the two endpoint ramps
        ($f_{\text{dip,src}} \cdot f_{\text{dip,dst}}$), so a path
        with one endpoint near the dip equator gradually loses TEP
        eligibility as that endpoint approaches the equator and
        loses it entirely once $|\phi_{\text{dip}}| < 5\degree$.
\end{itemize}
The bonus is the product of the plateau value and the three
factors:
\begin{equation}\label{eq:tep}
  B_{\text{TEP}} =
    B_{\text{TEP,plateau}}(F_{10.7\text{A}}) \cdot f_{\text{freq}} \cdot
    f_{\text{LT}} \cdot f_{\text{dip,src}} \cdot f_{\text{dip,dst}}
\end{equation}
where the plateau magnitude is solar-cycle-keyed via a smooth
sigmoid:
\begin{equation}\label{eq:tepplateau}
  B_{\text{TEP,plateau}}(F_{10.7\text{A}}) =
    8 + \frac{7}{1 + e^{-(F_{10.7\text{A}} - 125)/30}} \,\si{\dB}
\end{equation}
The sigmoid asymptotes to $\sim\!8\,\si{\dB}$ at solar minimum and
$\sim\!15\,\si{\dB}$ at solar maximum, with the transition centred
near $F_{10.7\text{A}} = 125$.  At quiet sun the plateau is
$\sim\!9\,\si{\dB}$; at moderate cycle ($F_{10.7\text{A}} = 120$)
it sits at $\sim\!11\,\si{\dB}$; near solar peak ($F_{10.7\text{A}}
= 180$) it reaches $\sim\!14\,\si{\dB}$.  Real TEP intensity scales
with EUV (and therefore F10.7), so the plateau follows F10.7A
rather than holding at the peak-cycle magnitude across all
conditions.

The full bonus saturates at the central
22--50 MHz / 17--23 LT / opposite-hemisphere near-equatorial-pair
window (frequency factor saturates at \SI{22}{\MHz}, both
endpoints' $|\phi_{\text{dip}}|$ above the 10\textdegree\
saturation knee).  This is a first-order approximation --- real
TEP magnitude varies with Es index, EIA crest strength, and local
evening $E \times B$ drift --- but it captures the main effect
with the smooth-ramp philosophy applied to the rest of the
budget: bands that the F2 model says are closed can in fact be
wide open across the equator for a few-hour window.

\subsection{Meteor Scatter Floor}\label{sec:ms}

6\,m and 2\,m can sustain brief burst-mode QSOs off meteor ionisation
trails, particularly in the predawn hours when Earth's leading face
sweeps up the most meteors.  ionocast lifts the VHF verdict floor
from \emph{closed} to \emph{poor} during the predawn window
($[2, 10]\,\text{h}$ local at the QTH) under either of two
conditions:
\begin{itemize}[nosep]
  \item A major IMO shower with $\text{ZHR} \geq 20$ in its
        \emph{active} or \emph{building} window per the IMO
        catalogue (the active window is $\pm 2$\,d of peak; the
        building window is the leading edge of the shower as IMO
        catalogues it, typically a few days before active begins,
        depending on shower).  The verdict note names the shower
        (e.g.\ ``Perseids'').  The
        $\text{ZHR} \geq 20$ threshold catches eight to ten
        showers per year at typical peak ZHR, including (in
        approximate ZHR order) Quadrantids ${\sim}\!120$,
        Geminids ${\sim}\!120$, Perseids ${\sim}\!100$,
        Eta Aquarids ${\sim}\!50$, Southern $\delta$~Aquariids
        ${\sim}\!25$, Orionids ${\sim}\!20$, Lyrids ${\sim}\!18$
        (just below the gate; included in some IMO catalogue
        years), Leonids ${\sim}\!15$ (outside outburst years; an
        outburst year clears the gate easily), Ursids
        ${\sim}\!10$ (below gate), and several minor regulars.
        The exact count varies year to year as IMO refines the
        catalogue.  Strong-but-sub-threshold annual showers like
        Lyrids and outside-outburst Leonids sit just below the
        gate, and the $>$100 minor showers IMO catalogues with
        $\text{ZHR} < 20$ are excluded, on the assumption that
        their incremental ionisation above the sporadic background
        isn't large enough to shift the verdict from \emph{poor}
        to anything higher.
  \item Otherwise, \emph{sporadic background} -- the
        ${\sim}5\text{--}10$~ZHR continuous meteoroid influx that
        sustains routine MS QSOs every dawn even on shower-free
        days.  The verdict note reads ``sporadic background''.
\end{itemize}
\emph{Poor} signals ``viable with effort, not sustained'' --- MS is
bursty, not continuous.  Earlier code missed the sporadic-background
case and only triggered on shower days, leaving 6\,m / 2\,m flagged
\emph{closed} during the routine MS window most of the year.  The
$[2, 10]\,\text{h}$ centre window is now flanked by 30-minute linear
ramps on each side (the activation $\text{weight} \in [0,1]$ ramps
$0 \to 1$ over $[1.5, 2.0]\,\text{h}$ LT and $1 \to 0$ over
$[10.0, 10.5]\,\text{h}$ LT).  The boolean MS-active flag fires on
any non-zero weight, so the operationally-firing window extends
30~minutes on each side of the centre band; the previous tier cliff
at the boundary has been moved to the weight=0 edges where MS
activity is genuinely below useful threshold.  This matches the
smooth-ramp philosophy applied at every other gate (auroral
$c(\phi)$, Bz, twilight $r$, near-MUF, Es, terminator $\sigma$,
forecast-Kp).

\subsection{VHF Auroral-E Scatter}\label{sec:vhfaur}

When the auroral oval is illuminated (OVATION
HP $\ge \SI{30}{\GW}$), 6\,m and 2\,m can open via E-region
scatter off the auroral electron precipitation column.
\texttt{snrMarginVhfAurora} (in \texttt{src/physics/snr.js})
returns a simple lumped-loss budget gated on HP --- the
function is a hard \texttt{return null} for HP $< 30$\,GW, so
no VHF-aurora verdict is even computed below the threshold,
which keeps the path off the band-table when auroral activity
is too low for E-scatter to be physically supported --- with an
auroral-MUF model that rises with hemispheric power
($\text{MUF}_{\text{aur}} = 50 + (\text{HP}_{\text{GW}} - 30)$\,MHz,
so 50\,MHz at the threshold and ${\sim}\!70$\,MHz at
HP $= \SI{50}{\GW}$, with no upper cap in code so an extreme
HP\,${\sim}\,200$\,GW would notionally lift $\text{MUF}_{\text{aur}}$
to $\sim 220$\,MHz; in practice the gate at HP $\ge 30$\,GW
rarely exceeds $\sim 100$\,GW even during G5 events so the
unbounded form has not been observed to misbehave, but a hard
cap at e.g.\ $\sim 100$\,MHz is candidate hardening tracked in
\S\ref{sec:limits}):
\begin{equation}\label{eq:vhfaurmargin}
  M_{\text{aur,VHF}} = P_{\text{tx}} + G_{\text{ant}}
    - L_{\text{fs}} - L_{\text{MUF}}
    - L_{\text{iono,aur}} - N - S_{\text{req}}
\end{equation}
with $L_{\text{iono,aur}} = \SI{25}{\dB}$
(\texttt{L\_IONO\_AUR\_DB} in \texttt{src/physics/loss.js}).
This is an \emph{empirical lump} (no single-source citation):
the value collapses several mechanisms operators report on
auroral-E paths --- polarisation scrambling on the auroral
E-layer column ($\sim\!4$--$\SI{8}{\dB}$ swing), aspect-
sensitivity loss against operator-side antenna-pattern
misalignment ($\sim\!3$--$\SI{6}{\dB}$ on a typical
narrow-beam VHF setup), and the residual ionospheric
population variability that swings hour-to-hour during
active-aurora windows ($\sim\!4$--$\SI{8}{\dB}$) --- which
sum to a 25\,dB midpoint operationally.  The three
contributing ranges are themselves anecdotal in the VHF
contest / DXer literature rather than measured against a
calibrated reference; the 25\,dB number is therefore
\emph{calibration-pending} (needs author confirmation against
6\,m / 2\,m auroral logs) and is held as a defensible
default until the next harness pass.  $L_{\text{aur}}$
(\S\ref{sec:laur}) does not appear in
Eq.~\ref{eq:vhfaurmargin} because the path is reading
\emph{on} the auroral E-region scatter rather than passing
through it as absorption; the same physics that closes HF over
auroral hops opens 6\,m and 2\,m scatter on the same hops.

\subsection{F2 Scatter Recovery}\label{sec:scatter}

When the band is nominally above MUF the F2 budget says
\emph{closed}, but real F-region irregularities (TIDs, plasma blobs,
gradient-driven instabilities) often keep the path partly viable via
scatter modes that do not require strict over-MUF compliance at
every hop.  ionocast reads inverse-square-distance-weighted foF2 from
all GIRO digisondes within \SI{3000}{\km} of each F2 reflection point
along the great circle, falling back to climatology
(Section~\ref{sec:climo}) at hops out of station range, and computes
the per-hop foF2 standard deviation $\sigma_{\!f}$.  The
inverse-square weighting biases the spread estimate toward the
nearest station: on a hop with one station at \SI{200}{\km} and
five stations at \SI{1500}{\km}, the near station carries
${\sim}80\%$ of the total weight, so $\sigma_{\!f}$ collapses
toward the near reading and the variance signal that
$B_{\text{scatter}}$ is supposed to detect can be artificially
narrowed.  This is a known statistical bias rather than a coding
mistake --- the weighted-variance formula is well-defined and the
resulting estimator is consistent, just not the unbiased
equal-weight variance.  The choice trades a slightly under-detected
$B_{\text{scatter}}$ on station-asymmetric hops against larger
variance on equal-weight estimates from the long-tail stations,
which empirically matter less to the path-averaged spread.  A
distance-window-uniform weighting is a candidate alternative;
re-fitting $w_{\text{sc}}$ under that scheme is gated on the
per-pair WSPR ground-truth mode (Section~\ref{sec:perpath}).
The scatter bonus fires only when all three gating conditions are
met simultaneously: $\geq 2$ GIRO stations contribute to
$\sigma_{\!f}$, the path takes $\geq 2$ hops, and
$f / \text{MUF} > 1.0$ (below MUF the standard model already says
``open'').  When fired, an additive scatter bonus is applied to
the SNR margin:
\begin{equation}\label{eq:scatterbonus}
  B_{\text{scatter}} =
    w_{\text{sc}}\cdot
    \min\!\bigl(1,\,\sigma_{\!f}/\SI{1.5}{\MHz}\bigr)\cdot
    \min\!\bigl(2,\,f/\text{MUF} - 1\bigr)\cdot
    \SI{5}{\dB}
\end{equation}
with $w_{\text{sc}} = 1.5$ from the joint
coordinate descent of \S\ref{sec:calibrate} (the harness's
unrestricted sweep keeps lifting $w_{\text{sc}}$ toward 4 against
the global ``alive somewhere'' metric, but that range exploits a
known structural ceiling of the metric rather than a real
propagation effect; the \SI{15}{\dB}-cap physical edge $1.5$
ships).  The foF2-variance term
saturates at \SI{1.5}{\MHz} spread (substantial gradients across the
path); the above-MUF excess saturates at $f/\text{MUF} = 3$ (wildly
over-MUF paths get no recovery).  At full saturation
$w_{\text{sc}} = 1.5$ contributes up to \SI{15}{\dB} of recovery,
sitting inside the operator-experience F2-scatter range
(10--25\,dB, consistent with K9LA's
propagation-tutorial archive~\cite{k9la} for the upper HF
above-MUF regime, where formal published F2-scatter
loss-magnitude measurements are sparse) and
matching the upper-band lift the model needed to track 30~d of
WSPR aggregate openings on 17--10\,m.  The two saturating
$\min(\cdot)$ terms are multiplied independently, but
$\sigma_{f}$ and $f/\text{MUF}$ are not strictly independent in
practice: paths with high $\sigma_{f}$ (large gradients across
the path) tend to have at least one hop near or above MUF
(observed empirically in the per-(path, band, hour) cells of
\texttt{harness.report.json}, not separately quantified or
literature-cited; treat as a calibration-side observation
rather than a published correlation).  The independent
saturation form slightly inflates the joint upper-bound
probability; the effect is small ($\leq \SI{1}{\dB}$ on the
$\SI{15}{\dB}$ ceiling) and does not change the calibrated
$w_{\text{sc}}$.

The weight was set by joint coordinate descent over
$\{L_{\text{iono}},\,D,\,w_{\text{sc}},\,w_{\text{NVIS-tail}},
w_{\text{Es-prim}},\,\sigma\text{-scale},\,\text{fusion-flag}\}$
from three deterministic starting configurations
(Appendix~\ref{app:repro}, ``baseline'', ``fusion-up'',
``modes-on'') across 35 reference paths
(Section~\ref{sec:calibrate}).  All seeds converged to the same
basin: only $w_{\text{sc}}$ produced measurable lift; an
unrestricted sweep keeps improving as $w_{\text{sc}}\!\to\!4$, but
that range exploits the WSPR-global-vs-path-specific structural
mismatch in the harness rather than a real propagation effect, so
the shipped weight is held at the \SI{15}{\dB}-cap physical edge.
The result is that the
calibrated $w_{\text{sc}} = 1.5$ has been validated against a
metric whose calibrators do not fully trust beyond a certain point.
Closing this loop properly requires re-running the sweep against
the per-pair WSPR ground truth (Section~\ref{sec:perpath}) once
that mode is wired into \texttt{replayMarginFromCell}'s alternative-mode
bonuses; the new optimum, whatever it is, can then be reported
against a metric that the calibrators \emph{do} trust at all weights
in its range.  Until that closure, treat $w_{\text{sc}} = 1.5$ as a
defensible-but-not-final value tied to a metric with a known
ceiling.

\paragraph{Per-hop fusion (deferred).}  The same per-hop foF2
sampling is also wired as an optional second-opinion source for
the MUF consensus blend (\S\ref{sec:muf}, where the
\texttt{fusion-flag} is introduced as one of the calibration
parameters of Appendix~\ref{app:repro}), behind the
\texttt{FUSION\_PRIMARY\_MUF} flag.  In the same calibration sweep
the flag stayed off: single-midpoint fusion did not beat the
kc2g\,$\leftrightarrow$\,climatology blend in either Brier or
accuracy, and per-hop minimum fusion regressed binary accuracy by
${\sim}1$\,pp because the rebuilt climatology
(Section~\ref{sec:climo}) tightened the residual it would have to
beat.  A future calibration that retunes the budget jointly with
fusion enabled may flip this; the wiring is intentionally preserved.

\subsection{Tropospheric Ducting}\label{sec:tropo}

The four modes above all live inside the SNR budget: each one
contributes a margin or bonus that competes with the F2 verdict for
the same path.  Tropospheric ducting is structurally different. It
is a VHF-only line-of-sight extension whose viability depends on the
shape of the lowest \SI{1}{\km} of the troposphere on the great
circle between TX and RX, not on any quantity the global F2 model
already evaluates. ionocast surfaces it as a measured per-station
panel (Section~\ref{sec:output}) rather than folding it into the
budget; the rest of this subsection documents the physics, the
classification, and how the panel is wired.

\paragraph{Refractivity and the duct condition.} Atmospheric
refractivity $N$, in N-units, is computed from radiosonde
temperature, pressure, and water-vapour pressure
(ITU-R P.453~\cite{itup453}):
\begin{equation}\label{eq:refractivity}
  N = \frac{77.6}{T} \left( P + \frac{4810\,e}{T} \right),
\end{equation}
with $T$ in kelvin, $P$ and $e$ in millibar. Water-vapour pressure
$e$ is obtained from the dew point via the Magnus saturation
formula
\begin{equation}\label{eq:esat}
  e_{\text{sat}}(t_C) = 6.112 \cdot
    \exp\!\left(\frac{17.62\,t_C}{243.12 + t_C}\right),
\end{equation}
evaluated at the dew-point temperature (so that
$e = e_{\text{sat}}(t_{\text{dew}})$ is the actual partial
pressure).  The constants $(17.62, 243.12)$ are the
Sonntag~1990~\cite{sonntag1990} coefficients for saturation
\emph{over liquid water} (recommended by WMO~2008 over the
Alduchov--Eskridge alternative, which gives slightly different
values $a=17.625$, $b=243.04$).  Below freezing, saturation over
ice uses different
constants $(22.46, 272.62)$ producing a ${\sim}10\%$ lower
$e_{\text{sat}}$ at $t_C = -20\,\degree\text{C}$.  ionocast applies
the liquid-water form across all temperatures, which is correct
for the warm-water duct cases where ducting is operationally
relevant (Mediterranean, Sea of Japan, California coast) but
slightly over-estimates $e$ on continental winter soundings where
the surface dew point is below freezing.  The resulting bias on
$N$ is ${\le}\!2$\,N-units across the operationally relevant range
$t_C \in [-30, +35]\,\degree\text{C}$ (worst case
${\sim}\!2$\,N-units near $t_C = -20\,\degree\text{C}$ where the
ice-vs-liquid divergence peaks; below $-30\,\degree\text{C}$ the
absolute moisture content is so low that the absolute $e$ error
shrinks even though the percentage error grows), and the regime
classification (standard / super-refractive / ducting) is robust
to it; refining this further would require a piecewise formula
gated on $t_C \gtrless 0\,\degree\text{C}$, which is candidate
cleanup tracked in Section~\ref{sec:limits}.  Modified refractivity $M$ adds a curvature correction
for the Earth-radius bend that makes a horizontally-stratified
duct look horizontal in $M$-coordinates,
\begin{equation}\label{eq:modM}
  M(z) = N(z) + 0.157\,z,
\end{equation}
with $z$ in metres.  A horizontal duct exists wherever
$\mathrm{d}M/\mathrm{d}z < 0$; equivalently, in the $N$-coordinate
that ionocast actually computes,
\begin{equation}\label{eq:ductthreshold}
  \frac{\mathrm{d}N}{\mathrm{d}h}\bigg|_{0\to\SI{1}{\km}}
    < -157~\text{N-units}/\si{\km}.
\end{equation}
The $-157$ value is the modified-refractivity coefficient
$0.157\,\text{N/m}$ rescaled to N-units per kilometre. It is not a
loss budget --- a duct either exists or it does not --- so the
treatment here is classification, not a margin contribution.

\paragraph{Bean \& Dutton~\cite{beandutton1968} classification.}
The lowest-\SI{1}{\km} gradient is bucketed into three regimes
(thresholds and labels follow Bean \& Dutton; ITU-R
P.453~\cite{itup453} provides the underlying refractivity
formulation but tabulates only $\le -100~\text{N-units}/\si{\km}$
gradients without printing the $-79$ / $-157$ classification
boundaries):
\begin{itemize}[nosep]
  \item $\mathrm{d}N/\mathrm{d}h > -79~\text{N-units}/\si{\km}$:
        \emph{standard atmosphere}.  Radio rays bend gently toward
        Earth (the textbook $\frac{4}{3}$ effective Earth radius);
        no range extension beyond the geometric horizon.
  \item $-157 \le \mathrm{d}N/\mathrm{d}h \le -79~\text{N-units}/\si{\km}$:
        \emph{super-refractive}.  The radio horizon extends past
        the geometric horizon; tropospheric scatter paths run
        noticeably above their fair-weather budget.  Common
        precursor to ducting.
  \item $\mathrm{d}N/\mathrm{d}h < -157~\text{N-units}/\si{\km}$:
        \emph{ducting}.  A surface-based or elevated atmospheric
        duct exists; VHF/UHF signals trapped in the duct can reach
        hundreds to thousands of kilometres independent of the
        ionosphere.  Operationally this opens 6\,m and 2\,m DX
        between stations sharing the duct, particularly across
        warm-water surfaces (Mediterranean, Sea of Japan,
        California coast) and under summer subsidence inversions.
\end{itemize}

\paragraph{Implementation.} The $\mathrm{d}N/\mathrm{d}h$ panel is
built from a 24-station basket of UWyo radiosonde sites with global
coverage (Europe~$\times 10$, North America~$\times 7$,
SE~Asia~$\times 3$, Africa~$\times 2$, South~America~$\times 1$,
Antarctica~$\times 1$; \texttt{SONDE\_STATIONS} in
\texttt{functions/\_handlers/refractivity.js}).
The Cloudflare worker (\texttt{functions/\_handlers/tropo.js}) fans
out over the basket in parallel for the most recent 00\,Z or 12\,Z
slot (with a \SI{12}{s} per-fetch timeout to bound the worst case),
parses each PRE-block sounding, and for each station:
\begin{enumerate}[nosep]
  \item Locates the surface sample (lowest row) and the sample
        nearest $z_{\text{surface}} + \SI{1}{\km}$ that is at least
        \SI{100}{m} above the surface.
  \item Computes $N_{\text{surface}}$ and $N_{\text{upper}}$ via
        Eq.~\ref{eq:refractivity} from temperature, pressure, and
        dew-point at each level.
  \item Returns $\mathrm{d}N/\mathrm{d}h = (N_{\text{upper}} -
        N_{\text{surface}})/\Delta z_{\text{km}}$ and the regime
        classification.
\end{enumerate}
Stations whose 00\,Z / 12\,Z release is missing from the upstream
archive return \texttt{ok: false} with reason
\texttt{"no data"}, which the ducting panel renders as a placeholder
row.  The nearest station with valid data is also surfaced as a
single $\mathrm{d}N/\mathrm{d}h$ cell on the VHF band-table,
sharing one cell across 6\,m and 2\,m because the gradient is
per-station, not per-band; the cell's tier color uses the same
classification (good $=$ ducting, warn $=$ super-refractive, no
color $=$ standard) so the band-table verdict and the per-station
panel never disagree on what the local atmosphere is doing.

\paragraph{Why this is informational, not a budget term.} A duct is
only useful if both endpoints sit inside it.  The radiosonde
network samples vertical structure at fixed sites every
\SI{12}{\hour}, which is enough to tell an operator at one of
those sites whether the local atmosphere is currently trapping
signals, but it is not enough to claim a specific TX-to-RX path is
ducted: marine-layer ducts can hug a coastline for hundreds of
kilometres while inland soundings show standard atmosphere only
\SI{50}{\km} away, and elevated subsidence ducts can sit several
hundred metres above a surface that a radiosonde column doesn't
cleanly resolve.  Folding the gradient into a path-level SNR bonus
would over-promise the model's spatial coverage; surfacing the
per-station classification lets the operator make the geographic
inference where the radiosonde basket is dense (Western Europe,
the contiguous US) and lets it stay silent where the basket is
thin (most of Africa, Central Asia, the open ocean).
The companion gridded surface (\S\ref{sec:tropomap}) fills the
spatial coverage gap with a NOAA GFS-derived per-cell index that
the radiosonde basket is then used to calibrate.

\subsection{Global Tropospheric Ducting Heatmap}\label{sec:tropomap}

The 24-station radiosonde basket of \S\ref{sec:tropo} answers
"is the local atmosphere ducting?" wherever a sounding lands within
the user's region of interest, but most of the planet sits hundreds
of kilometres from the nearest radiosonde site. To give operators a
global at-a-glance view, ionocast also ingests the NOAA Global
Forecast System (GFS), computes a per-cell composite ducting index
on a regular lat/lon grid, and renders a Hepburn-class heatmap
embedded in the VHF panel.  The radiosonde-per-station table of
\S\ref{sec:tropo} is retained as the calibration anchor: its P.453
classifications are the falsifiable ground truth against which the
gridded index is tuned.

\paragraph{Forecast input.} GFS at 0.25\textdegree{} is the densest
public NWP product covering the global troposphere with a refresh
cadence faster than 6\,h.  ionocast pulls 13 pressure levels
between 1000 and 500\,hPa
(1000 / 975 / 950 / 925 / 900 at 25\,hPa spacing, then
850 / 800 / 750 / 700 / 650 / 600 / 550 / 500 at 50\,hPa, the
densest set NOMADS publishes on its \texttt{pgrb2.0p25}
endpoint~\cite{noaagfs}) plus the surface bundle: 2\,m
temperature, 2\,m dew point, skin temperature, land-sea mask,
surface pressure, model orography, planetary-boundary-layer height,
and 10\,m wind components.  Each cycle is filtered server-side by
the NOMADS subset URL down to ${\sim}21$\,MB (vs ${\sim}500$\,MB
for the unfiltered 0.25\textdegree{} file).  The 4\,$\times$ daily
refresh runs ${\sim}5$\,h after each cycle (00 / 06 / 12 / 18\,Z)
via a GitHub Actions workflow, packs the result to a compact binary
(${\sim}2.1$\,MB row-major \texttt{Float32}, ${\sim}10\times$
smaller than the JSON form), and uploads to Cloudflare R2 for
browser delivery.

\paragraph{Index composition.} The per-cell index $\mathcal{T}$ is
a seven-term composite, height-weighted in the lowest 2.5\,km AGL
(weight 1.0 below 1.5\,km, 0.5 between 1.5 and 2.5\,km, zero above
since amateur VHF/UHF cannot couple meaningfully into upper
tropospheric ducts) and attenuated by surface mixing:
\begin{equation}\label{eq:tropoindex}
  \mathcal{T} = (1 - w_p) \cdot \!\left[\,
    5\,\Delta_{\text{duct}}
    + S_{\text{sr}}
    + S_{\ell}
    + S_{\text{surf}}
    + S_{\text{marine}}
    + S_{\text{HPBL}}
    + S_{\text{evap}}
  \,\right],
\end{equation}
where the terms, in M-units, are:
\begin{itemize}[nosep]
  \item $\Delta_{\text{duct}}$, the ITU-R P.834~\cite{itup834}
        canonical duct intensity $M_{\text{top}} - M_{\text{min}}$
        per trapping layer, taken as the maximum across all
        trapping layers in the column.  Weighted $5\times$ so a
        30-unit duct contributes 150 to $\mathcal{T}$ and saturates
        the renderer's 200-unit colormap.
  \item $S_{\text{sr}}$, the super-refractive sum: per-layer
        $(-79 - \mathrm{d}N/\mathrm{d}h)$ accumulated over every
        layer whose gradient sits in the P.453 super-refractive
        band $[-157, -79]$\,N-units$/$\si{\km}.  This term carries
        the broad marine baseline (subsidence inversions over the
        subtropical highs, sub-textbook stratification) that the
        binary "is there a duct" check misses.
  \item $S_{\ell}$, the per-layer temperature-inversion sum,
        triggered when a layer's $\mathrm{d}T$ exceeds the
        standard-lapse value by ${\ge}\!3$\,K.
  \item $S_{\text{surf}}$, the surface inversion bonus, comparing
        $T_{2m}$ to the lapse-adjusted temperature at the first
        pressure level $\ge\!200$\,m above ground; threshold
        1\,K, capped at 100.
  \item $S_{\text{marine}}$, the marine inversion bonus, fires
        when SST exceeds 2\,m air temperature by $\ge\!1.5$\,K
        over an ocean cell (gated by GFS land-sea mask), capturing
        Benguela / California-current cold-tongue and Mediterranean
        warm-water stratification.  Capped at 60.
  \item $S_{\text{HPBL}}$, a capping-inversion bonus that activates
        when GFS's planetary-boundary-layer height $z_{\text{HPBL}}
        < 1500$\,m \emph{and} another inversion term has already
        fired.  This is the textbook capping-inversion duct setup:
        a shallow well-mixed BL trapped under a strong inversion.
        Up to $+25$.
  \item $S_{\text{evap}}$, the saturation-deficit evaporation duct
        bonus.  Over open water the air immediately at the sea
        surface is saturated by definition; the deficit
        $e_{\text{sat}}(T_{\text{sea}}) - e(T_{2m,\text{dew}})$
        scales the strength of the resulting near-surface
        refractivity gradient.  Replaces an earlier
        column-precipitable-water proxy that fired on bulk
        humidity rather than the surface gradient that actually
        drives evap-duct height.  Capped at 30.
  \item $w_p \in [0, 0.6]$, the wind-mixing penalty: 10\,m wind
        speeds above 8\,m/s degrade the laminar boundary-layer
        structure that ducts depend on, attenuating $\mathcal{T}$
        by 10\% per m/s above 8\,m/s up to a 60\% reduction at
        14\,m/s.
\end{itemize}
A super-adiabatic lapse anywhere in the lowest 2\,km marks the
column as convectively-mixed and divides $S_{\ell}$ by 5 and
$\Delta_{\text{duct}}$ by 2; surface, marine, and evaporation
bonuses are unaffected because they measure surface-anchored
quantities the convective check (a layer-gradient probe) cannot
invalidate.

\paragraph{Pressure-level density and ECMWF comparison.} A
side-by-side calibration against the same 24-station P.453 panel
of \S\ref{sec:tropo} makes the case for GFS over ECMWF Open Data on
this specific application: the ECMWF Open Data feed publishes only
five pressure levels between 1000 and 500\,hPa
(1000 / 925 / 850 / 700 / 500), giving 75 to 150\,hPa gaps that
average across the 50 to 300\,m thick layers where most operational
ducts physically live.  Run with the same index formula, the ECMWF
feed scored 0\% recall on observed super-refractive layers; the
same harness against the GFS feed scored 80\%.  ECMWF retains a
synoptic-skill edge on multi-day large-system forecasts that is
not relevant for current-state ducting.  The full ECMWF-IFS
137-model-level product would resolve sub-100\,m structure but is
not part of the free Open Data feed.

\paragraph{Calibration.} The calibration harness
(\texttt{src/tropo/calibrate.mjs}) fetches the same 24-station
UWyo basket, scans each sounding for its strongest sub-3\,km
gradient (rather than the bulk-1\,km gradient \S\ref{sec:tropo}
reports per-station, which averages over thin surface inversions),
classifies it per Bean \& Dutton~\cite{beandutton1968}, samples
$\mathcal{T}$ at the station coordinates from the most recent grid,
and reports three-class agreement plus per-class precision and
recall.  Across 13 amateur-relevant sondes per cycle (those whose
strongest gradient sits at or below 2.5\,km AGL, matching the
height-weighting cap), the optimal cuts come out near
\begin{equation}\label{eq:tropocuts}
  \mathcal{T} < 5: \text{standard}; \quad
  5 \le \mathcal{T} \le 90: \text{super-refractive}; \quad
  \mathcal{T} > 90: \text{ducting},
\end{equation}
with 100\% precision in the ducting band (when the index reads
$\ge\!90$\,M-units a sonde at the same coordinates agrees the
column is ducting) and ${\sim}17$\% recall.  The recall ceiling is
GFS's vertical resolution rather than the formula: surface and
shallow ducts in 50 to 300\,m thick layers fall between the model's
25 to 50\,hPa pressure-level spacing and read as $\mathcal{T} = 0$.
That is a known and bounded data-source limit; the operational
implication is that bright cells on the heatmap are reliable but
the absence of color does not rule out ducting.

\paragraph{Rendering.} The browser fetches the packed binary from
Cloudflare R2 via a same-origin redirect, decodes the row-major
\texttt{Float32} arrays, applies a separable
$[1,4,6,4,1]/16$ binomial Gaussian to soften per-cell jitter
(matching the spatial smoothing every operational forecast product
applies before contouring), and runs marching-squares contour
extraction~\cite{d3contour} to produce closed polygon boundaries
at 15 thresholds spaced
\begin{equation}
  \mathcal{T}_i = \mathcal{T}_{\min} +
    \left(\frac{i}{N}\right)^{\!1/0.7}\!
    (\mathcal{T}_{\max} - \mathcal{T}_{\min}),
  \qquad i = 0, 1, \dots, 14
\end{equation}
with $\mathcal{T}_{\min} = 20$ and $\mathcal{T}_{\max} = 200$.  The
$1/0.7$ exponent stretches the lower bands so the weak-but-real
subtropical-marine baseline gets two or three colored bands rather
than a single navy stripe; the strongest 1\% of the globe saturates
at the deepest red without compressing the more common 30 to 100
range into purple alone.  MapLibre paints each band as a fill
layer with a hard color edge, identical to the
meteorological-forecast contour-fill look used by William R.\
Hepburn's worldwide tropo forecasts~\cite{hepburntropo} and Met
Office surface charts.  Black isolines from a marching-squares
overlay run on the source grid (unsmoothed) and trace the data
boundaries on top of the band fills.

\paragraph{Why this is not folded into the SNR budget.} The heatmap
is a global field of point classifications, not path-level loss
data.  A bright cell over the Mediterranean tells operators on
both Italian and Spanish coasts that their local atmosphere is
trapping VHF; whether a Genoa-Barcelona path is itself ducted
depends on \emph{both} endpoints sitting inside the same duct
\emph{and} the great-circle staying inside the duct between them,
which the gridded forecast cannot adjudicate without ray-tracing
through the M-profile along the path.  Surfacing the field as a
heatmap lets operators make that geographic inference at a glance
the way Hepburn's maps have been used for decades, while keeping
the SNR-budget side of ionocast honest about what a single forecast
column can support.

% ═════════════════════════════════════════════════════════════════════
\section{Ensemble Blend and Self-Calibration}\label{sec:ensemble}

\subsection{Verdict Margin}

The verdict margin is the physics-budget margin
$M_{\text{physics}}$ (Eq.~\ref{eq:budget}) plus the per-hop and
alternative-mode bonuses, with no further blending or bias
correction.  Gray-line is additive (it describes a distinct D-region
attenuation drop at the terminator), but TEP and scatter recovery
both describe F-region irregularity-driven recovery and so are
\emph{combined by maximum} rather than summed --- on a TEP-eligible
15\,m path in the 17--23 LT window above MUF, both fire on the same
physical channel and stacking them double-counts the recovery:
\begin{equation}\label{eq:blend}
  M_{\text{verdict}} = M_{\text{physics}} + B_{\text{grayline}}
                     + \max\!\bigl(B_{\text{TEP}},\, B_{\text{scatter}}\bigr)
\end{equation}

\paragraph{$\sigma$ tracks $M_{\text{physics}}$, not
$M_{\text{verdict}}$ -- caveat.}  The reliability formula
$R(M, \sigma)$ of Eq.~\ref{eq:reliability} is evaluated with
$M = M_{\text{verdict}}$ (the bonus-augmented margin) but $\sigma$
is computed from $M_{\text{physics}}$ inputs (per-band $\sigma_g$
plus the situational widenings of Section~\ref{sec:tier-confidence}).
When the bonuses fire, the bonus terms themselves carry uncertainty
that is not folded into $\sigma$: $B_{\text{TEP}} = 15\,\si{\dB}
\pm \SI{5}{\dB}$ at the published-experience plateau, and
$B_{\text{scatter}}$ is calibrated through
$w_{\text{sc}}$ but the $\pm$ on the underlying spread
($\sigma_{f_{\text{oF2}}}$ measurement noise, $f/\text{MUF}$
position) is non-trivial.  A strictly correct treatment would add
the bonus-term variances in quadrature to $\sigma$ when bonuses
are active; under a conservative ${\sim}\SI{5}{\dB}$ bonus-term
uncertainty the correction is $\sqrt{\sigma_g^2 + 25} - \sigma_g$,
which is $+\SI{1.0}{\dB}$ on a $\sigma_g = \SI{12}{\dB}$ upper-band
($\sqrt{169} - 12 = 1.0$) and $+\SI{1.4}{\dB}$ on a
$\sigma_g = \SI{8}{\dB}$ low-band ($\sqrt{89} - 8 \approx 1.43$).
The current model omits this widening: bonus-active verdicts on
both upper HF and the low/mid bands read slightly more confident
than they should, with the under-counting at ${\sim}1$--$1.4$\,dB
of $\sigma$ depending on band.  At the operationally most
relevant case --- 15\,m near-MUF with a moderate-cycle TEP firing
at $B_{\text{TEP}} \approx \SI{13}{\dB}$, $\sigma_g(\text{15\,m}) =
\SI{10}{\dB}$ --- the same $\sqrt{\sigma_g^2 + 25} - \sigma_g$
correction gives $\sqrt{125} - 10 \approx \SI{1.18}{\dB}$ of
under-counted $\sigma$, shifting the $z$-score by
${\sim}\!10\,\%$ at typical near-MUF margins (e.g.\ $z = 1.0$
nominal becomes $z = 1.0 \cdot 10/11.18 \approx 0.89$ once
corrected, dropping the predicted reliability from
$\Phi(1.0) \approx 84\,\%$ to $\Phi(0.89) \approx 81\,\%$).  The omission is
tracked in Section~\ref{sec:limits}; folding it in requires
calibrating per-bonus uncertainty estimates against the per-pair
WSPR truth (Section~\ref{sec:perpath}), which is gated on that
mode's wiring.

The TEP plateau saturates at \SI{15}{\dB} only at solar maximum
($F_{10.7\text{A}} \to \infty$); at moderate cycle ($F_{10.7\text{A}}
= 120$) the plateau sits at $\sim\!11.2$\,dB and at quiet sun
($F_{10.7\text{A}} = 70$) at $\sim\!9.0$\,dB
(Eq.~\ref{eq:tepplateau}: $8 + 7/(1 + e^{55/30}) = 8 + 0.97 \approx 8.97$).  The scatter upper bound
(\SI{15}{\dB} at $w_{\text{sc}} = 1.5$ when both
$\sigma_{f_{\text{oF2}}} = \SI{1.5}{\MHz}$ and $f / \text{MUF} \ge
3$ are simultaneously hit, see Section~\ref{sec:scatter}) shares
the asymptotic numerical ceiling with the solar-max TEP plateau,
but in practice the scatter term rarely saturates: a typical
TEP-eligible 15\,m
cell at 17\,LT with $f / \text{MUF} = 1.3$ and
$\sigma_{f_{\text{oF2}}} = \SI{0.5}{\MHz}$ produces
$B_{\text{scatter}} = w_{\text{sc}} \cdot \min(1, 0.5/1.5) \cdot
\min(2, 1.3 - 1) \cdot 5
= 1.5 \cdot 0.333 \cdot 0.3 \cdot 5
\approx \SI{0.75}{\dB}$, more than an order of magnitude below
the moderate-cycle TEP plateau (${\sim}11$\,dB at
$F_{10.7\text{A}} = 120$).  In a TEP-active
cell, $\max(B_{\text{TEP}},\,B_{\text{scatter}})$ therefore
predominantly selects $B_{\text{TEP}}$; the value of the $\max$
versus a sum is to prevent the small-but-non-zero $B_{\text{scatter}}$
from stacking on top of the calibrated TEP plateau.  The
$\max(\cdot)$ choice is a policy made on physics grounds rather
than an empirical retune: the calibration harness operates on the
climatology code path and does not compute either bonus at replay
time, so it cannot directly distinguish $B_{\text{TEP}} +
B_{\text{scatter}}$ from $\max(B_{\text{TEP}},\,B_{\text{scatter}})$.
Per-cell attribution of which mechanism dominates on TEP-eligible
15\,m paths (EU--Brazil, NA--Brazil, NA--Argentina, EU--Argentina,
EU--South Africa) is deferred to a future harness extension that
incorporates the alternative-mode bonuses into
\texttt{replayMarginFromCell}, with VOACAP cross-checks against
the 7-path fixture set in \texttt{scripts/tests/voacap.mjs}
(regenerated by the heavy \texttt{voacap-fixtures.mjs} suite).
Pure physics drives the verdict; the N0NBH-style
\texttt{heuristicTier} function in \texttt{src/physics/physics.js}
is retained as an independent reference predictor (and is what
\texttt{tune-blend.mjs} exercises as a regression canary) but
does not enter the verdict path.  Tier verdicts are defined by
fixed dB cuts (+18 / +6 / -5 / -14, Section~\ref{sec:tiers}) and
do not scale with sigma; the per-band absolute margin bias is
therefore meaningful as a fixed reference, but the harness
deliberately does not fit to it because that would couple the
verdict labels to the WSPR-truth artefact rather than the
underlying physics.

\subsection{Offline Constant Calibration}\label{sec:calibrate}

\paragraph{Two layers, one set of constants.}  The ionocast pipeline
runs in two distinct contexts.  The \emph{online} layer is the
browser at runtime: it fetches live data feeds (SWPC, kc2g, GIRO,
GFZ, Kyoto, DONKI, wspr.live), evaluates the physics in
\texttt{src/physics/}, and emits the band-table verdict.  The
\emph{offline} layer is \texttt{scripts/harness.mjs} on a
developer workstation: it imports the same \texttt{src/physics/}
code (no inline copy that can drift), replays the budget over the
35-path basket against 30~d of WSPR aggregates, and writes the
calibration baselines (\texttt{harness.baseline.json},
\texttt{spot-baselines.mjs}) that the online layer ships with.
The two layers exchange a single committed-artefact bundle: the
calibrated constants in \texttt{src/constants.js} and the WSPR
spot baselines in \texttt{src/data/spot-baselines.mjs}, both
written by the offline harness and shipped with the online
build.  Everything else
(downstream live data, the ClickHouse query, the per-(path, band)
baselines used for regression detection) lives entirely on one
side or the other.  A formal data-flow diagram is candidate
documentation work tracked in Section~\ref{sec:limits}; until
then this paragraph is the explicit description.

Several of the budget's empirical constants --- the lumped ionospheric
loss $L_{\text{iono}}$, the per-extra-hop defocusing constant $D$
(Section~\ref{sec:lhop}), and the F2-scatter weight $w_{\text{sc}}$
(Section~\ref{sec:scatter}) --- were originally set from published
ITU-R decompositions and ham-operator experience rather than measured
in place. The per-band $\sigma$ table (Table~\ref{tab:bandsigma}) is
\emph{not} fitted by the harness: $\sigma$ is anchored to the
ITU-R P.533 day-to-day spread (literature range 6--12\,dB; ionocast
picks 8--12\,dB across the per-band table, see
Table~\ref{tab:bandsigma}) and held fixed. The harness scores binary accuracy and Brier against WSPR but
does not adjust $\sigma$, since under the P.842 reliability framework
the tier boundary scales with whatever $\sigma$ is set, and a
machine-fitted $\sigma$ would chase the WSPR-truth artefact rather
than the underlying physics spread.

The current calibration harness is \texttt{scripts/harness.mjs}.
It imports the production physics directly from
\texttt{src/physics/} (no inline copy that can drift), replays the
SNR budget over a
35-path basket (Appendix~\ref{app:paths}) spanning 1--5 hops and
every major continent pair plus polar / transequatorial / NVIS-tail
geometries against
30~d of WSPR hourly spot counts (\texttt{wspr.live}, ClickHouse),
aligned hour-by-hour with the SWPC $K_p$ history, the
F\textsubscript{10.7} 81-day mean, and pre-fetched 30~d GIRO
foF2 / foEs histories from 27 digisonde stations
(Appendix~\ref{app:giro}) spanning Europe, North America, the South
Atlantic, Australia, the Pacific, and an expanded equatorial-belt
set (Ascension and Jicamarca for the Atlantic / Americas
equatorial-trough anchor pair --- Jicamarca on the dip equator,
Ascension at dip lat ${\sim}\!-3\degree$ also deep in the trough
region; Townsville and Niue for the Pacific southern crest;
BVJ03 Boa Vista at $+12\degree$ for the lone northern-crest
sample).  The Pacific / Australia block (Townsville, Niue, Perth,
Canberra, Hobart) is included to cover the South Pacific, and the
multi-snapshot subcommand \texttt{scripts/harness.mjs t1} (emitting
\texttt{scripts/data/.cache/t1-snapshots.jsonl}) is the per-station
bias diagnostic used when adding stations.  The fetcher chunks
queries 3-at-a-time with retry-on-503 because DIDB rate-limits
parallel fetches; coordinates are verified against the kc2g
station registry by \texttt{scripts/harness.mjs verify}.  Per-(path,
timestamp) cells precompute band-independent state once
(climatology, geometry, solar zenith) and the ${\sim}25{,}000$
unique cells are reused across all 10 HF bands per scoring call,
keeping a full sweep cycle well under \SI{1}{s}.

For each candidate config the harness computes the Brier loss of the
predicted \emph{open} probability against a binary ``alive'' floor on
the WSPR aggregate.  This is a sanity check on the model's binary
verdict, not a calibration target: WSPR-spot activity and link
viability for the configured station are correlated but not
identical, so binary accuracy is reported with that caveat
(see~Section~\ref{sec:tiers}).  An earlier per-band tier-midpoint
residual emitter targeted a WSPR-percentile tier truth; that
target was retired when the verdict moved to fixed dB cuts
(Section~\ref{sec:tiers}), since fitting per-band absolute
margins to WSPR percentiles couples the verdict labels to the
WSPR-truth artefact rather than the underlying physics.  In its place the harness writes a per-(path, band)
margin/$\sigma$ baseline and flags any cell that drifts more than
\SI{2}{\dB} in mean margin or 5\,pp in $P(\text{open})$ across runs
--- a self-consistency regression detector that needs no external
truth.

The ``alive'' floor itself comes in two forms, selectable via
\texttt{--ground-truth=\{global,per-path\}}:
\begin{itemize}[nosep]
  \item \textbf{Global} (default, baseline file
    \texttt{harness.baseline.json}):
    band-level ``alive somewhere in the world'' floor of $\geq 50$
    spots/hour, computed from a single bbox-unrestricted WSPR
    aggregate query.  Cheap (one query per harness run, $\sim$7\,k
    rows for a 30-day window), but every reference path sees the same
    open-rate truth, so the per-(path, band) drift signal is
    dominated by the path-independent open-rate bias.
  \item \textbf{Per-path}\label{sec:perpath} (baseline file
    \texttt{harness.baseline.perpath.json}): TX bbox + RX bbox
    aggregate per reference path (bidirectional, default half-width
    $5^\circ$), with a much sparser floor of $\geq 1$ spot/hour
    reflecting the per-pair WSPR density.  Adds 35 queries per
    harness run, but the open-rate truth is path-specific, so the
    per-(path, band) drift signal directly reflects per-pair
    behaviour.  Used to flag a per-mechanism regression that the
    global mode averages out --- for example, the \S\ref{sec:muf-consensus}
    asymmetric-MUF retest, where the bias was concentrated on the
    minority of cells where kc2g sat above climatology.
\end{itemize}
Both modes write to separate baseline files so an operator can
maintain both anchors simultaneously and compare drift in either
register.

A separate driver script,
\texttt{scripts/tests/tune-r7.mjs}, runs coordinate-descent
joint optimisation over the seven-parameter set
$\{L_{\text{iono}},\,D,\,w_{\text{sc}},\,w_{\text{NVIS-tail}},
w_{\text{Es-prim}},\,\sigma\text{-scale},\,\text{fusion-flag}\}$
from three deterministic starting configurations
(Appendix~\ref{app:repro}; ``baseline'', ``fusion-up'',
``modes-on'').  Production values: $L_{\text{iono}} =
\SI{1}{\dB}$, $D = \SI{0.25}{\dB}$, $w_{\text{sc}} = 1.5$,
$w_{\text{NVIS-tail}} = 1.0$ (no scaling vs the bare NVIS-tail
formula of Eq.~\ref{eq:nvistail}),
$w_{\text{Es-prim}} = 1.0$ (no scaling vs the bare Es-mode
formula in \S\ref{sec:es-prop-mode}), $\sigma\text{-scale} = 1.0$
(per-band $\sigma_g$ from Table~\ref{tab:bandsigma} held), and
$\text{fusion-flag}$ false (\texttt{FUSION\_PRIMARY\_MUF =
false} in \texttt{src/constants.js}; the R3 per-hop foF2 fusion
path is implemented but kept off in production until the
per-pair WSPR ground-truth wiring of \S\ref{sec:perpath} can
calibrate it without exploiting the global-truth structural
ceiling).  At the 2026-04-28 calibration freeze, all three
seeds converged to the same basin in $\leq 2$ iterations: the
final $\{L_{\text{iono}},\,D,\,w_{\text{sc}}\}$ landed within
$\pm 10\%$ across seeds; $\sigma\text{-scale}$ and the two
$w_{\text{NVIS-tail}} / w_{\text{Es-prim}}$ weights settled at
$1.0$ from every seed at that freeze (their natural-units
fixed point at the time); $\text{fusion-flag}$ converged to
\emph{false} on every seed.  Final Brier landed within
$\pm 0.001$ of the median seed.  \emph{Drift caveat:} solar
activity has moved the harness target since the freeze (see
the ``Solar-driven drift on re-run'' note in
Table~\ref{tab:harnessruns}).  Re-running the same coordinate
descent against later 30-day windows pulls
$\sigma\text{-scale}$ and $w_{\text{Es-prim}}$ off the $1.0$
fixed point: a 2026-05-06 re-run lands $\sigma\text{-scale}
\approx 2.0$ and $w_{\text{Es-prim}} \approx 1.5$ as the local
optima against that window.  Production constants stay at
their 2026-04-28-freeze values per the explicit no-auto-commit
policy below; the re-run divergence is a calibration-frame
drift signal, not a regression.  Lift across the freeze cycle
was concentrated on 12\,m / 10\,m, driven entirely by the
$w_{\text{sc}} = 1.5$ scatter recovery; the other six
parameters showed no measurable Brier change at the freeze
when swept individually.

The fixed reference values cited throughout the paper are
consolidated in Table~\ref{tab:harnessruns} so each percentage can
be tied to a specific basket / config / metric.

\begin{table}[H]
\centering
\caption{Canonical harness numbers cited in the paper, each tied to
its (basket, ground-truth mode, configuration) so a reader can
disambiguate.  Binary ``open'' threshold is $\geq 50$ spots/h on
the global aggregate or $\geq 1$ spot/h on the per-path aggregate.}
\label{tab:harnessruns}
\begin{tabular}{@{}llrrrr@{}}
\toprule
\textbf{Run} & \textbf{Basket} &
\multicolumn{2}{c}{\textbf{Global truth}} &
\multicolumn{2}{c}{\textbf{Per-path truth}} \\
& & \textbf{accBin} & \textbf{Brier} & \textbf{accBin} & \textbf{Brier} \\
\midrule
Pre-scatter baseline (2026-04 freeze)           & 35 paths   & 91.08\,\% & 0.0625 & --- & --- \\
Post-scatter sweep (2026-04 freeze)             & 35 paths   & 92.35\,\% & 0.0535 & --- & --- \\
Pre-B6 production (2026-04-28 freeze)           & 35 paths   & 94.25\,\% & 0.0386 & 30.97\,\%\,${}^{\mathsection}$ & 0.6353\,${}^{\mathsection}$ \\
Post-B6 production (2026-05-08, current)        & 35 paths   & 84.08\,\% & 0.1081 & 40.17\,\%\,${}^{\mathsection}$ & 0.5575\,${}^{\mathsection}$ \\
\bottomrule
\end{tabular}\\[2pt]
\footnotesize
${}^{\mathsection}$ Per-path-truth metric uses TX/RX-bbox-restricted aggregates with the
$\geq 1$ spot/h floor; the resulting accBin / Brier numbers are
\emph{not directly comparable} to the global-truth column because
the per-path floor is far stricter (most cells read closed) and
because operator-activity sparsity dominates the binary outcome
on minority-traffic bands.  The pre-B6 freeze headline
$94.25\,\% / 0.0386$ and the post-B6 current headline
$84.08\,\% / 0.1081$ are both the global-truth metric and both
inherit the structural ``open somewhere'' ceiling that the
per-path mode was specifically built to expose
(Section~\ref{sec:limits} \#9); the post-B6 numbers are
\emph{lower} on global truth because the new D-region formula
predicts 160\,m closed more often (correctly, per the per-path
metric: 160\,m accBin rose from 19.9\,\% to 37.6\,\% across the
B6 cut).  The per-path column is the more rigorous metric and
is the one that should drive constant retunes; B6 improved it
materially on 160\,m and left the other bands unchanged.  Treat the global number as a
regression-detection signal, not a quality score; the per-path
column is for cell-level drift detection against the saved
baseline.

\smallskip
\textbf{Solar-driven drift on re-run.}  The headline numbers are
the values produced by the 30-day window centred on the
2026-04-28 calibration freeze.  Re-running the same harness
against a later 30-day window will give different absolute
numbers as solar activity shifts the band-by-band open-rate
distribution: a 2026-05-06 fresh run ($F_{10.7}$ rose
$105 \to 156$ over the intervening week) produced
$86.06\,\%$ / $0.0937$ global and $38.29\,\%$ / $0.5740$
per-path.  This is calibration-frame drift, not regression: the
sweep-derived constants are unchanged but the truth signal
itself has moved with the sun.  The committed
\texttt{scripts/data/harness.baseline.json} is the regression
anchor and the ${\sim}\!2$\,dB / $5$\,pp cell thresholds are the
actual drift gates.
\end{table}

The harness does not auto-commit: a human reviews the suggested
constants and pastes the winners into \texttt{src/constants.js}.

\subsection{Circuit Reliability}\label{sec:reliability}

Given the margin $M$ and spread $\sigma$, the circuit reliability is
the probability that the achieved SNR meets or exceeds the mode's
required SNR --- in other words, the probability the link is usable on
a given attempt:
\begin{equation}\label{eq:reliability}
  R(M, \sigma) = \Phi\!\left(\frac{M}{\sigma}\right)
\end{equation}
where $\Phi$ is the standard normal CDF.  JavaScript has no built-in
$\Phi$, so the runtime evaluates it via the Abramowitz \& Stegun
26.2.17 rational approximation (\texttt{src/physics/tier.js},
\texttt{normCdf}); maximum absolute error is ${\sim}\!7.5\times 10^{-8}$
in the tails, well below the verdict-relevant resolution near the
fixed-dB tier boundaries ($M = +18, +6, -5, -14$\,\si{\dB}, which
sit at $z = M/\sigma = +1.5$ to $+2.25$, $+0.5$ to $+0.75$, $-0.4$
to $-0.6$, and $-1.2$ to $-1.75$ respectively for typical per-band
$\sigma$ in 8 to 12\,\si{\dB}; $\partial R/\partial z = \varphi(z)$
at those points runs 0.10 to 0.40).  The approximation choice is
therefore not a budget-shaping parameter; it is fixed for
reproducibility.
Eq.~\ref{eq:reliability} is the Gaussian-spread basic-circuit-
reliability model that underlies ITU-R P.842~\cite{itup842}: the
recommendation itself presents the calculation in piecewise /
tabular form rather than the literal $\Phi(M/\sigma)$ identity
used here, but both reduce to the same Gaussian assumption on
day-to-day SNR fluctuations.  ionocast evaluates this model with
live nowcast inputs --- $M$ is the current-hour budget margin
(Eq.~\ref{eq:blend}), and $\sigma$ is the per-band base $\sigma_g$
(Table~\ref{tab:bandsigma}, set within the ITU-R P.533 day-to-day
spread range 6--12\,dB; ionocast picks 8--12\,dB by band)
augmented in quadrature with the situational widenings of
\S\ref{sec:tier-confidence}.  $R$ drives
the tier verdict (Section~\ref{sec:tiers}) but the band-table UI
surfaces a different percentile alongside the tier label, the
\emph{Stability} metric defined in Section~\ref{sec:tier-confidence}.

\paragraph{Strict P.842 vs.\ instantaneous: an operator-relevant
divergence.}  P.842 as written is a \emph{climatological} predictor:
the formula is evaluated with the monthly-median margin at a given
hour and the day-to-day spread, so $R$ answers ``on what fraction of
days does the median link at this hour exceed the requirement?''
ionocast evaluates the same algebra with live data --- the
current-hour observed kc2g MUF, the live F\textsubscript{10.7A},
$K_p^{\text{eff}}$ from the past 7\,d --- and so $R$ answers
``what is the probability the link is viable \emph{right now}, given
the model's residual uncertainty?''  Both questions use the same Gaussian-spread machinery, but they
are not the same question.  An operator deciding whether to call
CQ at this minute wants the latter; a frequency-coordination
engineer planning circuit allocations across a season wants the
former.  ionocast surfaces the nowcast interpretation; $R$ is
exposed as a side metric for callers that want it (e.g.\ the
\emph{Stability} percentage in Section~\ref{sec:tier-confidence})
but does not drive the tier label.

Tier thresholds (Section~\ref{sec:tiers}) are pinned to fixed dB
margins ($+18 / +6 / -5 / -14$) that do not scale with the band's
$\sigma$.  This is a deliberate revert from an earlier
sigma-percentile cut scheme that ran optimistic in real-world
contests: the percentile boundaries placed Good at $+0.25\sigma$,
which corresponds to a $\sim$2\,\si{\dB} margin on lower HF
($\sigma_g = 8\,\si{\dB}$) and routinely disappeared under the
3 to 6\,\si{\dB} of QRM that contest environments add on top of the
ITU-R P.372 baseline noise the model assumes.  The fixed-dB cuts
take operator intuition as primary: $+6\,\si{\dB}$ Good means the
model expects to clear required SNR by 6\,\si{\dB} \emph{before}
real-world noise adds anything, which lines up with what an
operator calls a usable band on-air.  Per-band $\sigma_g$
continues to drive the Stability and Confidence metrics defined
in Section~\ref{sec:tier-confidence}, just not the boundary
labels.

\paragraph{What $R$ does and does not measure.}  $R$ is the
probability that the \emph{physics} of the link permits a QSO at
the operator's configured rig, conditional on an operator on the
other end transmitting and listening at that hour.  It does not
include operator-availability factors (band-time-of-day social bias,
WSPR-spot population density on the destination side, holiday
patterns, contests, etc.), which are independent of ionospheric
state and not represented in the SNR budget.  An empirical
calibration against 30\,d of WSPR per-path aggregates (TX/RX bbox
restricted, floor 1\,spot/h; the per-path scoring path inside
\texttt{scripts/harness.mjs} when run with \texttt{--per-path})
finds that predicted tier ranks correctly --- excellent $>$ good $>$
fair $>$ poor $>$ closed in observed open-rate ordering --- but the
absolute per-path completion rate runs $\sim$60--65\,pp below the
predicted reliability across the upper four tiers (the predicted
``excellent'' bucket sees a $\sim$30\,\% per-path open rate against
the WSPR cache; ``good'' $\sim$14\,\%; ``fair'' $\sim$3\,\%; ``poor''
$\sim$1\,\%; ``closed'' $\sim$0.1\,\%; these per-tier rates are
from an ad-hoc post-processing of the per-cell
\texttt{harness.report.json}, not from a byTier field that the
standard report currently emits).  The held-out window
(2026-04-26 onwards, post the 2026-04-25 loss-constant calibration)
matches the in-sample window within 1\,pp on every tier, ruling out
a calibration-drift explanation.  The gap is a
physics-vs-operator-availability semantics issue: $R$ captures
``the path's physics permit a QSO with the typical operator's rig''
(physics-permitted reliability), while WSPR per-path open-rate
captures ``a WSPR-active operator pair completed a transmit at
this hour'' (operator-attempt completion rate), which is uniformly
lower because operators do not attempt every available
(band, hour, path) cell.

\subsubsection{Verdict Stability}\label{sec:tier-confidence}

The band-table surfaces a per-band \emph{Stability} percentage,
defined as
\begin{equation}\label{eq:tierstab}
  S(M, \sigma) = \Phi\!\bigl(\min_{b \in \mathcal{B}} |M - b| / \sigma\bigr),
\end{equation}
where $\mathcal{B} = \{-14, -5, +6, +18\}\,\si{\dB}$ is the set
of fixed-dB tier boundaries (Table~\ref{tab:tiers}).
Operator reading:
\emph{how likely is the verdict to NOT cross its nearest tier
boundary if the true margin moves to its expected value?} The
metric is one-sided, it asks about the closest boundary only, not
both flanks of the bucket, and bucket-width-independent: a Fair
verdict $\Delta z$ from its nearest boundary reads the same
percentage as an Excellent verdict $\Delta z$ from its boundary.

\paragraph{Reachable values.} What $S$ can show depends on the
bucket, because the maximum nearest-boundary distance depends on
bucket width:
\begin{itemize}[nosep]
  \item \textbf{Closed} ($z < -1.28$) and \textbf{Excellent}
        ($z \ge 1.28$): open-ended on one side. $S$ rises
        monotonically the further $z$ moves into the tail:
        $\sim 84\%$ at $1\sigma$ past the boundary, $\sim 97\%$ at
        $2\sigma$, $\to 1.0$ deep in the tail.
  \item \textbf{Poor} ($\sim 0.90\sigma$ wide), \textbf{Fair}
        ($\sim 0.64\sigma$ wide), \textbf{Good} ($\sim 1.03\sigma$
        wide): max nearest-boundary distance is half each bucket's
        width, so $S$ caps at $\Phi(0.45) \approx 67\%$ for Poor,
        $\Phi(0.32) \approx 63\%$ for Fair, and $\Phi(0.51) \approx
        70\%$ for Good at each bucket centre.  $S = 50\%$ on a
        boundary; the centred-bucket maxima are the most stable a
        middle-tier verdict in that bucket can read.
\end{itemize}
Read $S$ on the open buckets against an absolute scale (deep
Excellent / Closed approach 100\%); read $S$ on the middle buckets
relative to their per-bucket centred maxima.

\paragraph{One-sided semantics.} $S$ measures the probability that
pure-$\sigma$ noise does not cross the \emph{nearest} boundary; it
ignores the far boundary on a finite-width bucket. For a centred
Fair verdict ($z = -0.066$, halfway between $-0.385$ and $+0.253$)
the far boundary is also $0.32\sigma$ away and is not counted, so
$S = 63\%$ rather than the two-sided in-bucket probability
$2\Phi(0.32) - 1 \approx 25\%$. The one-sided form is the correct
``how locked in is this verdict against the most likely flip''
indicator and is what operators typically want; the two-sided form
is the literal $P(\text{predicted} = \text{true})$ retained as
\texttt{tierConfidence} in the source.  The earlier band-table
column showed the two-sided form, which capped at $\sim 25$--$35\%$
for centred middle-tier verdicts and required a ``\,(peak)''
annotation; the new $S$ replaces it.

\paragraph{Operator-attempt completion rate (deferred).}\label{sec:completion}
The band-table currently surfaces only the physics-permitted tier
label.  A planned \emph{Expected Completion} (\emph{Comp.})
column would fold the operator-availability gap discussed above
into a single percentage; the formulation below documents that
deferred display so the activity-prior factor it depends on
(already computed via \texttt{spotBaselineMean} in
\texttt{src/derive/spots.js}) has a published target shape:
\begin{equation}\label{eq:completion}
  \text{Comp.}(b, h, M, \sigma) =
    R(M, \sigma) \cdot a(b, h),
\end{equation}
where $a(b, h) = \text{baseline}(b, h) / \max_h \text{baseline}(b, h)$
is the per-band operator-availability prior, derived from the
30-day per-(band, UTC-hour) WSPR baselines that already drive the
spot-override (\S\ref{sec:spot-override}). $a(b, h) = 1$ at the
band's peak activity hour; $a(b, h) \to 0$ deep in the band's
dead-of-night for that band's typical user population.  Comp.\
would read as ``what fraction of attempts at this rig in this
hour actually complete a QSO,'' which is the question an operator
picking a band to call CQ on is asking. Tier and Comp.\ together
would disambiguate the physics-vs-availability gap: an Excellent
tier on 12\,m at 0300\,UTC reads $\sim$95\,\% physics-permitted
(R) and $\sim$5--10\,\% Comp.\ --- the band is open, but nobody
is listening. Once the column ships, the intended operator
guidance is: use Tier when planning around the physics, use
Comp.\ when picking a band to actually call CQ on right now.

The factorisation $R \cdot a(b, h)$ assumes physics-permitted
reliability and operator-availability are independent.  This holds
strictly within a band in steady state -- $a(b, h)$ is normalised
against the band's own 24-hour maximum, so it encodes
\emph{intra-band} hourly activity variation and not inter-band
reliability.  Across bands the absolute Comp.\ values are not
directly comparable (a band with chronically thin WSPR-pair density
can never reach Comp.\ $\to 1$ even when fully open), which is why
the planned Comp.\ column would surface the percentage per row
rather than as a cross-band ranking.  In the limit of a future per-pair availability
prior with band-specific ceilings, the factorisation would need
refining to account for the fact that bands with high WSPR activity
have those operators \emph{because} the band is reliably open ---
a feedback that the current intra-band normalisation sidesteps but
does not eliminate.

\paragraph{Storm-time correlation breaks the independence
assumption.}  During a major geomagnetic storm both $R$ drops
(physics) and $a(b, h)$ should drop too (operators on the affected
bands stop calling because they observe the band is dead) ---
i.e.\ $R$ and $a$ would be positively correlated under exactly
the disturbance conditions where Comp.\ would be most
operationally relevant.  The factorisation $R \cdot a(b, h)$
treats them as independent, which would bias Comp.\ \emph{high}
during storms (the product of the depressed $R$ with the
steady-state $a$ overstates expected completion versus the true
$R \cdot a_{\text{actual}}$ where $a_{\text{actual}} < a$).  The
intra-band normalisation softens but does not remove this:
storm-time activity dropouts span all bands roughly together, so
the per-band 24-hour maximum baseline still encodes a quiet-day
shape that storm hours genuinely depart from.  When the column
ships, Comp.\ during $K_p^{\text{eff}} \geq 5$ conditions should
be read as an upper bound on expected completion, not a point
estimate; the storm-aware refinement is tracked in
Section~\ref{sec:limits} and is gated on the per-pair
availability data wiring of Section~\ref{sec:perpath}.

The spread $\sigma$ is \emph{per-band and condition-dependent},
computed as the RSS of a per-band base $\sigma_g$
(Table~\ref{tab:bandsigma}; values anchored within the ITU-R
P.533 day-to-day spread literature range 6--12\,dB, with
ionocast picking the upper-half subset 8--12\,dB across the
table) and seven situational penalties
(six enumerated below plus the Es-active mode contribution
defined in \S\ref{sec:es-prop-mode}).\footnote{Notation: $\sigma$ is overloaded across the
paper for distinct quantities --- the prediction-spread family in
this section ($\sigma$, $\sigma_g$, $\sigma_{\text{MUF}}$,
$\sigma_{\text{storm}}$, $\sigma_{\text{forecast}}$,
$\sigma_{\text{term}}$, $\sigma_{\text{recovery}}$,
$\sigma_{\text{night}}$, $\sigma_{\text{Es}}$); ground
conductivity in Eq.~\ref{eq:lgr} (in \si{S/m}); the Gaussian
kernel widths $\sigma_{\text{cr}}$, $\sigma_{\text{tr}}$ in the
EIA term (Eq.~\ref{eq:eia}, in degrees); the per-hop foF2
standard deviation $\sigma_{\!f}$ in the F2-scatter recovery
(Section~\ref{sec:scatter}); the standard-normal CDF argument
in Eq.~\ref{eq:reliability}'s $\Phi(M/\sigma)$; and the
dimensionless $\sigma$-scale tuning parameter on the
coordinate-descent grid in Appendix~\ref{app:repro}
(Table~\ref{tab:tunegrid}, which uniformly multiplies the
per-band $\sigma_g$ table during calibration sweeps; held at
1.0 in production).  The unit-ramp clamp formerly written as
$\sigma(x)$ has been renamed $R(x)$ in \S\ref{sec:espersistence}
to avoid collision.  The remaining distinct meanings are
recoverable from local context (units, subscripts) but the
overloading is real.}
\begin{itemize}[nosep]
  \item \textbf{Near-MUF:} when $f/\text{MUF} > 0.85$, fading and
        multipath intensify.  Adds up to \SI{+4}{\dB} at
        $f = \text{MUF}$.
  \item \textbf{Storm:} $K_p^{\text{eff}} \geq 5$ (the
        storm-lagged effective $K_p$ from
        Section~\ref{sec:kplag}, after Bz-forward and Dst bumps)
        produces ionospheric irregularities.  Linear in
        $K_p^{\text{eff}}$ above the activity threshold:
        $\sigma_{\text{storm}} = 3 + 0.75\,(K_p^{\text{eff}} - 5)$\,dB
        for $K_p^{\text{eff}} \geq 5$, zero otherwise; that
        gives \SI{+3}{\dB} at $K_p^{\text{eff}} = 5$ and
        \SI{+6}{\dB} at $K_p^{\text{eff}} = 9$.
        Using the effective rather than live $K_p$ means the storm
        $\sigma$ persists into the recovery phase (when the live
        index has already dropped but the F-region thermosphere is
        still recovering).  $\sigma_{\text{recovery}}$ below fires
        only when the phase classifier identifies recovery, so the
        two penalties stack during recovery (covering both the
        residual lagged-Kp irregularities and the LSTID ripple)
        and the storm penalty alone applies during main phase.
        The same $K_p^{\text{eff}}$ also drives the auroral
        amplification described later in this section.\footnote{In
        the source code, \texttt{snrMarginHf} (and the auroral
        loss terms it consumes) take a parameter named
        \texttt{opts.kp}.  This is \emph{always}
        $K_p^{\text{eff}}$ (i.e.\ the storm-lagged value with
        Bz-forward and Dst bumps applied) rather than the live
        3-hour $K_p$ sample; the bump-and-lag composition happens
        upstream in \texttt{src/derive/conditions.js} before the
        budget is invoked.  A reader auditing the code without
        the data-flow context could mistake the unqualified
        \texttt{kp} name for the live index.}
  \item \textbf{Forecast storm:} the SWPC 3-day Kp forecast shows a
        peak $K_p^{\text{peak}} \geq 5$ within the next 6\,h.  Sharp
        electron-density gradients build hours before the index reading,
        so $\sigma$ widens ahead of the disturbance.  Same magnitude
        as $\sigma_{\text{storm}}$ above evaluated at $K_p^{\text{peak}}$
        (\,\SI{+3}{\dB} at $K_p^{\text{peak}} = 5$, \SI{+6}{\dB} at
        $K_p^{\text{peak}} = 9$\,), attenuated by $0.7\times$ for
        forecast confidence, then ramped down linearly as the
        current $K_p$ catches up to the forecast peak:
        \begin{equation}
          \sigma_{\text{forecast}} =
            0.7\,\sigma_{\text{storm}}\!\left(K_p^{\text{peak}}\right)
            \cdot \min\!\bigl(1,\,\max(0,\,K_p^{\text{peak}} - K_p^{\text{eff}})\bigr)
        \end{equation}
        The catch-up ramp is gated on $K_p^{\text{eff}}$ rather than
        the live 3-hour $K_p$, matching the $\sigma_{\text{storm}}$
        branch above and preserving the smooth-ramp philosophy applied
        at every other gate; using the live index here would step the
        forecast bump down on each 3-hour tick rather than as the
        thermosphere actually catches up.
        Without this gap-based gate, summing in quadrature with the
        current-Kp $\sigma_{\text{storm}}$ branch would have produced
        $\sqrt{\sigma_{\text{storm}}^2 + 0.49\,\sigma_{\text{storm}}^2}
        \approx 1.22\,\sigma_{\text{storm}}$ at the moment of catch-up
        (a 22\% inflation), instead of the intended hand-off to
        $\sigma_{\text{storm}}$ alone.  The 1\,Kp-unit saturation
        width of the catch-up ramp is empirical: $K_p^{\text{eff}}$
        typically climbs by ${\sim}1$ unit per hour during onset,
        so a 1-unit gap maps to roughly an hour of forecast
        prominence ahead of the live index, which matches the
        operationally-useful early-warning horizon of the SWPC
        3-day forecast.  A wider saturation window would extend the
        forecast bump deeper into peak hours and double-count with
        $\sigma_{\text{storm}}$; a narrower window would cut the
        forecast bump off before the live index has caught up at
        slow-onset events.
  \item \textbf{Cross-terminator:} when the path midpoint has
        $|\cos\chi| \le 0.15$ (sunrise/sunset line), sharp
        electron-density gradients increase variance by \SI{+3}{\dB}
        (the variance contribution is $9\,\text{dB}^2$, added in
        quadrature to the per-band $\sigma_g^2$).  Above $|\cos\chi|
        \ge 0.20$ the contribution is zero; in the
        $|\cos\chi| \in [0.15, 0.20]$ band the variance contribution
        ramps linearly from $9\,\text{dB}^2$ down to $0$, matching
        the smooth-ramp philosophy applied at every other gate
        (auroral $c(\phi)$, Bz bump, twilight $r(\cos\chi_k)$,
        near-MUF, Es threshold) and avoiding a tier-confidence
        cliff at terminator approach.
  \item \textbf{Storm-recovery TID:} during the recovery phase of a
        geomagnetic storm (Dst depressed but $K_p^{\text{eff}}$
        trailing the live $K_p$, see Section~\ref{sec:kplag}),
        large-scale travelling ionospheric disturbances ripple
        through the F-region for hours after the surface-field
        index settles.  The mean margin is no longer suppressed but
        the path variance is visibly elevated; ionocast adds
        \SI{+4}{\dB} to $\sigma$ across all paths until the storm
        phase classifier exits ``recovery''.
  \item \textbf{Night-time low/mid HF inflation:} ITU-R P.533
        within-month variability for night-time multi-Mm paths on
        160\,m through 30\,m runs ${\sim}\!8\text{--}10$\,dB; the
        per-band $\sigma_g$ table sits at the 8\,dB floor on those
        bands, anchored to a quiet-day reference.  Without an extra
        bump, an ``excellent'' verdict at $+11$\,dB margin on a
        quiet night-time 30\,m / 20\,m path over-claims reliability
        by about one tier.  ionocast adds \SI{+3}{\dB} in
        quadrature when the path midpoint is in darkness
        ($\cos\chi_{\text{path}} < 0$) and $f \le \SI{16}{\MHz}$
        (i.e.\ 17\,m and below, where the tabulated $\sigma_g$ is
        not already inflated by near-MUF variability).  Above
        17\,m the table value already reflects the higher
        upper-band spread, so further inflation would
        double-count.
\end{itemize}
All penalties are added in quadrature:
\begin{equation*}
\sigma = \sqrt{\sigma_g^2 + \sigma_{\text{MUF}}^2 + \sigma_{\text{storm}}^2
  + \sigma_{\text{forecast}}^2 + \sigma_{\text{term}}^2 +
  \sigma_{\text{recovery}}^2 + \sigma_{\text{night}}^2 +
  \sigma_{\text{Es}}^2}
\end{equation*}
with the night-time inflation
$\sigma_{\text{night}}^2 = (\SI{3}{\dB})^2$ entering only when
$\cos\chi_{\text{path}} < 0$ and $f \le \SI{16}{\MHz}$, and the
Es-active contribution $\sigma_{\text{Es}}^2 = (\SI{2}{\dB})^2$
entering only when Es is selected as the propagation channel
(\S\ref{sec:es-prop-mode}; the Es-mode budget consumes the same
RSS via the shared \texttt{\_conditionalSigmaDb} helper, so
storm-time, terminator, recovery, and night-low/mid penalties
all propagate into Es-mode verdicts the same way they do for
F2).

\paragraph{Why $\sigma_{\text{storm}}$ and $\sigma_{\text{recovery}}$
are path-blind.}  $\sigma_{\text{MUF}}$ and $\sigma_{\text{term}}$
are evaluated at the path midpoint (so they vary path-by-path
across the 10-path basket), but
$\sigma_{\text{storm}}$, $\sigma_{\text{forecast}}$, and
$\sigma_{\text{recovery}}$ are global: they fire on the same
$K_p^{\text{eff}}$ and storm-phase signals regardless of whether
any hop on the path actually crosses an auroral CGM latitude.  This
is intentional rather than a simplification.  Storm-time variance
arises from at least three mechanisms whose footprints are not
confined to high latitude: large-scale travelling ionospheric
disturbances (LSTIDs) launched at the auroral oval propagate
equatorward at ${\sim}\SI{300}{m/s}$ to ${\sim}\SI{1000}{m/s}$ and
reach midlatitudes within hours; thermospheric Joule heating
shifts the global [O]/[N\textsubscript{2}] ratio, depressing
foF2 by 10--20\,\% on midlat paths during main phase; and
plasma-bubble / spread-F activity at low and equatorial latitudes
during the recovery phase produces irregularity-driven fading on
transequatorial paths that share no auroral hops.  The global
$\sigma$ widening captures this collectively.  The
\emph{path-localized} component of storm response --- additional
auroral D-region absorption on hops crossing the storm-expanded
$\phi_{\text{CGM}}$ threshold --- is captured separately in
$L_{\text{aur}}$ (Section~\ref{sec:laur}, with the
storm-main amplification described below), so a midlat path during a
G3 storm sees the global $\sigma_{\text{storm}}$ widening but no
$L_{\text{aur}}$ contribution, while an auroral path sees both.
The combination correctly under-widens neither: midlat paths get
the global-irregularity penalty without the absorption term, and
auroral paths get both the absorption (mean) and the widening
(spread).  The remaining bias is a slight over-widening of
\emph{deep equatorial} midlat paths during main phase (where LSTID
arrival is delayed by an hour or two relative to the global signal),
worth at most ${\sim}\SI{1}{\dB}$ of $\sigma$ inflation on those
paths --- below the per-band $\sigma_g$ resolution and accepted as
the cost of the global formulation.

\paragraph{Storm-phase auroral amplification.}\label{par:laur-stormamp}  Independent of the
$\sigma$ widening above, the auroral absorption term $L_{\text{aur}}$
(Section~\ref{sec:laur}) is incremented by \SI{+4}{\dB} per hop
during the storm \emph{main} phase, then re-clamped against the
\SI{30}{\dB} per-hop cap (\texttt{AUR\_PER\_HOP\_CAP\_DB}).
During main, the auroral oval expands equatorward and the
particle precipitation that drives auroral D/E absorption is
strongest; the steady-state $K_p$-keyed $L_{\text{aur}}$
undercounts the depth.  The \SI{+4}{\dB} additive bump is
calibrated to operator-experience reports of typical
extra absorption (3--6\,dB) on high-latitude amateur HF paths
during peak Dst depression.  ITU-R P.533~\cite{itup533}
Annex 1 Table 2 tabulates climatological auroral-and-other
losses ($L_h$) keyed on geomagnetic latitude, season, and
local time --- with values reaching 12--16\,dB on
auroral-zone (62.5--72.5° geomagnetic) winter dawn paths and
under 3\,dB on midlatitude paths --- but does not isolate a
``main-phase extra'' increment as a separate column.  The
3--6\,dB range cited here is therefore an
operator-experience figure consistent with the disturbed-vs-
quiet difference implicit in P.533's seasonal climatology
rather than a literal P.533 entry.  An earlier draft used a $1.4\times$ multiplier,
which would have produced ${+}\SI{10}{\dB}$ on a 25\,dB nominal
$L_{\text{aur}}$ (well past the 3--6 dB range cited as the
target); the additive form lands at the cited midpoint
regardless of the steady-state value, with the per-hop cap as
the hard ceiling for already-saturated terms.  Recovery and
quiet phases keep the steady-state value.  Storm phase is
classified from Dst trajectory plus the storm-lag kernel
(initial / main / recovery / quiet / active); main is
identified by $\text{Dst} \le -50$\,nT with $K_p^{\text{eff}}$
still loading.

\begin{table}[H]
\centering
\caption{Per-band base $\sigma_g$, anchored to the ITU-R P.533
day-to-day spread range and validated against per-spot wspr-snr
residual standard deviation bucketed by (band, distance,
hour-of-day, Kp, tx-latitude band) at $\geq$10 samples per
bucket; the measured within-bucket $\sigma$ was stable across
three bucket-granularity settings (coarse 1500\,km / 6\,h vs
tight 500\,km / 1\,h), confirming the spread is within-condition
rather than cross-condition contamination.  12\,m / 10\,m fall
back to operator-experience values (fewer than five buckets per
band; the WSPR signal is too weak to drive a within-condition
fit there).}
\label{tab:bandsigma}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Band} & {$\sigma_g$ (\si{\dB})} & {within-cond data (\si{\dB})} \\
\midrule
160\,m &  8 &  8.5 \\
80\,m  &  8 &  8.0 \\
60\,m  &  8 &  8.0 \\
40\,m  &  8 &  8.5 \\
30\,m  &  9 &  9.0 \\
20\,m  &  9 &  9.5 \\
17\,m  &  9 &  9.0 \\
15\,m  & 10 & 10.0 \\
12\,m  & 12 & (n=2 weak) \\
10\,m  & 12 & (n=3 weak) \\
6\,m   &  8 & (no VHF data) \\
2\,m   &  8 & (no VHF data) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{$\sigma_g$ sensitivity.}  Under the current fixed-dB
tier cuts (+18 / +6 / -5 / -14, Section~\ref{sec:tiers}), the
verdict label itself is independent of $\sigma_g$: the same
margin produces the same tier on every band regardless of
whether 10\,m's $\sigma_g$ is 10 or 12\,\si{\dB}.  $\sigma_g$
enters only the per-band Stability metric (the dB-distance to the
nearest fixed boundary, divided by $\sigma_g$, fed through $\Phi$;
Eq.~\ref{eq:tierstab}) and the side-metric reliability $R$.  This
is a deliberate trade-off against the earlier sigma-percentile
scheme, where a $\pm \SI{2}{\dB}$ miscalibration in $\sigma_g$
could flip the tier label on near-boundary bands; under fixed-dB
cuts the label is invariant to that calibration uncertainty and
$\sigma_g$ controls only the operator's read of how confident
the label is.  $\sigma_g$ is held at the P.533 anchor rather
than fitted, on the grounds that machine-fitted $\sigma$ would
chase the WSPR-truth structural ceiling
(Section~\ref{sec:limits}~\#9) instead of physical day-to-day
spread.

% ── Tier probability figure ──────────────────────────────────────────
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=13cm, height=7cm,
  xlabel={SNR Margin (\si{\dB})},
  ylabel={Probability Density},
  xmin=-35, xmax=35,
  ymin=0, ymax=0.065,
  ytick={0,0.01,0.02,0.03,0.04,0.05,0.06},
  xtick={-30,-25,-20,-15,-10,-5,0,5,10,15,20,25,30},
  xticklabel style={font=\scriptsize},
  grid=major,
  grid style={gray!30},
  title={Tier Probability Distribution ($M=+3$\,dB, $\sigma=8$\,dB,
         fixed-dB boundaries +18 / +6 / -5 / -14)},
  title style={font=\small\bfseries},
  axis on top,
]

% Gaussian PDF, mu=3, sigma=8.
% Fixed-dB tier boundaries (post 2026-05-10 revert from sigma-percentile):
%   excellent at +18 dB
%   good      at  +6 dB
%   fair      at  -5 dB
%   poor      at -14 dB
\pgfmathsetmacro{\mmu}{3}
\pgfmathsetmacro{\sig}{8}

\addplot[name path=pdf,black,very thick,domain=-35:35,samples=200]
  {1/(\sig*sqrt(2*pi))*exp(-(x-\mmu)^2/(2*\sig^2))};

\addplot[name path=zero,draw=none,domain=-35:35,samples=2] {0};

% Closed: < -14
\addplot[name path=pdfC,draw=none,domain=-35:-14,samples=100]
  {1/(\sig*sqrt(2*pi))*exp(-(x-\mmu)^2/(2*\sig^2))};
\addplot[name path=zeroC,draw=none,domain=-35:-14,samples=2] {0};
\addplot[red!30] fill between[of=pdfC and zeroC];

% Poor: -14 to -5
\addplot[name path=pdfP,draw=none,domain=-14:-5,samples=100]
  {1/(\sig*sqrt(2*pi))*exp(-(x-\mmu)^2/(2*\sig^2))};
\addplot[name path=zeroP,draw=none,domain=-14:-5,samples=2] {0};
\addplot[orange!30] fill between[of=pdfP and zeroP];

% Fair: -5 to +6
\addplot[name path=pdfF,draw=none,domain=-5:6,samples=100]
  {1/(\sig*sqrt(2*pi))*exp(-(x-\mmu)^2/(2*\sig^2))};
\addplot[name path=zeroF,draw=none,domain=-5:6,samples=2] {0};
\addplot[yellow!30] fill between[of=pdfF and zeroF];

% Good: +6 to +18
\addplot[name path=pdfG,draw=none,domain=6:18,samples=100]
  {1/(\sig*sqrt(2*pi))*exp(-(x-\mmu)^2/(2*\sig^2))};
\addplot[name path=zeroG,draw=none,domain=6:18,samples=2] {0};
\addplot[green!30] fill between[of=pdfG and zeroG];

% Excellent: > +18
\addplot[name path=pdfE,draw=none,domain=18:35,samples=100]
  {1/(\sig*sqrt(2*pi))*exp(-(x-\mmu)^2/(2*\sig^2))};
\addplot[name path=zeroE,draw=none,domain=18:35,samples=2] {0};
\addplot[blue!20] fill between[of=pdfE and zeroE];

% Threshold lines (fixed-dB)
\draw[gray,thin,dashed] (axis cs:-14,0) -- (axis cs:-14,0.065);
\draw[gray,thin,dashed] (axis cs:-5,0)  -- (axis cs:-5,0.065);
\draw[gray,thin,dashed] (axis cs:6,0)   -- (axis cs:6,0.065);
\draw[gray,thin,dashed] (axis cs:18,0)  -- (axis cs:18,0.065);

% Region labels
\node[font=\scriptsize,red!70!black]    at (axis cs:-22,0.005)  {Closed};
\node[font=\scriptsize,orange!80!black] at (axis cs:-9.5,0.014) {Poor};
\node[font=\scriptsize,olive]           at (axis cs:0.5,0.040)  {Fair};
\node[font=\scriptsize,green!50!black]  at (axis cs:12,0.022)   {Good};
\node[font=\scriptsize,blue!60!black]   at (axis cs:25,0.005)   {Excellent};

% Probabilities for M=+3, sigma=8 under fixed-dB boundaries:
%   z_p = (-14 - 3) / 8 = -2.125 ; Phi = 0.0168
%   z_f = ( -5 - 3) / 8 = -1.000 ; Phi = 0.1587
%   z_g = ( +6 - 3) / 8 = +0.375 ; Phi = 0.6462
%   z_e = (+18 - 3) / 8 = +1.875 ; Phi = 0.9696
%   P(Closed)    = 0.0168                ->  2 %
%   P(Poor)      = 0.1587 - 0.0168 = 0.1419 -> 14 %
%   P(Fair)      = 0.6462 - 0.1587 = 0.4875 -> 49 %
%   P(Good)      = 0.9696 - 0.6462 = 0.3234 -> 32 %
%   P(Excellent) = 1 - 0.9696      = 0.0304 ->  3 %
\node[font=\scriptsize,anchor=north] at (axis cs:-22,0.0035) {2\%};
\node[font=\scriptsize,anchor=north] at (axis cs:-9.5,0.012) {14\%};
\node[font=\scriptsize,anchor=north] at (axis cs:0.5,0.038)  {49\%};
\node[font=\scriptsize,anchor=north] at (axis cs:12,0.020)   {32\%};
\node[font=\scriptsize,anchor=north] at (axis cs:25,0.0035)  {3\%};

% Mean marker
\draw[black,thick,->] (axis cs:3,0.058) -- (axis cs:3,0.0510);
\node[font=\scriptsize,anchor=south] at (axis cs:3,0.058) {$M=+3$};

\end{axis}
\end{tikzpicture}
\caption{Tier probability distribution for a margin of
         $+3$\,dB with $\sigma=8$\,dB under the fixed-dB tier
         boundaries in Table~\ref{tab:tiers}.  Boundaries sit at
         $-14, -5, +6, +18$\,\si{\dB} and do not move with $\sigma$.
         The Fair band is widest at $11$\,\si{\dB} ($-5$ to $+6$);
         Good is $12$\,\si{\dB} wide ($+6$ to $+18$); Excellent and
         Closed are open-ended.  At $M=+3$ the most likely true
         tier is Fair (49\%), then Good (32\%): $M$ sits in the
         Fair bucket but only $3$\,\si{\dB} below the Good
         boundary, so a positive Gaussian fluctuation of $0.4\sigma$
         flips the verdict.  The $\sigma = \SI{8}{\dB}$ shown is the
         6\,m / 2\,m placeholder (Table~\ref{tab:bandsigma}); HF
         bands use $\sigma_g \in \{8, 9, 10, 12\}$\,dB.  $\sigma$
         changes the within-bucket density profile but not the
         bucket boundaries.}
\label{fig:tierprob}
\end{figure}

% ═════════════════════════════════════════════════════════════════════
\section{Tier Mapping}\label{sec:tiers}

Verdicts are computed \emph{per band} (160\,m through 10\,m
individually, plus 6\,m and 2\,m for VHF), not in groups.  For
each band the SNR margin is evaluated on every path in the radial
basket (12 bearings $\times$ 6 distance rings = 72 paths,
Section~\ref{sec:kc2gobs}); the \textbf{best-margin} path wins and
its margin drives the verdict, while the reason string and
\texttt{openDirs} list record which compass direction that was.
If a single path is viable, the band is not ``closed'' --- it is
open in that direction, which is material information for the
operator.  A median-margin aggregation would dominate the verdict
with the failing long paths because the basket samples six
distance rings out to 16\,Mm, most of which are over-MUF or
night-crossing at any given moment; the median would bury
genuinely open shorter paths (a typical mid-cycle daytime
\SI{2500}{\km} hop reporting \SI{+20}{\dB} margin would have its
verdict overwritten by the median of dozens of dead long-distance
samples).

\paragraph{Best-path summary, full-detail companion.}  The
band-table row shows the best-margin direction only (the
``Best Path'' column, formatted as a 16-point compass label
plus distance, e.g.\ ``WNW 9.0\,Mm''), which is a deliberate
compression: displaying all 72 path columns per row would be
visually unworkable.  An operator who wants the underlying
per-direction breakdown consults the Reference Paths panel in
the Ionosphere section, which lists the kc2g MUF on a 12-row
display subset (the longest viable ring per bearing) of the
72-path basket, with a colour-coded f/MUF column for the band
currently selected.  The two surfaces share the same underlying
budget; the band-table row is the headline, the Reference Paths
panel is the source-of-truth.  The Best-Path column itself is
therefore only a hint --- if it reads ``WNW 9\,Mm'' on 20\,m and
the operator wants south-east Asia, the Reference Paths panel
is what to consult next.

The continuous margin is discretised into five tiers using
fixed-dB cuts (Table~\ref{tab:tiers}).  The cuts are absolute and
do not scale with the per-band prediction uncertainty
$\sigma_g$.  The tier label answers ``does the model expect a
usable signal on this band right now, with the station the
operator has configured?'' and is therefore a property of the
user's actual rig.  The defaults (100\,W SSB, 5\,dBi horizontal
dipole, suburban noise) match the canonical first-visit operator
profile; any change to power, mode, antenna, or noise environment
recomputes the SNR budget and reshapes the verdict accordingly.

Tier verdicts are driven by the SNR budget plus the additive
mode bonuses (Eq.~\ref{eq:blend}: $B_{\text{grayline}} +
\max(B_{\text{TEP}}, B_{\text{scatter}})$); the WSPR
spot-override of \S\ref{sec:spot-override} can additionally
promote a band-group summary tier from below-Good up to Good
when the activity ratchet fires, but does not enter the per-band
row's tier or stability columns.  The harness retains a binary
``$\geq 50$ spots/h'' WSPR~\cite{wsprlive} activity check as a
sanity floor only (it does not feed the runtime verdict) since
spot activity measures who is transmitting rather than whether a
QSO would complete on the configured station; the per-path mode
of the harness uses a sparser $\geq 1$ spot/h floor over TX/RX
bbox-restricted aggregates, documented in \S\ref{sec:calibrate}
(Table~\ref{tab:harnessruns}).

\begin{table}[H]
\centering
\caption{Tier thresholds.  Margin is the dB difference between
predicted SNR and the mode's required SNR (Table~\ref{tab:modesnr})
including all additive mode bonuses; the cuts are absolute dB
values, identical across all bands.  The corresponding circuit
reliability $R = \Phi(M/\sigma)$ varies with the per-band
$\sigma_g$ but is no longer the boundary criterion.}
\label{tab:tiers}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Tier} & \textbf{Margin} & \textbf{Meaning} \\
\midrule
Excellent & $\geq +18\,\si{\dB}$       & Loud and easy, almost any antenna \\
Good      & $\geq +6\,\si{\dB}$        & Clear copy, normal QSO conditions \\
Fair      & $\geq -5\,\si{\dB}$        & Workable with effort, careful timing \\
Poor      & $\geq -14\,\si{\dB}$       & Long shot, weak signals at the floor \\
Closed    & $< -14\,\si{\dB}$          & Band effectively closed \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Why fixed dB rather than $R$-percentile cuts.}  An
earlier revision of the verdict mapped tiers to fixed
reliability buckets ($R = 0.90 / 0.60 / 0.35 / 0.10$, equivalent
to margin $\geq 1.2816\,\sigma / 0.2533\,\sigma / -0.3853\,\sigma /
-1.2816\,\sigma$).  It read well as operator-facing prose
(``Good $=$ QSO completes 3 in 5 attempts'') but a 2026-05-10
field-day audit by the author showed it labelled bands as Good
or Fair that were unworkable on-air; only 20\,m, 17\,m, and
15\,m carried real traffic at the time the model was reading
five HF bands as Good or better.  The mismatch came from two
sources.  First, the percentile cut at $+0.2533\,\sigma$
puts Good at margin $\approx +2$\,\si{\dB} on lower HF
($\sigma_g = \SI{8}{\dB}$); a 2\,\si{\dB} model margin
disappears under the 3 to 6\,\si{\dB} of QRM that a contest
environment adds on top of the ITU-R P.372~\cite{itup372}
baseline noise the model assumes.  Second, the formal
operator-readable framing turned out to over-promise: ``3 in 5
QSOs complete'' encoded a $60\%$ floor that does not survive
contact with on-air reality where adjacent-channel splatter,
fading, and operator-side antenna or noise mis-estimation all
push outcomes downward.  The fixed-dB cuts take operator
intuition as primary: $+6\,\si{\dB}$ as Good means the model
expects to clear required SNR by 6 dB \emph{before} real-world
noise adds anything, which lines up with what an operator
calls a usable band on-air.  The boundaries are now
$\sigma$-independent; band-specific $\sigma_g$ continues to
drive the stability and confidence columns.

\subsection{WSPR Activity Override}\label{sec:spot-override}

WSPR spot counts give an independent observational signal that the
physics budget does not consume directly.  When observed activity for
a band exceeds 1.3$\times$ the band's 30-day mean for the current
UTC hour \emph{and} the physics-derived tier is currently below
``Good'' (i.e., Fair, Poor, or Closed), the verdict is promoted to
Good.  This catches openings outside the budget --- weak
sporadic-E, off-axis scatter, propagation modes that the
climatology underweights.

The override is a strict one-way ratchet \emph{capped at Good as a
promotion ceiling, not a global ceiling}: a physics-Excellent
verdict is preserved (the tier-rank condition is false when the
prior tier is already at or above Good, so the override never
fires); a physics-Fair verdict can be promoted to Good but no
higher; physics-Poor and physics-Closed can also only be promoted
to Good. Low or absent activity does not invalidate the physics;
it might just mean nobody is on the air, so the override never
demotes.

\paragraph{Where the override appears in the UI.}  Two surfaces
display per-band conditions and they handle the override
differently --- by design, to avoid pathologies where the row tier
and the row Stability percent appear to disagree:
\begin{itemize}[nosep]
  \item \textbf{Per-band table rows} (HF Bands panel): the Tier
        column and the Stability column both come from the
        physics-derived $M, \sigma$ via
        \texttt{bestPerBand[name].tier} and
        \texttt{bestPerBand[name].stability}.  The override does
        \emph{not} enter individual rows.  Tier and Stability are
        therefore internally consistent within each row: a centred
        physics-Fair row reads ``Fair'' near the 66\,\% middle-tier
        ceiling; a tail-deep physics-Excellent row reads
        ``Excellent'' near 100\,\%.
  \item \textbf{Band-group summary line} (e.g., ``20\,m: Good ·
        margin 5\,dB · unusually active: 100 spots/h vs avg 30''):
        the Tier comes from the override-promoted value when the
        override fires.  In that case, the Stability percent is
        \emph{suppressed} from the line and replaced by the
        activity note --- the operator sees \emph{why} the group
        was promoted rather than a numerically inconsistent
        Stability computed against the wrong tier bucket.
\end{itemize}
The architecture deliberately splits per-band detail (physics
truth) from group-level summary (override-aware) so neither
display ever forces the operator to read ``Good · 27\,\%''.

\paragraph{Data-driven baselines.}  The reference is the
30-day \emph{mean} hourly spot count per (band, UTC-hour),
regenerated daily by the \texttt{wspr-baselines.yml} GitHub
Actions workflow (a 06:00 UTC cron job that runs
\texttt{node scripts/harness.mjs wspr-baselines} and commits the
diff to \texttt{src/data/spot-baselines.mjs}, providing a true
sliding 30-day window that tracks solar-cycle progression and
seasonal shifts without manual intervention):

\begin{quote}\small\ttfamily
SELECT band, hour\_utc, sum(hourly\_count) / 30.0 AS avg\_spots\\
FROM (\\
\phantom{x}SELECT band, toStartOfHour(time) AS h,\\
\phantom{xSELECT }toHour(time) AS hour\_utc,\\
\phantom{xSELECT }count() AS hourly\_count\\
\phantom{x}FROM wspr.rx\\
\phantom{x}WHERE time >= now() - INTERVAL 30 DAY\\
\phantom{x}GROUP BY band, h, hour\_utc\\
) GROUP BY band, hour\_utc
\end{quote}

The outer aggregator divides total spots in the (band, hour)
cell by the window length in days rather than averaging only over
the hours that had at least one spot, so quiet cells (e.g.\ a
band with rare openings at a given UTC hour) get a baseline that
includes the zero-spot days rather than only the days that
opened.  The inner \texttt{GROUP BY band, h, hour\_utc} groups by
both \texttt{toStartOfHour(time)} (full date+hour timestamp) and
\texttt{toHour(time)} (hour-of-day in [0, 23]); the redundancy is
intentional, since \texttt{h} drives the per-day-per-hour
\texttt{hourly\_count} and \texttt{hour\_utc} is what the outer
group averages over to collapse the 30-day window into a 24-hour
cycle.

This produces a $10 \times 24$ matrix of typical hourly spot rates
across all days of the week.  The override fires when observed
activity exceeds $1.3 \times$ the matrix entry for the current
(band, UTC-hour) cell --- ``the band is \emph{meaningfully}
busier than usual for this time of day.''  Weekends are included
in the rolling baseline so that the live observation (which
includes weekends) is compared against a like-for-like mean;
major contest weekends are $<\!10\,\%$ of weeks in a 30-day
window so their inflation of the mean is bounded, and they
affect the override ratio in both directions (a current contest
day reads proportionally less anomalous against a slightly
inflated mean).  WSPR
hourly spot counts are over-dispersed Poisson with strong diurnal
structure: spot density depends on operator population (heavily
clustered in NA + EU, see ``Reception-population bias'' below)
and on whether the band is open at all, so the variance scales
faster than $\sqrt{\text{mean}}$ and a per-cell coefficient of
variation in the ${\sim}0.3$--$1.5$ range is typical (lower on
heavily-trafficked daytime cells, higher on quiet upper-band
predawn cells).  At that CV range the $1.3\times$ threshold
corresponds to roughly mean\,+\,$1\sigma$ on the busy cells and
mean\,+\,$0.2\sigma$ on the quiet ones; the per-cell CV
distribution itself is not separately tabulated in the paper
(see Section~\ref{sec:limits}'s deferred-empirical-claims
backlog item for the harness extension that would compute it).  Tightening to a bare $\text{spots} > \text{mean}$
comparison would fire on every above-median hour and over-promote
borderline cases; the multiplier was empirically chosen as the
smallest value that filters out routine above-average hours
without missing genuinely anomalous activity.  The choice was
made by sweeping candidate thresholds in $\{1.0, 1.1, \ldots, 2.0\}$
against the same 30-day WSPR baseline data used to construct the
matrix (\texttt{wspr.rx} table, 2026-04 window): 1.3$\times$ was
the smallest ratio at which the override fired on roughly 5--10\%
of (band, hour) cells per day --- low enough to read as
``unusual'' rather than ``daily routine'', high enough to keep
the override actionable on the cells where it matters (off-axis
sporadic-E days, evening 15\,m enhancement during HSS recovery,
upper-band cells whose 30-day-mean is structurally low for the
current solar phase but where live counts climb sharply on a
genuine opening).  Combined with the Good promotion
ceiling, the threshold keeps the override an ``unusually active''
signal across all (band, hour) cells.

The baseline matrix is emitted as a plain ES module at
\texttt{src/data/spot-baselines.mjs} and imported at runtime by
the derive pipeline.  It is regenerated daily at 06:00 UTC by
the \texttt{wspr-baselines.yml} GitHub Actions workflow (running
\texttt{node scripts/harness.mjs wspr-baselines} and committing
the diff), giving a true sliding 30-day window that tracks
solar-cycle progression and seasonal shifts without manual
intervention.  The committed-and-shipped form means fresh
checkouts get valid baselines out of the box without needing to
hit ClickHouse on first run.

\paragraph{Reception-population bias in the baseline.}  The
\texttt{wspr.rx} aggregate counts spots received by listening
stations.  WSPR receivers are heavily concentrated in North America
and Europe (operator-population statistics from
\texttt{wsprnet.org}'s active-stations roster; ${\sim}60\%$ of
worldwide WSPR receivers sit in those two regions, with another
${\sim}15\%$ in the AU/NZ/JA cluster), so a 30-day baseline of
``${\sim}100$ spots/h on 20\,m at 0900\,UTC'' reflects predominantly
North American daytime listening rather than a globally-uniform
``activity at this hour''.  The override fires uniformly across the
operator population, but the activation threshold reflects
NA/EU/Pacific listening windows.  Operators outside those regions
(AF, low-income tropical-belt countries, Central Asia) see a
baseline whose 24-hour shape is set by other regions' UTC clocks
rather than by their local-time activity rhythm; the override
correctly fires when the global aggregate runs hot, but the
``unusually active for this hour'' interpretation holds most
crisply in the receiver-dense regions.  A per-region renormalisation
is candidate work, gated on enough per-pair receiver coverage to
fit regional baselines without over-fitting noise.

% ═════════════════════════════════════════════════════════════════════
% ── Additional Figures ───────────────────────────────────────────────

% ── Noise power-sum figure ───────────────────────────────────────────
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=12cm, height=7cm,
  xlabel={UTC Hour},
  ylabel={Noise Power (\si{\dBm})},
  xmin=0, xmax=24,
  ymin=-128, ymax=-92,
  xtick={0,3,6,9,12,15,18,21,24},
  ytick={-125,-120,-115,-110,-105,-100,-95},
  grid=major,
  grid style={gray!30},
  legend pos=north east,
  legend style={font=\small},
  title={80\,m Noise Components (Suburban)},
  title style={font=\small\bfseries},
]

% 80m: Nbase = -115
% Atmospheric: Natmo = -115 + (-10 * cos(chi))
% At midnight cos(chi) ≈ -1 → Natmo = -115 + 10 = -105
% At noon cos(chi) ≈ 1 → Natmo = -115 - 10 = -125
% Using cos(chi) ≈ cos((x-12)*pi/12) for simple diurnal

\addplot[blue,thick,domain=0:24,samples=200]
  {-115 + (-10)*cos(deg((x-12)/12*180))};
\addlegendentry{Atmospheric $N_{\text{atmo}}$}

% Man-made: Nmm = -115 + 15 = -100 (flat)
\addplot[red,thick,domain=0:24,samples=2]
  {-100};
\addlegendentry{Man-made $N_{\text{mm}}$ (suburban)}

% Total: power sum
% N = 10*log10(10^(Natmo/10) + 10^(-100/10))
\addplot[black,very thick,domain=0:24,samples=200]
  {10*log10(10^((-115 + (-10)*cos(deg((x-12)/12*180)))/10) + 10^(-100/10))};
\addlegendentry{Total $N$ (power sum)}

% Annotation
\draw[<-,thick] (axis cs:12,-124) -- (axis cs:14,-126)
  node[anchor=west,font=\scriptsize] {Atmospheric minimum};
\draw[<-,thick] (axis cs:12,-100) -- (axis cs:14.5,-97)
  node[anchor=west,font=\scriptsize] {Man-made dominates};

\end{axis}
\end{tikzpicture}
\caption{Noise power components on 80\,m in a suburban environment.
         The man-made noise (\SI{-100}{\dBm}) dominates at all
         hours; the atmospheric diurnal swing barely affects the
         total.  Power-summing the two components in linear units
         (the model used here) gives the black curve, which sits
         essentially at the man-made level since
         $10^{-115/10} + 10^{-100/10} \approx 10^{-100/10}$ when
         the components differ by ${\sim}\!15$\,dB.  An earlier
         formulation that combined the two by an incorrect
         dB-domain operation produced a noise estimate
         several dB above the man-made-dominated truth on
         suburban / urban sites, inflating the margin and biasing
         verdicts open; the power-sum form is therefore a
         correctness constraint, not a stylistic choice.}
\label{fig:noise}
\end{figure}

% ── D-region absorption by band ──────────────────────────────────────
\begin{figure}[H]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=12cm, height=7cm,
  xlabel={UTC Hour},
  ylabel={$L_{\text{Dreg}}$ (\si{\dB})},
  xmin=0, xmax=24,
  ymin=0, ymax=40,
  xtick={0,3,6,9,12,15,18,21,24},
  grid=major,
  grid style={gray!30},
  legend pos=north west,
  legend style={font=\small},
  title={Quiet-Day D-Region Absorption by Band (Eq.\,\ref{eq:labsd})},
  title style={font=\small\bfseries},
]

% Post-S0-#2 paired rewrite (2026-05-07):
% LabsD(f, chi) = Lday(f) * max(0, cos(chi))^0.7
% Lday(f) = 200 / (f + 0.5)^2
% Night: cos(chi) < 0.05 → 0 (gate omitted from the smooth illustrative curve)

% 160m (1.838 MHz): Lday = 200/5.466 = 36.6
\addplot[red,very thick,domain=6:18,samples=100]
  {36.6 * (max(0, cos(deg((x-12)/12*180))))^0.7};
\addplot[red,very thick,domain=0:6,samples=20] {0};
\addplot[red,very thick,domain=18:24,samples=20] {0};
\addlegendentry{160\,m ($L_{\text{day}}=36.6$\,dB)}

% 80m (3.570 MHz): Lday = 200/16.56 = 12.1
\addplot[orange,thick,domain=6:18,samples=100]
  {12.1 * (max(0, cos(deg((x-12)/12*180))))^0.7};
\addplot[orange,thick,domain=0:6,samples=20] {0};
\addplot[orange,thick,domain=18:24,samples=20] {0};
\addlegendentry{80\,m ($L_{\text{day}}=12.1$\,dB)}

% 60m (5.366 MHz): Lday = 200/34.49 = 5.80
\addplot[yellow!80!black,thick,domain=6:18,samples=100]
  {5.80 * (max(0, cos(deg((x-12)/12*180))))^0.7};
\addplot[yellow!80!black,thick,domain=0:6,samples=20] {0};
\addplot[yellow!80!black,thick,domain=18:24,samples=20] {0};
\addlegendentry{60\,m ($L_{\text{day}}=5.80$\,dB)}

% 40m (7.040 MHz): Lday = 200/56.85 = 3.52
\addplot[green!60!black,thick,domain=6:18,samples=100]
  {3.52 * (max(0, cos(deg((x-12)/12*180))))^0.7};
\addplot[green!60!black,thick,domain=0:6,samples=20] {0};
\addplot[green!60!black,thick,domain=18:24,samples=20] {0};
\addlegendentry{40\,m ($L_{\text{day}}=3.52$\,dB)}

% 30m (10.140 MHz): Lday = 200/113.4 = 1.76
\addplot[blue,thick,domain=6:18,samples=100]
  {1.76 * (max(0, cos(deg((x-12)/12*180))))^0.7};
\addplot[blue,thick,domain=0:6,samples=20] {0};
\addplot[blue,thick,domain=18:24,samples=20] {0};
\addlegendentry{30\,m ($L_{\text{day}}=1.76$\,dB)}

% Night shading
\addplot[name path=top,draw=none,domain=0:6,samples=2] {40};
\addplot[name path=bot,draw=none,domain=0:6,samples=2] {0};
\addplot[gray!10] fill between[of=top and bot];
\addplot[name path=top2,draw=none,domain=18:24,samples=2] {40};
\addplot[name path=bot2,draw=none,domain=18:24,samples=2] {0};
\addplot[gray!10] fill between[of=top2 and bot2];

\node[font=\scriptsize,gray] at (axis cs:3,38) {Night};
\node[font=\scriptsize,gray] at (axis cs:21,38) {Night};

\end{axis}
\end{tikzpicture}
\caption{Quiet-day D-region absorption by band under the
         $L_{\text{day}}(f) \cdot \cos^{0.7}(\chi)$ form of
         Eq.~\ref{eq:labsd}.  160\,m reaches
         \SI{36.6}{\dB} at local noon; 30\,m and above are
         essentially unaffected.  Night absorption is zero.  The
         illustrative curves show a smooth descent toward zero at
         low $\cos\chi$, but the actual implementation hard-gates
         $L_{\text{Dreg}} = 0$ at $\cos\chi < 0.05$
         (Eq.~\ref{eq:labsd}); the figure smooths that gate for
         readability rather than showing the cliff.  The grayline
         bonus of \S\ref{sec:grayline} adds the
         $L_{\text{day}}(f) - L_{\text{Dreg}}(f, \chi)$ refund on
         top of the day-side charge plotted here, capped at
         \SI{25}{\dB}.}
\label{fig:dabs}
\end{figure}

% ═════════════════════════════════════════════════════════════════════
\section{Output Formatting and Alerts}\label{sec:output}

The tiers and indices produced by the foregoing pipeline are
surfaced across an eight-panel web UI:

\begin{itemize}[nosep]
  \item \textbf{Active Alerts} (\S\ref{sec:alerts}): SWPC-bulletin
        feed plus rule-derived ``soft'' alerts in one
        severity-sorted list at the top of the page.
  \item \textbf{HF Bands}: the ten HF amateur bands as a per-band
        table.  Each row shows the band frequency, the predicted
        verdict tier (Excellent / Good / Fair / Poor / Closed
        per Section~\ref{sec:tiers}), the SNR margin in dB, the
        Stability percent
        (Eq.~\ref{eq:tierstab}: $\Phi$ of $\sigma$-distance to the
        nearest tier boundary; bucket-width-independent and so
        directly readable on an absolute scale --- 50\,\% at a
        boundary, ${\sim}84$\,\% one $\sigma$ inside, $\to 100$\,\%
        deep inside), the dominant propagation mode (F2, NVIS, Es,
        TEP, F2-scatter, MS, or VHF Auroral-E
        [Section~\ref{sec:vhfaur}], displayed as ``Aurora''), and
        the best-margin destination from the
        path basket.  Distinct from circuit reliability $R$ of
        Eq.~\ref{eq:reliability}: $R$ is the probability the link
        is viable, Stability is the probability the verdict
        \emph{label} is robust to within-$\sigma$ noise.  The
        ``Best Path'' cell is a compression of the underlying
        10-path evaluation; for the per-destination breakdown
        consult the Reference Paths sub-panel in the Ionosphere
        section (cross-referenced by the cell's info popover).
  \item \textbf{VHF Bands}: 6\,m and 2\,m mirrored from the HF
        layout, with Es-MUF (\S\ref{sec:es-prop-mode}), meteor
        scatter (\S\ref{sec:ms}), and an ``aurora'' regime label (path contributes a non-zero $L_{\text{aur}}$
        from \S\ref{sec:laur} and the E-region scatter that
        accompanies auroral activity is what the path is reading
        on, rather than F2 refraction; this is a UI mode-label
        rather than a separate physics treatment).  A second
        panel below the band table renders per-station
        tropospheric ducting status from
        the 24-station UWyo radiosonde basket
        (Section~\ref{sec:tropo}): each row shows the lowest-1\,km
        $\mathrm{d}N/\mathrm{d}h$ and an ITU-R P.453 classification
        (standard / super-refractive / ducting), sorted ducting
        first so an active event surfaces at the top.
  \item \textbf{Upcoming Disruptions} (outlook): SWPC's 3-day
        R/S/G probability table, the SWPC 3-h $K_p$ forecast,
        Earth-directed CMEs from NASA DONKI, IMO-catalogue meteor
        showers, top active solar regions, and the 27-day
        ISES SFI / $A_p$ outlook.
  \item \textbf{Ionosphere}: nearest-GIRO-station digisonde
        readings (foF2 / foEs / hmF2), GNSS TEC, 30\,MHz cosmic
        noise, the Reference Paths panel --- the per-QTH path
        table listing kc2g MUF on the longest viable ring per
        bearing in the radial basket (up to 12 rows summarising
        the 72-path 12 bearings $\times$ 6 distance rings basket)
        --- plus the kc2g global foF2 map and the GloTEC TEC map.
        The Reference Paths panel is the source-of-truth that
        Section~\ref{sec:tiers}'s band-table Best-Path column
        compresses; an operator who wants per-destination detail
        rather than the headline winner consults this panel.
  \item \textbf{Geomagnetic}: $K_p$ / $A_p$ / $D_{\text{st}}$ /
        $H_{p30}$ / Sym-H / AE / aurora HP indices, a 7-day
        $K_p$ trend chart, and the OVATION aurora forecast and
        D-RAP map.
  \item \textbf{Solar}: SFI / SN / X-ray class / proton flux
        indices, real-time DSCOVR / ACE solar wind ($B_z$ /
        $V_{sw}$ / $n$), and SDO imagery (HMI visible-light
        sunspots, EUV 193\,\AA{} corona / coronal holes, EUV
        304\,\AA{} chromosphere; per
        \texttt{src/ui/sections.js}).
  \item \textbf{Links}: data source attribution, license credits,
        and links to the upstream services.
\end{itemize}

This section documents the assembly logic for the two surfaces
that depend on derived numeric thresholds (the alerts panel
soft-rule grader, and the per-band tier verdict feeding the HF /
VHF tables); the layout choices for the other panels are
data-source documentation rather than algorithm.

\subsection{Algorithmic surfaces and data-source surfaces}

Of the eight panels above, three have non-trivial assembly logic
that this section documents:
\begin{itemize}[nosep]
  \item \emph{Active Alerts} (\S\ref{sec:alerts}, below) combines
        SWPC bulletins with rule-derived ``soft'' alerts whose
        thresholds are model-side numeric constants rather than
        upstream fields.
  \item the \emph{HF Bands} and \emph{VHF Bands} tables surface
        per-band $(M, \sigma) \to$ tier verdicts already
        documented in detail in Sections~\ref{sec:snr},
        \ref{sec:reliability}, and \ref{sec:tiers}; their layout
        is just a transposition of those into table rows.
  \item the \emph{Reference Paths} sub-panel (inside the
        Ionosphere panel) computes per-QTH great-circle midpoints
        for the 72-path radial basket (12 bearings $\times$ 6
        distance rings, Section~\ref{sec:kc2gobs}), looks up the
        nearest kc2g ionosonde for each midpoint subject to a
        \SI{30}{\minute} freshness gate, and renders the per-path
        MUF with a colour-coded f/MUF column for the band
        currently selected by the band table.  The display is
        filtered to the longest viable ring per bearing
        ($\le 12$ rows) so the table stays scannable; the band
        table's per-band best-path selector still consumes the
        full 72-path basket.  This is the full-detail companion
        to the band table's compressed Best Path column
        (Section~\ref{sec:tiers}).
\end{itemize}
The remaining five panels (Upcoming Disruptions, the rest of
Ionosphere outside Reference Paths, Geomagnetic, Solar, Links)
are data-source documentation: each block
is a labelled rendering of a fetched upstream (SWPC / DONKI / IMO
/ kc2g / GIRO / SDO) without ionocast-side processing.  Their
assembly logic lives in the data fetchers and the
\texttt{src/ui/sections.js} layout description rather than in
this section; an operator-facing description of each block sits
inline in the panel's \texttt{interp} attribute and is displayed
when the panel's info badge is clicked.

\paragraph{Reading the Stability column at a glance.}  The
band-table column displays $S(M, \sigma)$ from
Eq.~\ref{eq:tierstab}: $\Phi$ of the $\sigma$-distance to the
nearest tier boundary.  Bucket-width-independent, so a centred
Fair verdict and a centred Excellent verdict at the same
$\sigma$-offset from their respective nearest boundaries read the
same percentage.  Operator reading: 50\,\% at the boundary itself,
${\sim}84$\,\% one $\sigma$ inside the bucket, asymptotic to
100\,\% deep inside; finite-width middle buckets (Poor / Fair /
Good) cap at distinct per-bucket centred ceilings,
$\Phi(0.45) \approx 67\,\%$ for Poor, $\Phi(0.32) \approx 63\,\%$
for Fair, $\Phi(0.51) \approx 70\,\%$ for Good (average
${\sim}66\,\%$, often referenced as a single round figure in
plain-prose contexts), so read mid-tier values relative to their
own bucket-specific ceiling and open-ended Closed / Excellent
values against the absolute scale.
The two-sided literal $P(\text{predicted}=\text{true})$ form
$C(M, \sigma)$ retained in source as \texttt{tierConfidence}
caps at $\sim 32\,\%$ on centred middle tiers; the production UI
(\texttt{src/ui/builders/tables.js}) surfaces $S$ instead.

\subsection{Active Alerts Panel}\label{sec:alerts}

The alerts panel combines two rendering pipelines into one
severity-sorted list.

\paragraph{Official SWPC feed.}  A polling fetch of
\texttt{products/alerts.json} emits up to four recent items.
Severity is derived from the canonical 6-letter Space Weather
Message Code embedded in the bulletin body
(e.g.~\texttt{ALTTP2}, \texttt{WARK04}); the 3-letter prefix
maps as \texttt{ALT*} / \texttt{WAR*}~$\to$~\emph{alert},
\texttt{WAT*}~$\to$~\emph{watch}, and \texttt{SUM*} /
\texttt{FOR*} / \texttt{ADV*}~$\to$~\emph{info}.  When the
message code is missing or unrecognised, the parser falls back
to the leading severity word in the body (``ALERT:'',
``WARNING:'', ``WATCH:'', ``SUMMARY:'', ``FORECAST:'',
``ADVISORY:'') with the same mapping.  Anything else lands in
\emph{info}.  The mustard \emph{warn} tier was retired during
the pill-revamp pass and is no longer emitted by either the
SWPC pipeline or the soft-alert rules.

\paragraph{Rule-derived (soft) alerts.}  Eleven rules run on
every refresh against the same data bundle the per-band budget
consumes ($K_p$, $K_p^{\text{eff}}$, $B_z$, $D_{\text{st}}$,
storm phase, aurora HP, DRAP, GOES X-ray, GOES $\ge 10$\,MeV
protons, solar-wind speed, and forecast $\sigma$).  Each rule
is \emph{graded} and short-circuits at the highest matching
tier, so $K_p = 9$ does not simultaneously fire G5, G4, G3, G2,
and G1.  Thresholds reuse the constants documented in
Sections~\ref{sec:snr}--\ref{sec:projection}; no new magic
numbers are introduced.

\paragraph{Intra-tier sort order.}  Alerts are sorted primarily
by severity tier (\emph{extreme} > \emph{alert} > \emph{watch} >
\emph{info}); within a single tier the rule iteration order is
the deterministic tiebreak: FLARE first (most time-urgent ---
X-ray ionisation peaks within seconds of the flare event,
operator response window is minutes), then geomagnetic indices
($K_p$, $B_z$, $D_{\text{st}}$ in that order), then aurora,
DRAP, PCA, storm recovery, HSS arrival, forecast~$\sigma$
inflation, and finally the geomagnetic-storm composite chip
(storm-phase summary).  The geomagnetic block is sorted by
operator familiarity rather than causal chain: $K_p$ first
because it is the most-cited storm-severity index in amateur
literature (the NOAA G-scale headlines map directly to it),
$B_z$ next as the L1 forward-warning input that leads
geomagnetic activity at Earth by ${\sim}30$--$60$\,min via
solar-wind transit ($\SI{1.5e6}{\km}/v_{\text{sw}}$), and
$D_{\text{st}}$ last as the ring-current depth signal.  The
underlying causal chain runs in the opposite direction:
$B_z \to D_{\text{st}} \to K_p$ ($\S$\ref{sec:fast-storm-drivers}
records that $D_{\text{st}}$ responds within minutes of storm
onset, 1--3\,h before the 3-hour $K_p$ index publishes).  The
fixed display order means repeated refreshes show the same chip
arrangement, so an operator scanning the panel between refreshes
can pattern-match position rather than re-read each line.

\begin{table}[H]
\centering
\caption{Soft-alert rule set.  Severity maps to the same
         \emph{info / watch / alert / extreme} levels used by
         the SWPC renderer; \emph{extreme} ranks highest.  Rule
         order matches the deterministic intra-tier tiebreak in
         the prose above.  Subscripts on labels are the NOAA
         G/R/S scale codes the rule emits as a clickable
         inline term in the chip body.}
\label{tab:soft-alerts}
\begin{tabular}{@{}l l l l@{}}
\toprule
\textbf{Trigger} & \textbf{Threshold} & \textbf{Level} & \textbf{Label} \\
\midrule
X-ray flare                  & class $\ge$ X20            & extreme & FLARE R5 \\
                             & class $\in$ X10--X19.9     & extreme & FLARE R4 \\
                             & class $\in$ X1--X9.9       & alert   & FLARE R3 \\
                             & class $\in$ M5--M9.9       & alert   & FLARE R2 \\
                             & class $\in$ M1--M4.9       & watch   & FLARE R1 \\
\midrule
Geomagnetic $K_p$            & $K_p \ge 9$                & extreme & G5 \\
                             & $K_p \ge 8$                & extreme & G4 \\
                             & $K_p \ge 7$                & alert   & G3 \\
                             & $K_p \ge 6$                & alert   & G2 \\
                             & $K_p \ge 5$                & watch   & G1 \\
\midrule
IMF $B_z$ southward          & $B_z \le -10$\,nT          & alert   & BZ SOUTH \\
                             & $B_z \le -5$\,nT           & watch   & BZ SOUTH \\
\midrule
Ring current $D_{\text{st}}$ & $D_{\text{st}} \le -250$\,nT & extreme & DST EXTREME \\
                             & $D_{\text{st}} \le -100$\,nT & alert   & DST STRONG \\
                             & $D_{\text{st}} \le -50$\,nT  & watch   & DST MODERATE \\
\midrule
Aurora HP                    & HP $\ge \SI{100}{\GW}$     & alert   & AURORA \\
                             & HP $\ge \SI{50}{\GW}$      & watch   & AURORA \\
\midrule
D-region absorption at QTH   & DRAP $\ge \SI{10}{\MHz}$   & alert   & ABSORPTION \\
                             & DRAP $\ge \SI{5}{\MHz}$    & watch   & ABSORPTION \\
\midrule
Polar cap absorption         & flux $\ge 100\,000$\,pfu   & extreme & PCA S5 \\
                             & flux $\ge 10\,000$\,pfu    & extreme & PCA S4 \\
                             & flux $\ge 1\,000$\,pfu     & alert   & PCA S3 \\
                             & flux $\ge 100$\,pfu        & alert   & PCA S2 \\
                             & flux $\ge 10$\,pfu         & watch   & PCA S1 \\
\midrule
Storm recovery tail          & $K_p^{\text{eff}} - K_p \ge 1$ \& $K_p^{\text{eff}} \ge 4$ & info & STORM TAIL \\
\midrule
HSS arrival                  & $v_{\text{sw}} \ge \SI{800}{\km/s}$       & alert & HSS \\
                             & $v_{\text{sw}} \ge \SI{600}{\km/s}$       & watch & HSS \\
\midrule
Forecast $\sigma$ inflation  & $\sigma_{\text{forecast}} \ge \SI{4}{\dB}$ & watch & FCAST \\
                             & $\sigma_{\text{forecast}} \ge \SI{2}{\dB}$ & info  & FCAST \\
\midrule
Geomagnetic-storm composite  & \texttt{stormPhase = "main"} (per \S\ref{sec:fast-storm-drivers}) & alert & GSTM \\
\bottomrule
\end{tabular}
\end{table}

Soft alerts render alongside SWPC notices rather than only when
the official feed is empty: the derived context --- southward
$B_z$, falling $D_{\text{st}}$, DRAP blanketing QTH, an HSS
front arriving --- is material information SWPC does not
publish until formal watch or warning bulletins are issued,
which may lag the underlying event by several hours.

% ═════════════════════════════════════════════════════════════════════
\section{Known Limitations and Future Work}\label{sec:limits}

The items below are the model's currently-known shortcomings and
the cleanups they imply.  They mix three classes that a reader
prioritising follow-up work may want to disentangle:
\emph{physics gaps} where the model omits a propagation mechanism
or approximates one (\#2 TX pattern, \#3 NVIS height, \#4 RX-side
noise rejection, \#5 EIA crests, \#6 flare SID decay, \#7
storm-lag kernel, \#10--\#11 equatorial / polar residuals, \#13
Es coverage, \#15 generic chordal-hop propagation outside TEP
and F2-scatter);
\emph{validation gaps} where the metric or the calibration target
has a known ceiling (\#9 binary-accuracy ceiling, \#12
$w_{\text{sc}}$ tied to a metric with a structural cap);
\emph{process / calibration cadence} items where the
infrastructure rather than the physics is the bottleneck
(\#1 in-app feedback loop not yet built, \#8 upper-band
scatter-recovery dependence as an implementation status note);
and \emph{closed items retained for calibration history}
(\#14 $N_{\text{base}}$ re-anchor, kept as the most recent
ground-truth alignment but not gating any current verdict).
The numbering is chronological by when the item entered the
backlog; readers wanting priority order should weight
\#1--\#7 (active gaps) above \#9, \#12, \#14 (calibration /
metric-ceiling items).

\begin{enumerate}[leftmargin=*]
  \item \textbf{No per-user bias correction loop yet.}  An
        on-device validation feedback loop (operator scores
        predictions, per-(band, forecast-horizon) bias and
        $\sigma$ tighten as samples accumulate) is the intended
        target shape but is not yet implemented in the live
        runtime; every operator currently sees the same global
        tune.  Until
        such a loop ships, the per-band $\sigma$ is the static
        base from Table~\ref{tab:bandsigma} (\SI{8}{\dB} on
        160\,m--40\,m, \SI{9}{\dB} on 30\,m / 20\,m / 17\,m,
        \SI{10}{\dB} on 15\,m, \SI{12}{\dB} on 12\,m / 10\,m,
        \SI{8}{\dB} on the VHF placeholders; the unconditional
        \SI{8}{\dB} fallback is reached only when the
        band-frequency lookup misses entirely, which the BANDS
        table makes impossible in production), and the offline
        harness (Section~\ref{sec:calibrate}) is the only
        feedback mechanism --- it validates the global tune
        against 30~d of WSPR aggregates but does not specialise
        to any one operator.

  \item \textbf{TX antenna pattern is an elevation-only cut.}
        The type + height + $\lambda$ model (Section~\ref{sec:antpattern})
        captures the dominant TX-side physics -- ground-reflection
        lobe position, free-space directivity, and polarisation --
        but misses second-order effects on the transmit budget:
        azimuth is not modelled (we don't know where a beam is
        pointed, so beams get their elevation-cut gain only);
        ground conductivity is fixed at ``average soil'' for all
        verticals (salt-water / sea-shore verticals are
        ${\sim}\SI{3}{\dB}$ better); the real-earth pseudo-Brewster
        null on verticals (${\sim}5$--$\SI{8}{\dB}$ dip near the
        pseudo-Brewster angle, ${\sim}\!13$--$15\degree$ elevation
        over average soil per the ARRL Antenna
        Book~\cite{arrl2023antenna} pseudo-Brewster-angle
        tabulation; the sea-water / very-good-ground angles drop
        to $3$--$8\degree$) is absent;
        and beams are treated as having the same elevation-lobe
        shape as a horizontal wire at the same height, which is
        correct for the main lobe but misses the additional
        off-broadside suppression a multi-element array provides.

  \item \textbf{NVIS secant model is single-layer.}  For paths
        $<$\,\SI{500}{\km} the model substitutes GIRO $f_{\text{oF2}}$
        with a secant correction (Eq.~\ref{eq:nvis}) anchored at the
        observed $h_{m\!F2}$ when available, defaulting to
        \SI{300}{\km}.  A more accurate NVIS model would account
        for E-/F-layer mode coexistence at low elevations and the
        full angular dependence up to ${\sim}\SI{70}{\degree}$
        off-vertical, where the simple secant overestimates the
        oblique gain.

  \item \textbf{No receive-side directional noise rejection.}  A
        physically distinct effect from \#2 above: the
        transmit-side pattern affects the radiated signal toward
        the destination; the receive-side pattern affects the
        \emph{noise} captured.  A Yagi pointed at the target
        rejects off-axis noise by its front-to-back ratio
        (\SI{5}{\dB}--\SI{15}{\dB}), which improves SNR even
        though signal-side gain on a one-way link is unchanged
        by the orientation choice.  This effective SNR improvement
        is not yet modelled.  Splitting transmit and receive
        antenna directivity would also let the budget account for
        asymmetric station geometry (e.g.\ Yagi at one end, dipole
        at the other).

  \item \textbf{Equatorial anomaly crests are approximated, not
        measured.}  The Gaussian correction in
        Eq.~\ref{eq:eia} captures the daytime crest enhancement at
        dip latitude $\pm \SI{15}{\degree}$ (+50\% of the base foF2
        at $F_{10.7\text{A}} = 70$, saturating to +85\% at
        $F_{10.7\text{A}} \ge \!120$) plus the equatorial-trough
        depression on the dip equator ($-30\%$ at $F_{10.7\text{A}}
        = 70$, saturating to $-50\%$ at $F_{10.7\text{A}} \ge 170$),
        but uses a single Gaussian width and a single crest position
        for all seasons and cycle phases.
        The true EIA migrates in latitude and varies with the
        equatorial $E\times B$ drift; a data-driven crest locator
        would be more faithful.

  \item \textbf{Flare SID decay not tracked.}  The flare absorption
        term uses the current X-ray class as an instantaneous proxy
        and therefore misses the 10--60 minute exponential decay of
        the SID after an X-ray peak.  The 10-minute model refresh
        picks up the decayed class, so the error is bounded, but
        minute-scale transitions are smeared.

  \item \textbf{Storm-lag kernel uses two coarse parameters.}  The
        Laplace-kernel-weighted effective $K_p$
        (Eq.~\ref{eq:kplag}) has $\tau_{\text{peak}}$ fixed at
        \SI{2}{\hour} and $\tau_{\text{decay}}$ switched between
        \SI{8}{\hour} (CME) and \SI{24}{\hour} (CIR / HSS) per the
        storm-type classifier of \S\ref{sec:kplag}.  Real F-region
        dynamics depend on storm intensity, latitude, and season;
        a per-storm-intensity $\tau_{\text{decay}}$ schedule and a
        latitude-dependent $\tau_{\text{peak}}$ are good targets
        for the offline calibration harness once a longer archive
        is available.

  \item \textbf{Upper-band tracking is now scatter-recovery-bound.}
        After the climatology rebuild (Section~\ref{sec:climo}), the
        H-pol Fresnel multi-hop model (Section~\ref{sec:lhop}), the
        per-band $\sigma$ refit (Table~\ref{tab:bandsigma}), and the
        F2-scatter recovery weight calibrated by joint coordinate
        descent (Section~\ref{sec:scatter}, Section~\ref{sec:calibrate}),
        the upper-band binary-accuracy lift is concentrated on
        12\,m and 10\,m, where the F2-scatter recovery channel is
        the dominant remaining lever.  Lower bands (60\,m and
        below) saturate at $>$\,98\,\% accuracy under the
        global-truth metric and are not budget-limited.  Headline
        binary-accuracy values for each calibration run are in
        Table~\ref{tab:harnessruns}; \emph{read those numbers as
        regression-detection signals, not as operator-facing
        per-path accuracy} (see \#9 below for the structural
        ceiling that the per-path column exposes).

  \item \textbf{The harness's binary-accuracy metric has a structural
        ceiling that permissive predictors clear without effort.}
        The ``open $\geq\!50$~spots/h'' target is an OR over every
        path on the band: most hours of most upper bands are ``open
        somewhere'' globally, so a predictor that errs toward ``open''
        scores well even when its per-path verdicts are wrong.  The
        sharpest evidence is the gap between the two ground-truth
        modes \emph{on the same harness run}: at the pre-B6
        2026-04-28 freeze the global aggregate read $94.25\,\%$
        accBin while the per-path mode (TX/RX-bbox restriction with
        $\geq\!1$ spot/h floor) read $30.97\,\%$ on the identical
        predictions; the post-B6 2026-05-08 baseline reads
        $84.08\,\%$ global vs $40.17\,\%$ per-path, narrower but
        still substantial.  A 40--60\,pp gap across two
        encodings of the same underlying truth is not a model
        property; it is a metric property.  The headline number is
        sitting on the ``somewhere is open'' ceiling, not on a
        per-path-correct floor.  Replacing the WSPR aggregate with
        TX-RX-pair queries (per-path ground truth) and replacing the
        climatology fallback with a real-time IRI assimilation
        (IRTAM) are the next two highest-leverage improvements; both
        require a server-side fetcher with periodic grid retrieval,
        which the current Cloudflare Pages-only architecture does
        not host.

  \item \textbf{Equatorial-belt foF2 validation has a residual
        gradient.}  The equatorial set is Boa Vista (BVJ03,
        northern crest, dip lat ${\sim}+12\degree$\,N, BR),
        Ascension (Atlantic, dip lat ${\sim}-3\degree$, deep in
        the trough region), Townsville and Niue (southern crest,
        Pacific), and Jicamarca (essentially on the dip equator,
        a second equatorial-trough anchor rather than a crest one). The
        2026-04-29 BVJ03 addition closed the one-sided crest fit;
        the EIA grid sweep (Section~\ref{sec:climo}) now bottoms
        out near $\max\!|\text{bias}| \approx \SI{1.18}{\MHz}$
        rather than the pre-BVJ03 ${\sim}\SI{2.2}{\MHz}$ floor.
        Further reduction is bounded by station density across
        the longitude-varying dip equator: Indian-sector and SE
        Asian dip-latitude stations (Trivandrum, Hyderabad,
        Madimbo, etc.\ are candidates --- none currently return
        DIDB data) or a real-time IRI assimilation are the next
        leverage points.  This residual is coupled to the polar
        over-prediction in \#11 below: both are residuals on the
        same $\mathcal{P}(\phi)\cdot d(\cos\chi)$ dayBump and
        $A_{\text{EIA}}(F_{10.7\text{A}})$ term, so a clean fix
        needs to retune them jointly.

  \item \textbf{Polar over-prediction and equatorial residual are
        coupled in the same model term.}  High-latitude stations
        Tromsø, Gakona, and Eielson AFB over-predict daytime foF2
        by ${\sim}1$--\SI{1.7}{\MHz}, and the equatorial bias
        (\#10 above) floors at ${\sim}\SI{1.18}{\MHz}$ post-BVJ03.
        These are listed as separate items because their candidate
        fixes differ (polar wants a steeper sigmoid above
        $60^\circ$, partly addressed by
        Eq.~\ref{eq:fof2-poleward}; equatorial wants additional
        Indian / SE-Asian dip-latitude stations), but structurally
        they are residuals on the same
        $\mathcal{P}(\phi) \cdot d(\cos\chi)$ dayBump and
        $A_{\text{EIA}}(F_{10.7\text{A}})$ term in
        Eq.~\ref{eq:fof2-base}: fitting the polar fall-off and
        the EIA Gaussian amplitude / width separately --- as a
        sequence of single-region tunes --- can each regress the
        other.  A clean closure needs a joint refit holding both
        residuals visible in the cost function; the
        BVJ03-extended basket makes this tractable now and is
        tracked in \texttt{docs/BACKLOG.md} under the
        latitude-gradient joint refit item.

  \item \textbf{Scatter-recovery weight $w_{\text{sc}} = 1.5$ is
        defensible-but-not-final.}  The scatter weight was set by a
        coordinate-descent sweep against the harness's binary
        ``$\geq 50$ spots/h'' WSPR target (Section~\ref{sec:scatter}
        / Section~\ref{sec:calibrate}).  An unrestricted sweep keeps
        improving as $w_{\text{sc}} \to 4$, but that range is known
        to be exploiting the global-vs-path-specific structural
        mismatch in the harness rather than a real propagation
        effect, so the shipped weight is held at the
        \SI{15}{\dB}-cap physical edge.  The paper acknowledges
        this in Section~\ref{sec:scatter}, but the limitation is
        worth surfacing as its own line in this list: the
        calibrated $w_{\text{sc}}$ has been validated against a
        metric whose calibrators do not fully trust beyond a
        certain point.  Closing the loop properly requires
        re-running the sweep against the per-pair WSPR ground
        truth (Section~\ref{sec:perpath}) once that mode is wired
        into \texttt{replayMarginFromCell}'s alternative-mode bonuses; the
        new optimum is then reported against a metric that the
        calibrators \emph{do} trust at all weights in its range.

  \item \textbf{Sporadic-E budget under-fits rare 30\,m / 40\,m
        openings by design.}  The HF Es budget
        (\texttt{snrMarginHfEs}) returns null below \SI{14}{\MHz}
        regardless of $f_{\text{oEs}}$ (Section~\ref{sec:es-prop-mode}),
        even though the algebraic Es MUF gate $f \le 5 f_{\text{oEs}}$
        would otherwise admit lower bands on strong-Es days
        ($f_{\text{oEs}} \ge \SI{4}{\MHz}$ pushes
        $5 f_{\text{oEs}} \ge \SI{20}{\MHz}$).  The hard
        \SI{14}{\MHz} floor matches amateur-radio literature:
        Es is essentially unobserved below 20\,m at the canonical
        2000\,km single-hop distance the budget assumes.  But rare
        Es openings on 30\,m / 40\,m \emph{do} appear in the WSPR
        record during peak-cycle summer evenings with
        $f_{\text{oEs}} \ge \SI{8}{\MHz}$, typically on
        shorter-than-2000\,km paths where the geometry makes lower-band
        Es viable.  These events are not captured by the budget.
        An operator who sees 30\,m WSPR activity that the model
        calls F2-only should treat the verdict as ``F2 explains
        most of it; rare Es is possible on top and the budget will
        not tell you.''  The Es budget is also \emph{single-hop
        only} ($d_{\text{Es}} = \SI{2000}{\km}$ hard-coded
        regardless of actual path length, per
        \S\ref{sec:es-prop-mode}); multi-hop Es chains and
        Es-plus-F2 mode coexistence on long paths are not modelled.
        Removing the lower-band floor and adding multi-hop
        geometry would require a distance-aware Es model and a
        recalibrated $L_{\text{iono,Es}}$ for the shorter-hop
        regime; both are deferred to a future VHF-Es-aware
        harness.

  \item \textbf{$N_{\text{base}}$ table re-anchored to P.372 quiet-rural
        (closed 2026-04-30).}  Pre-2026-04-30, $N_{\text{base}}$
        was being compared against eyeball P.372 references that
        sat below the cosmic galactic floor.  Once the
        \texttt{tests.mjs} noise-floor suite was corrected to
        evaluate $N_{\text{base}} + \Delta N_{\text{diurnal}}$ at
        midnight against properly-anchored P.372 references, the
        model came out quieter than P.372 quiet-rural midnight by
        0--11\,dB on the upper bands.  The 2026-04-30 retune ports
        the corrected anchors directly into $N_{\text{base}}$ (net
        shift: 17--10\,m all $+11$\,dB, 30\,m $+5$, 20\,m $+8$,
        80--40\,m $+1$--$2$, 160\,m $+10$, 60\,m unchanged).  WSPR
        per-spot residual mean improved from $-23.97$ to
        $-18.73$\,dB; per-pair Brier improved 0.64$\to$0.57.  The
        item is retained in this list as the most recent calibration
        ground-truth alignment; it is closed for forward work and
        does not gate any current verdict.

  \item \textbf{Generic chordal-hop propagation not modelled
        outside TEP and F2-scatter.}  ionocast explicitly models
        one chordal mode --- trans-equatorial propagation
        (\S\ref{sec:tep}, \texttt{tepBonusDb}) --- which captures
        the chordal afternoon / evening F-region mode across
        the magnetic equator that skips multiple D-region
        traversals.  The F2-scatter recovery
        (\S\ref{sec:scatter}, \texttt{scatterBonusDb}) covers
        adjacent physics: above-MUF off-axis recovery via
        F-region irregularities and short-chord modes near
        MUF.  Other chordal regimes are \emph{not} separately
        modelled: generic mid-latitude chordal hops (signals
        reflecting between F-region tilts without intermediate
        ground bounces in non-equatorial regions), polar
        chordal modes along auroral-oval gradients, and
        long-haul N--S evening chordal modes outside the
        TEP-eligibility window (cross-equator $+$ 17--23\,LT
        $+$ $|\phi_{\text{dip}}| \ge 10\degree$ on both
        endpoints).  The link budget assumes ground reflection
        between F2 hops (Fresnel $L_{\text{gr}}$ $+$ defocus
        $D$ per intermediate bounce, plus $n_{\text{hops}}
        \cdot L_{\text{iono}}$), so any chordal trajectory
        that skips a ground bounce is over-charged on the
        loss side compared to reality.  TEP and scatter
        bonuses partially compensate where their
        eligibility gates fire; on a long-haul N--S
        midlatitude path outside those gates (e.g.\ a ZL--JA
        evening 20\,m path that anecdotally reads stronger
        than the standard ground-reflection budget would
        suggest), the verdict reads slightly pessimistic.
        Adding a generic chordal-mode bonus would require a
        calibration anchor distinct from the equatorial TEP
        record --- candidate paths are sparse in the WSPR
        basket and the bonus would risk over-rewarding paths
        the standard budget already calls open --- so the
        gap is acknowledged here rather than papered over.
        Closing it cleanly is gated on an extension of the
        per-pair WSPR ground-truth wiring (\#9) plus a
        separate calibration sweep specifically for
        non-equatorial chordal candidates.
\end{enumerate}

% ═════════════════════════════════════════════════════════════════════
\appendix

\section{Default Settings}\label{app:defaults}

\begin{table}[H]
\centering
\caption{Default operator settings.}
\label{tab:defaults}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Parameter} & \textbf{Default Value} & \textbf{Notes} \\
\midrule
Transmit power $P_{\text{tx}}$ & \SI{50}{\dBm} (\SI{100}{\W}) & Typical HF transceiver \\
Antenna gain $G_{\text{ant}}$ (peak) & \SI{+5}{\dBi} & Installed dipole w/ ground reflection; effective gain at the path's takeoff angle is $G_{\text{peak}} + G_{\text{rel}}(\theta_{\text{elev}})$, $G_{\text{rel}} \in [-15, 0]$\,dB (\S\ref{sec:antpattern}) \\
Mode & SSB & $S_{\text{req}} = +10$\,dB \\
Noise environment & Suburban & $F_a = +15$\,dB \\
Reference bandwidth & \SI{2.5}{\kHz} & SSB passband \\
Spread $\sigma_g$ (base) & 8--12\,\si{\dB} per band & ITU-R P.533 anchor, not fitted (Table~\ref{tab:bandsigma}) \\
kc2g freshness gate & \SI{30}{\minute} & Stale data discarded \\
Hop distance ceiling & \SI{4000}{\km} (at $h_F = \SI{300}{\km}$) & The runtime ceiling is $h_F$-dependent per Eq.~\ref{eq:hopceiling}; the \SI{4000}{\km} value is the typical $h_F = \SI{300}{\km}$ point used in worked examples \\
$L_{\text{iono,HF}}$ & \SI{1}{\dB} per F2 reflection & Lumped median-conditions residual (\S\ref{sec:liono}; \texttt{L\_IONO\_HF\_DB}) \\
$L_{\text{iono,Es}}$ & \SI{15}{\dB} per Es hop & Patchy / aspect / polarisation aggregate (\S\ref{sec:es-prop-mode}; \texttt{L\_IONO\_ES\_DB}) \\
$L_{\text{iono,aur}}$ & \SI{25}{\dB} per VHF auroral-E hop & Empirical lump (\S\ref{sec:vhfaur}; \texttt{L\_IONO\_AUR\_DB}) \\
$D$ (defocus per extra hop) & \SI{0.25}{\dB} & Calibrated against 30-day WSPR (\S\ref{sec:lhop}; \texttt{DEFOCUS\_DB\_PER\_EXTRA\_HOP}) \\
D-region prefactor $K_D$ & $200\,\si{\dB}\cdot\si{\MHz}^{2}$ & $L_{\text{day}}(f) = K_D / (f + 0.5)^{2}$ (Eq.~\ref{eq:labsd}; \texttt{D\_REGION\_PREFACTOR}) \\
\bottomrule
\end{tabular}
\end{table}

\section{Noise Floor Table}\label{app:noise}

The reference noise data feeding the noise model
(Section~\ref{sec:noise}) is in Table~\ref{tab:noisebase} of the
main text.  The operator's noise environment selection determines
how the man-made channel of Eq.~\ref{eq:noise} fires: rural sites
gate the channel off entirely (the indicator
$\mathbb{1}_{\text{env}\ne\text{rural}}$ is zero); suburban sites
add the man-made channel at $N_{\text{base}} + \SI{15}{\dB}$; urban
at $N_{\text{base}} + \SI{25}{\dB}$.  $F_a$ is a single additive
term, so a per-environment expansion of Table~\ref{tab:noisebase}
is unnecessary; the reader can apply the offset mentally.

\section{Calibration Path Basket}\label{app:paths}

The 35-path basket used by the offline calibration harness
(\texttt{scripts/harness.mjs}, Section~\ref{sec:calibrate}).
Each row is a great-circle TX$\to$RX pair specified by endpoint
coordinates; the basket is intentionally diverse across hop count
(1 to 5 hops, NVIS through long-path), midpoint geometry
(near-station vs ocean), and latitude regime (midlatitude / polar /
equatorial / transequatorial).  The same basket is the input to the
sweep that produced the binary-accuracy figure quoted in
Section~\ref{sec:tier-confidence}.

\begin{table}[H]
\centering
\small
\caption{The 35 calibration reference paths.  Columns: path name,
TX endpoint $(\phi, \lambda)$ in degrees, RX endpoint $(\phi, \lambda)$
in degrees, and the geometry / use-case note from
\texttt{scripts/paths.json}.}
\label{tab:paths}
\begin{tabular}{@{}lrrrrl@{}}
\toprule
\textbf{Name} & \textbf{TX $\phi$} & \textbf{TX $\lambda$}
              & \textbf{RX $\phi$} & \textbf{RX $\lambda$}
              & \textbf{Geometry / use case} \\
\midrule
EU--EU short      &  38.72 &   -9.14 &  50.45 &   30.52 & 3000\,km midlat 1-hop \\
EU--EU west       &  55.95 &   -3.20 &  37.40 &   -5.99 & 2400\,km midlat 1-hop \\
EU--EU east       &  60.17 &   24.94 &  37.97 &   23.73 & 2500\,km midlat 1-hop \\
EU--NA mid        &  38.72 &   -9.14 &  40.71 &  -74.01 & 5400\,km transatlantic 2-hop \\
EU--NA north      &  55.95 &   -3.20 &  45.51 &  -73.57 & 5100\,km transatlantic 2-hop \\
EU--NA south      &  37.40 &   -5.99 &  25.76 &  -80.19 & 7100\,km transatlantic 2-hop \\
NA--EU east       &  40.71 &  -74.01 &  50.45 &   30.52 & 7600\,km transatlantic 2-hop \\
NA--NA west-east  &  37.77 & -122.42 &  40.71 &  -74.01 & 4100\,km transcontinental 2-hop \\
EU--Asia          &  38.72 &   -9.14 &  35.68 &  139.69 & 11144\,km 3-hop, midpoint Siberia \\
NA--Asia          &  37.77 & -122.42 &  35.68 &  139.69 & 8300\,km transpacific 3-hop \\
Hawaii--VK        &  21.31 & -157.86 & -33.87 &  151.21 & 8200\,km transpacific equator-crossing 3-hop \\
Hawaii--NA        &  21.31 & -157.86 &  37.77 & -122.42 & 3900\,km midlat 1-hop, midpoint Pacific \\
Hawaii--JA        &  21.31 & -157.86 &  35.68 &  139.69 & 6200\,km midpoint Pacific 2-hop \\
EU--South Africa  &  38.72 &   -9.14 & -26.20 &   28.04 & 8500\,km TEP-eligible 3-hop \\
NA--Brazil        &  40.71 &  -74.01 & -23.55 &  -46.63 & 7700\,km TEP-eligible 2-hop \\
EU--Brazil        &  38.72 &   -9.14 & -23.55 &  -46.63 & 8700\,km TEP-eligible 3-hop \\
VK--EU            & -33.87 &  151.21 &  50.45 &   30.52 & 14944\,km Asia-transit 5-hop, midpoint SE Asia \\
VK--EU west       & -33.87 &  151.21 &  38.72 &   -9.14 & 18200\,km 5-hop, midpoint Indian Ocean \\
VK--NA            & -33.87 &  151.21 &  37.77 & -122.42 & 11900\,km transpacific 4-hop \\
JA--VK            &  35.68 &  139.69 & -33.87 &  151.21 & 7800\,km 2-hop, equator-crossing \\
JA--EU            &  35.68 &  139.69 &  50.45 &   30.52 & 7800\,km 2-hop, midpoint Siberia \\
NA--Argentina     &  40.71 &  -74.01 & -34.61 &  -58.38 & 8500\,km TEP-eligible \\
EU--Argentina     &  38.72 &   -9.14 & -34.61 &  -58.38 & 10100\,km 3-hop \\
Polar W--UA       &  40.71 &  -74.01 &  55.75 &   37.62 & 7500\,km via near-pole 2-hop \\
Polar W6--EU      &  37.77 & -122.42 &  50.45 &   30.52 & 9700\,km polar 3-hop \\
Polar W--NA       &  40.71 &  -74.01 &  37.77 & -122.42 & 4100\,km, hop-count consistency check \\
Equator NA        &  21.31 & -157.86 & -23.55 &  -46.63 & 13013\,km equator-crossing 4-hop \\
Asia--VK          &  22.32 &  114.17 & -33.87 &  151.21 & 7400\,km 2-hop \\
EU--Asia short    &  55.75 &   37.62 &  35.68 &  139.69 & 7500\,km 2-hop, midpoint Siberia/China \\
NVIS DE--NL       &  52.52 &   13.41 &  52.37 &    4.90 & 600\,km NVIS-tail short \\
NVIS GB--FR       &  51.51 &   -0.13 &  48.85 &    2.35 & 350\,km true NVIS short \\
NVIS US--CA       &  40.71 &  -74.01 &  45.42 &  -75.70 & 530\,km NVIS-tail \\
NA--IT            &  40.71 &  -74.01 &  41.90 &   12.50 & 6900\,km transatlantic midlat \\
EU--Mid-East      &  50.45 &   30.52 &  25.20 &   55.27 & 4200\,km 2-hop midlat-equatorial \\
ZL--W6            & -41.29 &  174.78 &  37.77 & -122.42 & 11200\,km transpacific 4-hop \\
\bottomrule
\end{tabular}
\end{table}

Endpoint coordinates are read by the harness from
\texttt{scripts/data/paths.json} (the source of truth); this
table is a paper-side mirror.  ``Polar W--NA'' is an intentional
duplicate of NA--NA west-east, retained as a hop-count regression
canary, so the basket has 35 entries but only 34 unique
(TX, RX) pairs.

\section{GIRO Digisonde Stations}\label{app:giro}

The 27 ionosonde stations queried by the harness for foF2 / foEs
ground-truth.  The list lives in two files that must stay in sync:
\texttt{scripts/harness.mjs}'s \texttt{GIRO\_STATIONS} (URSI code
plus coordinates, used by the offline calibration harness) and
\texttt{functions/\_handlers/giro.js}'s \texttt{GIRO\_STATIONS} (URSI
code, display name, coordinates, and provider attribution, used by
the live UI).  Coordinates are verified against the kc2g station
registry by \texttt{scripts/harness.mjs verify}.

\begin{table}[H]
\centering
\small
\caption{The 27 GIRO digisonde stations used for foF2 / foEs
ground-truth and the EIA grid sweep.  Coordinates are
$(\phi, \lambda)$ in degrees (East $> 0$, West $< 0$).  Dip
latitude $\phi_{\text{dip}}$ is computed via the tilted-dipole
approximation of Eq.~\ref{eq:cgm} with the IGRF-13 north magnetic
pole at $(80.7^\circ\text{N},\,72.7^\circ\text{W})$; positive
values are in the northern magnetic hemisphere.  Region groupings
follow the harness's structural blocks; ``Provider'' is the
institution operating the digisonde feed via the GIRO / DIDB
network.}
\label{tab:giro}
\begin{tabular}{@{}llrrrl@{}}
\toprule
\textbf{URSI} & \textbf{Name} & \textbf{$\phi$} & \textbf{$\lambda$}
              & \textbf{$\phi_{\text{dip}}$} & \textbf{Provider} \\
\midrule
\multicolumn{6}{@{}l}{\textit{Europe / midlatitude (10)}} \\
PQ052 & Pruhonice, CZ        &  50.0 &   14.6 & 49.5 & Inst.\ Atmospheric Physics, Czech Acad.\ Sci. \\
JR055 & Juliusruh, DE        &  54.6 &   13.4 & 54.2 & IAP Kühlungsborn \\
EB040 & Roquetes, ES         &  40.8 &    0.5 & 42.9 & Observatori de l'Ebre \\
FF051 & Fairford, GB         &  51.7 &   -1.5 & 53.8 & Rutherford Appleton Lab / STFC \\
SO148 & Sopron, HU           &  47.6 &   16.7 & 46.9 & Geodetic and Geophysical Inst., HAS \\
AT138 & Athens, GR           &  38.0 &   23.5 & 36.4 & National Observatory of Athens \\
RO041 & Rome, IT             &  41.8 &   12.5 & 41.9 & INGV Rome \\
GM037 & Gibilmanna, IT       &  37.9 &   14.0 & 37.8 & INGV Gibilmanna \\
NI135 & Nicosia, CY          &  35.0 &   33.2 & 32.0 & GIRO / DIDB \\
DB049 & Dourbes, BE          &  50.1 &    4.6 & 51.2 & Royal Meteorological Inst.\ of Belgium \\
\midrule
\multicolumn{6}{@{}l}{\textit{North America (4)}} \\
MHJ45 & Millstone Hill, MA   &  42.6 &  -71.5 & 51.9 & MIT Haystack Observatory \\
BC840 & Boulder, CO          &  40.0 & -105.3 & 47.6 & NOAA / NCEI Boulder \\
WP937 & Wallops Island, VA   &  37.9 &  -75.5 & 47.2 & NASA Wallops Flight Facility \\
IF843 & Idaho Natl Lab, ID   &  43.8 & -112.7 & 50.6 & Idaho National Laboratory \\
\midrule
\multicolumn{6}{@{}l}{\textit{Polar / high-latitude (3)}} \\
TR169 & Tromsø, NO           &  69.6 &   19.2 & 67.4 & EISCAT / UiT Tromsø \\
GA762 & Gakona, AK           &  62.4 & -145.0 & 63.8 & HAARP / Gakona \\
EI764 & Eielson AFB, AK      &  64.7 & -147.1 & 65.6 & GIRO / DIDB \\
\midrule
\multicolumn{6}{@{}l}{\textit{Equatorial / low-latitude (3)}} \\
AS00Q & Ascension Island     &  -7.9 &  -14.4 & $-3.0$ & UK Ionospheric Monitoring Service \\
BVJ03 & Boa Vista, BR        &   2.8 &  -60.7 & 11.9   & Instituto Nacional de Pesquisas Espaciais (INPE) \\
JI91J & Jicamarca, PE        & -11.9 &  -76.9 & $-2.6$ & Radio Observatorio de Jicamarca / IGP \\
\midrule
\multicolumn{6}{@{}l}{\textit{Africa / southern midlatitude (1)}} \\
LV12P & Louisvale, ZA        & -28.5 &   21.2 & $-28.7$ & SANSA \\
\midrule
\multicolumn{6}{@{}l}{\textit{South America / southern midlatitude (1)}} \\
CN53M & Cachoeira Paulista, BR & -22.7 & -45.0 & $-14.4$ & INPE \\
\midrule
\multicolumn{6}{@{}l}{\textit{Australasia / Pacific (5)}} \\
TV51R & Townsville, AU       & -19.6 &  146.8 & $-26.6$ & GIRO / DIDB \\
ND61R & Niue                 & -19.1 & -169.9 & $-20.0$ & GIRO / DIDB \\
PE43K & Perth, AU            & -32.0 &  116.1 & $-41.2$ & GIRO / DIDB \\
CB53N & Canberra, AU         & -35.3 &  149.0 & $-42.0$ & GIRO / DIDB \\
HO54K & Hobart, TAS          & -42.9 &  147.3 & $-49.7$ & GIRO / DIDB \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Coverage gaps.}  Boa Vista (BVJ03) anchors the
northern equatorial-anomaly crest at dip latitude
${\sim}+12\degree$\,N (Brazilian sector, where the South
Atlantic Anomaly shifts the magnetic equator south of its
geographic counterpart, lifting the dip latitude of geographic-
equatorial stations onto the northern magnetic flank).  Indian-
sector and SE-Asian dip-latitude stations near
$\phi_{\text{dip}} \approx +15\degree$\,N (Trivandrum,
Hyderabad, Madimbo, etc.) would close the longitude-coverage
gap but do not currently return DIDB data.  The southern
hemisphere is well-covered (Townsville, Niue, Perth, Canberra,
Hobart for the Pacific block), but Africa is sparse outside
SANSA's Louisvale, and Asia / Indian Ocean coverage is
essentially absent.  This shapes the residual EIA-amplitude
bias documented in Section~\ref{sec:limits}.

\section{Reproducibility Manifest}\label{app:repro}

The harness-fitted constants $\{L_{\text{iono}}, D, w_{\text{sc}},
w_{\text{NVIS-tail}}, w_{\text{Es-prim}}, \sigma\text{-scale},
\text{fusion-flag}\}$ were produced by joint coordinate descent
(Section~\ref{sec:calibrate},
\texttt{scripts/tests/tune-r7.mjs}) starting
from three deterministic seed configurations chosen to enter the
parameter space from distinct basins.  This appendix lists those
seeds, the per-parameter sweep grid, and the on-disk artefacts
needed to reproduce the harness-fitted numbers end-to-end.

\paragraph{Three deterministic starting configurations.}

\begin{table}[H]
\centering
\small
\caption{Coordinate-descent starting configurations.  All three
seeds run the same descent over the grid below; the winner is the
seed whose final state has the lowest binary Brier loss against
the WSPR ``alive somewhere'' truth.  ``$L_{\text{iono}}$'' here is
the per-hop HF lumped ionospheric loss in dB;
``defocusDbPerExtraHop'' is the $D$ constant of
Section~\ref{sec:lhop}; the four mode weights are the dimensionless
multipliers controlling whether F2-scatter / NVIS-tail blending /
Sporadic-E-as-mode contributions fire; ``fusionEnabled'' is the
per-hop foF2 fusion flag (deferred path of
Section~\ref{sec:scatter}); ``sigmaScale'' uniformly scales the
per-band $\sigma_g$ table (Table~\ref{tab:bandsigma}) to expose
whether the harness wants a wider or tighter prediction
spread.}
\label{tab:seeds}
\begin{tabular}{@{}lccccccc@{}}
\toprule
\textbf{Seed} & \textbf{$L_{\text{iono}}$} & \textbf{$D$} &
  \textbf{fusion} & \textbf{$w_{\text{sc}}$} &
  \textbf{$w_{\text{NVIS}}$} & \textbf{$w_{\text{Es}}$} &
  \textbf{$\sigma$-scale} \\
\midrule
baseline   & 1 & 0.25 & off & 0   & 0   & 0   & 1.0 \\
fusion-up  & 4 & 0.25 & on  & 0   & 0   & 0   & 1.0 \\
modes-on   & 1 & 0.25 & on  & 1.5 & 1.0 & 1.0 & 1.3 \\
\bottomrule
\end{tabular}
\end{table}

The three names encode the experimental hypothesis each seed
tests: \emph{baseline} freezes alternative-mode bonuses to zero
and asks whether the F2-only budget alone is sufficient;
\emph{fusion-up} asks whether enabling per-hop foF2 fusion plus a
deeper $L_{\text{iono}}$ shifts the Brier optimum; \emph{modes-on}
asks whether all alternative-propagation bonuses turned on at
moderate weights, plus a wider $\sigma$, finds a different basin.
All three converged to the same basin within ${\le}2$ iterations
(Section~\ref{sec:calibrate}), final
$\{L_{\text{iono}}, D, w_{\text{sc}}\}$ within $\pm 10\,\%$ across
seeds.

\paragraph{Per-parameter sweep grid.}  Each coordinate-descent step
sweeps one parameter through its grid below while holding the
others fixed at their current best values.  Grids were chosen wide
enough to cover both physical-edge values and obvious-overshoot
values; the descent converges inside the physical region.

\begin{table}[H]
\centering
\small
\caption{Coordinate-descent sweep grid
(\texttt{scripts/tests/tune-r7.mjs}'s \texttt{GRID}).}
\label{tab:tunegrid}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Parameter} & \textbf{Sweep values} \\
\midrule
$L_{\text{iono}}$ (\texttt{lIonoHfDb})           & 0, 0.5, 1, 2, 4, 6, 8 \\
$D$ (\texttt{defocusDbPerExtraHop})              & 0, 0.25, 0.5, 1.0 \\
\texttt{fusionEnabled}                            & false, true \\
$w_{\text{sc}}$ (\texttt{scatterWeight})          & 0, 0.5, 1.0, 1.5 \\
$w_{\text{NVIS-tail}}$ (\texttt{nvisTailWeight}) & 0, 0.5, 1.0, 1.5 \\
$w_{\text{Es-prim}}$ (\texttt{esWeight})          & 0, 0.5, 1.0, 1.5 \\
$\sigma$-scale (\texttt{sigmaScale})              & 0.7, 1.0, 1.3, 1.6, 2.0 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{On-disk calibration artefacts.}  The harness writes
two baseline files and one report file; together with the source
configurations above and the calibrated-constants file
\texttt{src/constants.js} they specify the calibration completely.

\begin{description}[nosep]
  \item[\texttt{src/constants.js}.]  The single source of truth
    for every harness-fitted dB constant
    ($L_{\text{iono}}$, $D$, $w_{\text{sc}}$, $w_{\text{NVIS-tail}}$,
    $w_{\text{Es-prim}}$, $\sigma\text{-scale}$, $\text{fusion-flag}$
    above plus EIA amplitudes / TEP plateau coefficients / Es
    persistence kernel constants / Bz bump table / storm-floor
    schedule).  Hand-edited after the harness reports a winner;
    every value in the budget traces to this file.
  \item[\texttt{scripts/data/harness.baseline.json}.]  Global
    ``open somewhere'' baseline computed from a single
    bbox-unrestricted WSPR aggregate query (cheap, ${\sim}\!7\,k$
    rows for a 30-day window).  Used as the default truth for the
    Brier / accBin numbers in the headline metrics
    (Table~\ref{tab:harnessruns}).
  \item[\texttt{scripts/data/harness.baseline.perpath.json}.]
    Per-path baseline using TX/RX bounding-box restricted aggregates
    (35 queries per harness run, $\geq 1$ spot/h floor).  Adds a
    path-specific drift signal that the global mode averages out.
  \item[\texttt{scripts/outputs/harness.report.json}.]  Latest harness
    run output.  Regenerated each \texttt{node scripts/harness.mjs}
    invocation; \texttt{scripts/outputs/} is gitignored so the
    file is not committed to source control.
  \item[\texttt{src/data/spot-baselines.mjs}.]  Per-(band, UTC-hour)
    30-day WSPR spot-count baseline used by the spot-override
    surface (Section~\ref{sec:spot-override}) and the operator
    Comp.\ availability prior of Eq.~\ref{eq:completion}.
    Regenerated daily at 06:00 UTC by the
    \texttt{wspr-baselines.yml} GitHub Actions workflow, which
    runs \texttt{node scripts/harness.mjs wspr-baselines} and
    commits the diff; this gives a true sliding 30-day window
    that tracks solar-cycle progression and seasonal shifts
    without manual intervention.
\end{description}

\paragraph{Auxiliary test, tune, and diagnostic suites.}  Three
script families live alongside the harness in
\texttt{scripts/tests/}:

\begin{description}[nosep]
  \item[Tune suites.] Coordinate-descent or grid-sweep scripts
    that the production calibration pulls from.
    \texttt{tune-r7.mjs} runs the seven-parameter sweep
    described above; \texttt{tune-eia.mjs} sweeps the EIA crest
    amplitude / width / latitude triplet against the equatorial
    GIRO basket (\S\ref{sec:climo}); \texttt{tune-blend.mjs}
    sweeps the legacy ensemble-blend weights as a regression
    canary against the retired N0NBH-style heuristicTier
    (\S\ref{sec:ensemble}).  All three write report cards to
    stdout rather than mutating committed files.
  \item[Validation suites.] Cross-checks against external
    sources: \texttt{wspr-snr.mjs} compares per-spot WSPR SNR
    to the model's predicted budget on the same path / hour;
    \texttt{psk.mjs} pulls PSKReporter spot density as an
    independent activity signal; \texttt{rbn.mjs} and
    \texttt{rbn-beacon.mjs} use Reverse Beacon Network spot data
    against CW skimmer references.  These cover decode-mode
    truth that WSPR-only validation misses.
  \item[Diagnostic suites.] Pre-fix investigation tooling
    retained for regression detection: \texttt{storm-split.mjs}
    bins per-path margin error by storm phase (CME-onset / main
    / recovery / quiet); \texttt{day-night.mjs} bins by sunlit
    fraction of the path; \texttt{hops.mjs} bins by integer hop
    count.  These produce per-bin Brier / accBin breakdowns and
    are the source of the band-distribution numbers cited in
    \S\ref{sec:limits}~\#8.
\end{description}

\paragraph{What this paper documents and what it does not.}  Every
constant in the SNR budget (Section~\ref{sec:snr}) and MUF
estimation (Section~\ref{sec:muf}) is given inline.  The
$\sigma_g$ table is in Table~\ref{tab:bandsigma}; the noise-floor
table in Table~\ref{tab:noisebase}; the full path basket in
Appendix~\ref{app:paths}; the GIRO station list in
Appendix~\ref{app:giro}; the $\Phi$ approximation in
Section~\ref{sec:reliability}; the IGRF pole and
EIA / winter / equatorial / polar amplitudes inline at their
respective equations.  What this paper does \emph{not} reproduce
is (a) the WSPR ground-truth aggregates themselves --- those are
queried live from \texttt{wspr.live} at harness-run time and
their values shift week-to-week as activity rebalances; and
(b) the SWPC, GIRO, GFZ, and Kyoto data feeds, which are also
live.  Reproducing the headline accBin / Brier numbers in
Table~\ref{tab:harnessruns} therefore requires running the harness
against contemporaneous data from those feeds, which means the
absolute numbers will differ slightly run-to-run; the regression
detection signal (drift $> \SI{2}{\dB}$ mean margin or 5\,pp in
$P(\text{open})$ across runs) is the reproducibility guarantee,
not bit-exact metric values.

% ═════════════════════════════════════════════════════════════════════
% \begin{thebibliography} auto-renders its own "References" h2 in the
% article class (via \refname); only addcontentsline is needed for TOC.
\addcontentsline{toc}{section}{References}\label{sec:refs}

\begin{thebibliography}{99}

\bibitem{itup533}
  ITU-R Recommendation P.533-14,
  \emph{Method for the Prediction of the Performance of HF Circuits},
  International Telecommunication Union, Geneva, 2019.
  \url{https://www.itu.int/rec/R-REC-P.533/}.

\bibitem{itup372}
  ITU-R Recommendation P.372-15,
  \emph{Radio Noise},
  International Telecommunication Union, Geneva, 2021.
  \url{https://www.itu.int/rec/R-REC-P.372/}.
  The 2026-04-30 noise-floor re-derivation in
  Section~\ref{sec:noise} (Table~\ref{tab:noisebase}) was done
  against this revision; figure references throughout the noise
  prose track P.372-15 numbering, which differs in places from
  the prior P.372-14 numbering (notably Fig.~38 vs Fig.~37 for
  the quiet-rural midlat reference).

\bibitem{itup1239}
  ITU-R Recommendation P.1239-3,
  \emph{ITU-R Reference Ionospheric Characteristics},
  International Telecommunication Union, Geneva, 2012.
  \url{https://www.itu.int/rec/R-REC-P.1239/}.

\bibitem{itup525}
  ITU-R Recommendation P.525-4,
  \emph{Calculation of Free-Space Attenuation},
  International Telecommunication Union, Geneva, 2019.
  \url{https://www.itu.int/rec/R-REC-P.525/}.

\bibitem{itup534}
  ITU-R Recommendation P.534-6,
  \emph{Method for Calculating Sporadic-E Field Strength},
  International Telecommunication Union, Geneva, 2019.
  \url{https://www.itu.int/rec/R-REC-P.534/}.

\bibitem{itup453}
  ITU-R Recommendation P.453-14,
  \emph{The Radio Refractive Index: Its Formula and Refractivity Data},
  International Telecommunication Union, Geneva, 2019.
  \url{https://www.itu.int/rec/R-REC-P.453/}.

\bibitem{itup527}
  ITU-R Recommendation P.527-6,
  \emph{Electrical Characteristics of the Surface of the Earth},
  International Telecommunication Union, Geneva, 2021.
  \url{https://www.itu.int/rec/R-REC-P.527/}.

\bibitem{itup842}
  ITU-R Recommendation P.842-5,
  \emph{Computation of Reliability and Compatibility of HF Radio Systems},
  International Telecommunication Union, Geneva, 2013.
  \url{https://www.itu.int/rec/R-REC-P.842/}.

\bibitem{itup834}
  ITU-R Recommendation P.834-9,
  \emph{Effects of Tropospheric Refraction on Radiowave Propagation},
  International Telecommunication Union, Geneva, 2017.
  Cited for the modified-refractivity duct-intensity definition
  $M_{\text{top}} - M_{\text{min}}$ in \S\ref{sec:tropomap}.
  \url{https://www.itu.int/rec/R-REC-P.834/}.

\bibitem{noaagfs}
  NOAA National Centers for Environmental Prediction (NCEP),
  \emph{Global Forecast System (GFS) Operational Documentation},
  served via the NOAA Operational Model Archive and Distribution
  System (NOMADS), College Park, Maryland.  ionocast pulls a
  server-side-filtered subset of the 0.25\textdegree{}
  \texttt{pgrb2} product (13 pressure levels plus surface bundle)
  from the NOMADS subset endpoint.
  \url{https://nomads.ncep.noaa.gov/cgi-bin/filter_gfs_0p25.pl};
  product description at
  \url{https://www.nco.ncep.noaa.gov/pmb/products/gfs/}.

\bibitem{hepburntropo}
  William R.\ Hepburn,
  \emph{Worldwide VHF/UHF Tropospheric Forecasts},
  DX Information Centre, dxinfocentre.com.  Long-running
  amateur-radio reference for the Hepburn-class banded-color
  tropospheric forecast aesthetic that ionocast's heatmap matches
  visually.
  \url{https://www.dxinfocentre.com/tropo.html}.

\bibitem{d3contour}
  Mike Bostock and contributors,
  \emph{d3-contour: Compute contour polygons using marching
  squares},
  ISC license, 2017--present.  Used by the renderer to extract
  per-band fill polygons from the smoothed grid; vendored locally
  alongside d3-array under \texttt{src/tropo/vendor/}.
  \url{https://github.com/d3/d3-contour}.

\bibitem{arrl2023antenna}
  ARRL,
  \emph{The ARRL Antenna Book}, 25th edition,
  American Radio Relay League, Newington, CT, 2023.
  Cited for ground-reflection factor and elevation-pattern
  derivations in \S\ref{sec:antpattern} (Chapter~3).
  \url{https://www.arrl.org/shop/}.

\bibitem{arrl2023handbook}
  ARRL,
  \emph{The ARRL Handbook for Radio Communications},
  centenary (100th) edition,
  American Radio Relay League, Newington, CT, 2022/2023.
  Cited for noise-environment classification background
  alongside ITU-R~P.372~\cite{itup372} and for the
  comfortable-copy SSB / trained-ear CW $S_{\text{req}}$
  conventions used in Table~\ref{tab:modesnr}.
  \url{https://www.arrl.org/shop/}.

\bibitem{k9la}
  C.~Luetzelschwab (K9LA),
  \emph{Propagation Tutorials and Monthly-Column Archive},
  k9la.us, 2008--2024.
  Cited as background for D-region absorption magnitude.  The
  prior pre-2026-05-07 per-band $A_{\text{base}}$ table was
  operator-experience consensus informed by this archive; that
  table has been retired in favour of the closed-form
  $L_{\text{day}}(f) = K_D / (f + 0.5)^{2}$ of
  Eq.~\ref{eq:labsd}, which tracks the ITU-R P.533~\cite{itup533}
  Annex 1 non-deviative-absorption shape directly.  K9LA remains
  the calibration cross-check for the magnitudes the new closed
  form produces on 160--30\,m amateur paths.
  \url{https://www.k9la.us/}.

\bibitem{k1jt}
  J.\,H.~Taylor (K1JT),
  ``WSJT-X: Digital Modes for Weak Signal Communications,''
  \url{https://wsjt.sourceforge.io/wsjtx.html}.

\bibitem{fullerrowell1994}
  T.\,J.~Fuller-Rowell, M.\,V.~Codrescu, R.\,G.~Roble, and
  A.\,D.~Richmond,
  ``How does the thermosphere and ionosphere react to a geomagnetic
  storm?''
  in \emph{Magnetic Storms}, Geophysical Monograph~98, pp.~203--225,
  American Geophysical Union, 1994.

\bibitem{mendillo2006}
  M.~Mendillo,
  ``Storms in the ionosphere: Patterns and processes for total
  electron content,''
  \emph{Reviews of Geophysics}, vol.~44, RG4001, 2006.
  \textsc{doi}: 10.1029/2005RG000193.

\bibitem{buonsanto1999}
  M.\,J.~Buonsanto,
  ``Ionospheric Storms --- A Review,''
  \emph{Space Science Reviews}, vol.~88, no.~3--4,
  pp.~563--601, 1999.
  \textsc{doi}: 10.1023/A:1005107532631.
  Cited for the F-region positive-phase magnitude
  (15--50\,\% above quiet-time baseline) and duration
  (1--4\,h) summarised in \S\ref{sec:stormdepression}.

\bibitem{prolss1995}
  G.\,W.~Pr\"olss,
  ``Ionospheric F-region storms,''
  in \emph{Handbook of Atmospheric Electrodynamics}, vol.~2,
  pp.~195--248, CRC Press, 1995.

\bibitem{sauer2008}
  H.\,H.~Sauer and D.\,C.~Wilkinson,
  ``Global mapping of ionospheric HF/VHF radio wave absorption due to
  solar energetic protons,''
  \emph{Space Weather}, vol.~6, S12002, 2008.
  \textsc{doi}: 10.1029/2008SW000399.

\bibitem{anderson1981}
  D.\,N.~Anderson,
  ``Modeling the ambient, low latitude F-region ionosphere~--
  a review,''
  \emph{Journal of Atmospheric and Terrestrial Physics},
  vol.~43, no.~8, pp.~753--762, 1981.

\bibitem{mitra1974}
  A.\,P.~Mitra,
  \emph{Ionospheric Effects of Solar Flares},
  D.~Reidel Publishing, Dordrecht, 1974.

\bibitem{beandutton1968}
  B.\,R.~Bean and E.\,J.~Dutton,
  \emph{Radio Meteorology},
  Dover Publications, New York, 1968 (reprint of NBS Monograph~92,
  1966), 435~pp.
  Canonical reference for atmospheric refractivity~$N$, the
  $\mathrm{d}N/\mathrm{d}h$ classification (sub-refractive,
  standard, super-refractive, trapping/ducting), and global
  gradient climatologies cited in \S\ref{sec:tropo}.

\bibitem{sonntag1990}
  D.~Sonntag,
  ``Important new values of the physical constants of 1986,
  vapor pressure formulations based on the ITS-90, and psychrometer
  formulae,''
  \emph{Zeitschrift f\"ur Meteorologie}, vol.~40, no.~5,
  pp.~340--344, 1990.
  Source of the Magnus saturation-pressure coefficients
  ($a=17.62$, $b=243.12\,^\circ$C) used in \S\ref{sec:tropo};
  recommended by WMO~(2008) over the Alduchov--Eskridge form.

\bibitem{middleton1977}
  D.~Middleton,
  ``Statistical-physical models of electromagnetic interference,''
  \emph{IEEE Transactions on Electromagnetic Compatibility},
  vol.~EMC-19, no.~3, pp.~106--127, 1977.
  \textsc{doi}: 10.1109/TEMC.1977.303527.

\bibitem{igrf13}
  P.~Alken, E.~Th\'ebault, C.\,D.~Beggan, et al.,
  ``International Geomagnetic Reference Field: the thirteenth generation,''
  \emph{Earth, Planets and Space}, vol.~73, no.~49, 2021.
  \textsc{doi}: 10.1186/s40623-020-01288-x.

\bibitem{kc2g}
  A.~Rodland (KC2G),
  \emph{HF Propagation Prediction Maps},
  prop.kc2g.com.  Real-time MUF and foF2 maps assimilated from
  GIRO ionosondes, funded by the Northern California DX
  Foundation / WWROF.  Primary external real-time MUF input
  throughout ionocast.
  \url{https://prop.kc2g.com/}.

\bibitem{giro}
  B.\,W.~Reinisch and I.\,A.~Galkin,
  ``Global Ionospheric Radio Observatory (GIRO),''
  \emph{Earth, Planets and Space}, vol.~63, no.~4,
  pp.~377--381, 2011.
  \textsc{doi}: 10.5047/eps.2011.03.001.
  Source of foF2, foEs, and hmF2 measurements via the Digital
  Ionosonde Database (DIDB) at the University of Massachusetts
  Lowell.
  \url{https://giro.uml.edu/}.

\bibitem{hp30}
  Y.~Yamazaki, J.~Matzka, C.~Stolle, G.~Kervalishvili,
  J.~Rauberg, O.~Bronkalla, et al.,
  ``Geomagnetic Activity Index Hpo,''
  \emph{Geophysical Research Letters}, vol.~49, no.~10,
  e2022GL098860, 2022.
  \textsc{doi}: 10.1029/2022GL098860.
  Hp30 / Hp60 high-cadence planetary indices distributed by GFZ
  Potsdam.
  \url{https://kp.gfz.de/en/hp30-hp60/}.

\bibitem{wdckyoto}
  World Data Center for Geomagnetism, Kyoto:
  M.~Nose, T.~Iyemori, M.~Sugiura, T.~Kamei, A.~Matsuoka,
  S.~Imajo, and T.~Kotani,
  \emph{Geomagnetic Dst Index}, 2015,
  \textsc{doi}: 10.17593/14515-74000;
  and S.~Imajo, A.~Matsuoka, H.~Toh, and T.~Iyemori,
  \emph{Mid-latitude Geomagnetic Indices ASY and SYM},
  2022, \textsc{doi}: 10.14989/267216.
  Kyoto University, Graduate School of Science.
  \url{https://wdc.kugi.kyoto-u.ac.jp/}.

\bibitem{silso}
  F.~Clette, L.~Svalgaard, J.\,M.~Vaquero, and E.\,W.~Cliver,
  ``Revisiting the Sunspot Number: A 400-Year Perspective on the
  Solar Cycle,''
  \emph{Space Science Reviews}, vol.~186, no.~1--4,
  pp.~35--103, 2014.
  \textsc{doi}: 10.1007/s11214-014-0074-2.
  See also F.~Clette and L.~Lef\`evre, ``The New Sunspot Number:
  Assembling All Corrections,'' \emph{Solar Physics}, vol.~291,
  2016, \textsc{doi}: 10.1007/s11207-016-1014-y.
  Reformed (v2) SSN time series distributed by SILSO, Royal
  Observatory of Belgium.
  \url{https://www.sidc.be/SILSO/}.

\bibitem{uwyo}
  University of Wyoming, Department of Atmospheric Science,
  \emph{Wyoming Weather Web Upper Air Sounding Archive},
  Laramie, WY.  Global radiosonde sounding archive used for
  tropospheric refractivity profiles in \S\ref{sec:tropo}.
  \url{https://weather.uwyo.edu/upperair/sounding.html}.

\bibitem{wsprlive}
  WSPR Live,
  \emph{ClickHouse-backed Queryable Mirror of the WSPRnet Spot
  Database}.
  The aggregate-query interface ionocast uses for activity-
  baseline calibration and the binary ``is this band alive''
  sanity check in the harness.
  \url{https://wspr.live/}.

\bibitem{wsprnet}
  J.\,H.~Taylor (K1JT) and B.~Walker (W1BW),
  ``WSPRing Around the World,''
  \emph{QST}, November 2010, pp.~30--32, American Radio Relay
  League, Newington, CT.
  Original WSPR protocol description; spots are aggregated at
  WSPRnet (\url{https://wsprnet.org/}), the upstream from which
  WSPR~Live~\cite{wsprlive} mirrors.
  \url{https://wsjt.sourceforge.io/WSPR_QST_Nov_2010.pdf}.

\bibitem{n0nbh}
  P.~Herrman (N0NBH),
  \emph{Solar Data / HF Propagation Conditions Widget},
  hamqsl.com.  Heuristic band-condition summary widely embedded
  in amateur-radio pages.
  \url{https://www.hamqsl.com/solar.html}.

\bibitem{voacap}
  G.\,R.~Hand and L.\,R.~Teters,
  \emph{Voice of America Coverage Analysis Program (VOACAP):
  User's Guide and Reference Manual},
  Institute for Telecommunication Sciences (NTIA/ITS),
  U.S.~Department of Commerce, Boulder, CO.  IONCAP-derived
  point-to-point HF prediction implementation, distinct from
  ITU-R~P.533~\cite{itup533}.
  \url{https://www.voacap.com/}.

\bibitem{dxtoolbox}
  Black Cat Systems,
  \emph{DX Toolbox: HF / Shortwave Propagation Software},
  Westminster, MD.  Heuristic propagation forecasting tool
  combining solar-flux, K/A-index, and grayline overlays.
  \url{https://www.blackcatsystems.com/software/ham-shortwave-radio-propagation-software.html}.

\bibitem{donki}
  NASA Goddard Space Flight Center,
  \emph{Space Weather Database Of Notifications, Knowledge,
  Information (DONKI)},
  Community Coordinated Modeling Center.
  Catalogue of CME, HSS, and SEP analyses used by ionocast for
  storm-type classification.
  \url{https://kauai.ccmc.gsfc.nasa.gov/DONKI/}.

\end{thebibliography}

\end{document}
